# SQL {#sql}

```{r, include=F}
# library(RMariaDB)
library(RSQLite)
library(DBI)

# Establish a connection to sqlite databases
chinook <- dbConnect(RSQLite::SQLite(), "chinook.db")
lahman <- dbConnect(RSQLite::SQLite(), "lahman.db")

# Establish a connection to mysql databases
# connection <- dbConnect(RMariaDB::MariaDB(),
#                       host="scholar-db.rcac.purdue.edu",
#                       db="elections",
#                       user="elections_user",
#                       password="Dataelect!98")
```

```{r, eval=F}
library(RMariaDB)
library(RSQLite)
library(DBI)

# Establish a connection to sqlite databases
chinook <- dbConnect(RSQLite::SQLite(), "chinook.db")
lahman <- dbConnect(RSQLite::SQLite(), "lahman.db")

# Establish a connection to mysql databases
connection <- dbConnect(RMariaDB::MariaDB(),
                      host="your-host.com",
                      db="your-database-name",
                      user="your-username",
                      password="your-password")
```

### RDBMS {#sql-rdbms}

### SQL in R {#sql-in-r}

#### Examples {#sql-in-r-examples}

Please see [here](https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/think-summer-examples-2020.pdf) for a variety of examples demonstrating using SQL within R.

### SQL in Python {#sql-in-python}

### Examples {#sql-examples}

The following examples use the `lahman.db` sqlite database.

#### Display the first 10 ballparks in the `ballparks` table.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT * FROM parks LIMIT 10;
```
</details>

#### Make a list of the names of all of the inactive teams in baseball history. {#sql-where}

<details>
    <summary>Click here for solution</summary>

Remove the LIMIT 10 for full results.
```{sql, connection=lahman}
SELECT franchName FROM teamsfranchises WHERE active=='N' LIMIT 10;
```
</details>

#### Find the player with the most Runs Batted In (RBIs) in a season in queries. In the first query find the playerID of the player with the most RBIs. In the second query find the player's name in the `people` table.

<details>
    <summary>Click here for solution</summary>
    
In addition to his RBI record, Hack Wilson also held the NL home run record for a long time as well with 56. In 1999, Manny Ramirez tried to pursue the RBI record, but only was able to accrue 165 RBIs.
```{sql, connection=lahman}
-- Find the playerID
SELECT playerID FROM batting WHERE RBI==191;

-- Display the name
SELECT nameFirst, nameLast FROM people WHERE playerID=='wilsoha01';
```
</details>

#### Who was the manager of the 1976 "Big Red Machine" (CIN)? Complete this in 2 queries.

<details>
    <summary>Click here for solution</summary>
  
The "Big Red Machine" was a famous nickname for the dominant Cincinnati Reds of the early 1970s. Many of its team members are Hall of Famers, including their manager, Sparky Anderson.
    
```{sql, connection=lahman}
SELECT playerID FROM managers 
  WHERE yearID==1976 AND teamID=='CIN';

SELECT nameFirst, nameLast FROM people 
  WHERE playerID=='andersp01';
```
</details>

#### Make a list of the teamIDs that were managed by Tony LaRussa. Complete this in 2 queries.

<details>
    <summary>Click here for solution</summary>
    
Tony LaRussa is very well known for being a manager that was involved in baseball for a very long time. He won the World Series with the St. Louis Cardinals and the Oakland Athletics.

```{sql, connection=lahman}
SELECT playerID FROM people WHERE nameLast=='LaRussa' AND nameFirst=='Tony';

SELECT DISTINCT teamID FROM managers WHERE playerID=='larusto01';
```
</details>

#### What was Cecil Fielder's salary in 1987? Display the teamID with the salary. 

<details>
    <summary>Click here for solution</summary>

Cecil Fielder was a power hitting DH in the 1980s and 1990s. His son, Prince Fielder, played in the major leagues as well.

```{sql, connection=lahman}
SELECT playerID FROM people 
  WHERE nameFirst=='Cecil' AND nameLast=='Fielder';

SELECT teamID, salary FROM salaries 
  wHERE playerID=='fieldce01' AND yearID==1987;
```
</details>

#### Make a list of all the teams who have lost a World Series (WS) since 1990. Put the list in ascending order by `yearID`.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT teamIDloser, yearID FROM seriespost
  WHERE yearID >= 1990 AND round=='WS'
  ORDER BY yearID ASC LIMIT 10;
```
</details>

#### Let's find out about Cal Ripken, Jr. What was his height and weight? Did he bat right or left handed? When did he play his final game? Find all of this information in one query.

<details>
    <summary>Click here for solution</summary>
  
Cal Ripken, Jr's nickname is the "Iron Man" of baseball due to the fact that he started in 2,632 straight games. That means in just over 16 seasons, Cal Ripken, Jr. never missed a game!

```{sql, connection=lahman}
SELECT height, weight, bats, finalgame FROM people 
  WHERE nameFirst=='Cal' AND nameLast=='Ripken'
  AND deathState IS NULL;
```
</details>

#### Select all the playerIDs and yearIDs of the players who were inducted in the hall of fame and voted in by the Veterans committee, between 1990 and 2000. Put the list in descending order.

<details>
    <summary>Click here for solution</summary>
    
The veterans committee in the Hall of Fame voting process place players in the Hall of Fame that are forgotten by the writers, fans, etc. This is a way for players to recognize who they think were the greatest players of all time, or are skipped over for a variety of reasons. This is one reason why there is a lot of scrutiny in the process for how players are selected to the baseball hall of fame.

```{sql, connection=lahman}
SELECT playerID, yearID FROM halloffame 
  WHERE votedBy=='Veterans' AND inducted=='Y'
  AND yearID BETWEEN 1990 AND 2000
  ORDER BY yearID DESC LIMIT 10; 
```
</details>

#### Get a list of the attendance by season of the Toronto Blue Jays (TOR). What season was the highest attendance?

<details>
    <summary>Click here for solution</summary>
    
The Toronto Blue Jays were the 1993 season's World Series champion. This means that, yes, a non-USA team has won the World Series for baseball!

```{sql, connection=lahman}
SELECT yearkey, attendance FROM homegames 
  WHERE teamkey=='TOR' 
  ORDER BY attendance DESC LIMIT 10;
```
</details>

#### How many different leagues have represented Major League Baseball over time?

<details>
    <summary>Click here for solution</summary>
    
Major League Baseball has had several leagues that have been represented in its history. There are only two current leagues: National League and the American League.

```{sql, connection=lahman}
SELECT DISTINCT league FROM leagues;
```
</details>

#### Find the teams that have won the World Series.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT teamID, yearID FROM teams WHERE WSWin=='Y' LIMIT 10;
```
</details>

#### List the top 10 season win totals of teams. Include the `yearID` and `teamID`.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT teamID, yearID, W FROM teams ORDER BY W DESC LIMIT 10;
```
</details>

#### List the pitchers with their `teamID`, wins (`W`), and losses (`L`) that threw complete games (`CG`) in the 1995 season. Include their number of complete games as well.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID, teamID, W, L, CG FROM pitching
  WHERE CG > 0 AND yearID==1995
  ORDER BY W DESC LIMIT 10;
```
</details>

#### Get a printout of the Hits (`H`), and home runs (`HR`) of Ichiro Suzuki's career. Do this is in two queries. In the first query, find Ichiro Suzuki's `playerID`. In the second one list the `teamID`, `yearID`, hits and home runs.

<details>
    <summary>Click here for solution</summary>
    
Ichiro Suzuki is regarded as one of the greatest hitters of all time because of his prowess in both American and Japanese professional baseball.

```{sql, connection=lahman}
SELECT playerID FROM people
  WHERE nameFirst=='Ichiro' AND nameLast=='Suzuki';
  
SELECT teamID, yearID, H, HR FROM batting
  WHERE playerID=='suzukic01';
```
</details>

#### How many walks (`BB`) and strikeouts (`SO`) did Mariano Rivera achieve in the playoffs? Which year did Mariano Rivera give up the most post-season walks?

<details>
    <summary>Click here for solution</summary>
    
More men have walked on the moon than have scored a run on Mariano Rivera in a playoff game. Mariano Rivera made the hall of fame in 2019.

```{sql, connection=lahman}
SELECT playerID FROM people 
  WHERE nameFirst=='Mariano' AND nameLast=='Rivera';
  
SELECT yearID, teamID, BB, SO FROM pitchingpost
  WHERE playerID=='riverma01'
  ORDER BY BB DESC;
```
</details>

#### Find the pitcher with most strikeouts (`SO`), and the batter that struck out the most in the 2014 season. Get the first and last name of the pitcher and batter, respectively.

<details>
    <summary>Click here for solution</summary>
    
Corey Kluber is a two-time AL Cy Young winner. He is well known for his two-seam fastball that is difficult to hit.

```{sql, connection=lahman}
SELECT playerID, SO FROM pitching
  WHERE yearID==2014
  ORDER BY SO DESC 
  LIMIT(10);
  
SELECT playerID, SO FROM batting
  WHERE yearID==2014
  ORDER BY SO DESC
  LIMIT(10);

SELECT nameFirst,nameLast FROM people
  WHERE playerID=="klubeco01" OR playerID=="howarry01";
```
</details>

#### How many different teams did Bartolo Colon pitch for?

<details>
    <summary>Click here for solution</summary>
    
Bartolo Colon is a well-known journeyman pitcher in baseball. He has pitched with a lot of teams, but it wasn't until he played for the New York Mets when he needed to come to the plate. He had a weird batting stance that is funny to watch. He even [hit a home run](https://www.youtube.com/watch?v=OVFsq9FQBlc) one season!

```{sql, connection=lahman}
SELECT playerID FROM people
  WHERE nameFirst=='Bartolo' AND nameLast=='Colon';
  
SELECT DISTINCT teamID FROM pitching
  WHERE playerID=='colonba01';
```
</details>

#### How many times did Trevor Bauer come to bat (`AB`) in 2016? How many hits (`H`) did he get?

<details>
    <summary>Click here for solution</summary>

Trevor Bauer is much more known for his pitching than he is known for hitting. This is common for pitchers, as many are not very good at hitting.    

```{sql, connection=lahman}
SELECT playerID FROM people
  WHERE nameFirst=="Trevor" AND nameLast=="Bauer";
```

```{sql, connection=lahman}
SELECT AB, H FROM batting
 WHERE playerID=="bauertr01" AND yearID=="2016";
```
</details>

#### Let's compare Mike Trout and Giancarlo Stanton by season. Who has hit more RBIs in a season? Who has been caught stealing (`CS`) more in a season?

<details>
    <summary>Click here for solution</summary>

Mike Trout and Giancarlo Stanton are considered two of the of the best hitters in Major League Baseball for very different reasons. Trout is an all-around player known for being indispensible, where Stanton is known as a power hitter.    

```{sql, connection=lahman}
SELECT playerID, nameFirst, nameLast FROM people 
  WHERE (nameFirst=='Giancarlo' AND nameLast=='Stanton')
  OR  (nameFirst=='Mike' AND nameLast=='Trout');
```

```{sql, connection=lahman}
SELECT playerID, yearID, teamID, RBI, CS FROM batting
  WHERE playerID=='stantmi03' OR playerID=='troutmi01'
  ORDER BY RBI DESC LIMIT 1;
```

```{sql, connection=lahman}
SELECT playerID, yearID, teamID, RBI, CS FROM batting
  WHERE playerID=='stantmi03' OR playerID=='troutmi01'
  ORDER BY CS DESC LIMIT 1;
```
</details>

#### Make a list of players who walked (`BB`) more than they struck out (`SO`) between 1980 and 1985. Of these players, who walked the most? Use the `BETWEEN` command in this queries. Use a second query to get the player's first and last name.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID, yearID, teamID, BB, SO FROM batting
  WHERE BB > SO LIMIT 10;
```

```{sql, connection=lahman}
SELECT nameFirst, nameLast FROM people WHERE playerID=='randowi01';
```
</details>

#### How many different NL catchers (C) won gold glove winners between 1990 and 2000?

<details>
    <summary>Click here for solution</summary>
    
There were 6 different catchers.
    
```{sql, connection=lahman}
SELECT DISTINCT playerID FROM awardsplayers
  WHERE awardID=='Gold Glove' AND notes=='C'
  AND lgID=='NL' AND yearID BETWEEN 1990 AND 2000;
```
</details>

#### How many different 3rd Basemen played for the Seattle Mariners between 2000 and 2005? Who had the most Errors?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT DISTINCT playerID, yearID, E FROM fielding WHERE 
    yearID BETWEEN 2000 AND 2005 AND teamID=='SEA'
    AND POS=='3B'
    ORDER BY E DESC LIMIT 10;
```

```{sql, connection=lahman}
SELECT nameFirst, nameLast FROM people
  WHERE playerID=='camermi01';
```
</details>

#### Craig Biggio was more known for his play at second base over his major league baseball career, but he didn't always play second base. What seasons did Craig Biggio play Catcher? 

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID FROM people
  WHERE nameFirst=='Craig' AND nameLast=='Biggio';
```

```{sql, connection=lahman}
SELECT teamID, yearID, POS FROM fielding
  WHERE playerID=='biggicr01' AND POS=='C';
```
</details>

#### Find the teams that have won the World Series that represented the National League. Display the list with the `yearID` and `teamID` in ascending order.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT teamID, yearID FROM teams 
  WHERE WSWin=='Y' AND lgID=='NL'
  ORDER BY yearID ASC LIMIT 10;
```
</details>

#### List the pitchers that threw at least one complete game (CG) in the 1995 season. Please include the wins and losses of the top 10 pitchers. Use the playerID of the pitcher who threw the most complete games to find out the name of the pitcher that had the most complete games.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID, W, L, CG FROM pitching 
  WHERE CG > 0 AND yearID==1995
  ORDER BY CG DESC
  LIMIT 10;
```

```{sql, connection=lahman}
SELECT nameFirst, nameLast FROM people
  WHERE playerID=='maddugr01';
```
</details>

#### Who was the most recent player manager?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID, yearID FROM managers 
  WHERE plyrMgr=='Y' 
  ORDER BY yearID DESC LIMIT 10;
```

```{sql, connection=lahman}
SELECT nameFirst, nameLast FROM people WHERE playerID=='rosepe01';
```
</details>

#### Get the at-bats, homeruns, stolen bases for Roberto Clemente by year in ascending order.

<details>
    <summary>Click here for solution</summary>
    
Roberto Clemente is known as being a leader for the Pittsburgh Pirates. He died in a 1972 plane crash on a humanitarian mission to Puerto Rico, where he grew up.

```{sql, connection=lahman}
SELECT playerID FROM people 
  WHERE nameFirst=='Roberto' AND nameLast=='Clemente';
```

```{sql, connection=lahman}
SELECT yearID,AB,HR,SB FROM battingpost 
  WHERE playerID=='clemero01'
  ORDER BY yearID ASC;
```
</details>

#### Get a list of distinct World Series winners from the years Tom Lasorda managed the Los Angeles Dodgers (LAN). First find the years Tom Lasorda was the manager of the Los Angeles Dodgers, and then find the distinct teams that won a World Series in that time frame.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID FROM people 
  WHERE nameFirst=='Tom' AND nameLast=='Lasorda';
```

```{sql, connection=lahman}
SELECT yearID FROM managers
  WHERE playerID=='lasorto01' LIMIT 10;
```

```{sql, connection=lahman}
SELECT DISTINCT teamID FROM teams
  WHERE WSWin=='Y' AND yearID BETWEEN 1976 AND 1996;
```
</details>

#### Which teams did Kenny Lofton steal more than 20 bases in a season after the year 2000?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID FROM people 
  WHERE nameFirst=='Kenny' AND nameLast=='Lofton';
```

```{sql, connection=lahman}
SELECT teamID, yearID, SB FROM batting
  WHERE playerID=='loftoke01' AND SB > 20
  AND yearID >2000;
```
</details>

#### How much did the Tampa Bay Rays (TBL) pay Wade Boggs in 1998? Who paid Boggs the most in a season during his career?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID FROM people 
  WHERE nameFirst=='Wade' AND nameLast=='Boggs';
```

```{sql, connection=lahman}
SELECT teamID, yearID, salary FROM salaries
    WHERE playerID=='boggswa01'
    AND yearID==1998;
```

```{sql, connection=lahman}
SELECT teamID, yearID, salary FROM salaries
    WHERE playerID=='boggswa01'
    ORDER BY salary DESC LIMIT 10;
```
</details>

#### 

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT teamID, yearID, W, L, HR, HRA, attendance FROM teams
  WHERE teamID=='DET' AND (WSWin=='Y' OR LgWin=='Y');
```
</details>

#### The standings you would find in a newspaper often have Wins and Losses in order of most to least wins. There are often other numbers that are involved like winning percentage, and other team statistics, but we won't deal with that for now. Get the NL East Standings in 2015. 

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT teamID, W, L FROM teams
  WHERE divID=='E' AND lgID=='NL'
  AND yearID==2015
  ORDER BY teamrank ASC;
```
</details>

#### Make a list of the teams, wins, losses, years for NL East teams that have won the World Series. Which team had the most wins?

<details>
    <summary>Click here for solution</summary>

```{sql, connection=lahman}
SELECT teamID, yearID, W, L FROM teams
  WHERE lgID=='NL' AND divID=='E' AND WSWin=='Y'
  ORDER BY W DESC;
```
</details>

#### Get a list of the `playerIDs` of managers who won more games than they lost between 1930 and 1950. Get the manager's name, and the name of the team of the manager with the most wins on the list.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID, teamID, yearID, W, L FROM managers
    WHERE yearID BETWEEN 1930 AND 1950 AND W > L
    ORDER BY W DESC LIMIT 10;
```

```{sql, connection=lahman}
SELECT nameFirst, nameLast FROM people
  WHERE playerID=='mackco01';
```

```{sql, connection=lahman}
SELECT franchName FROM teamsfranchises 
  WHERE franchID=='PHA';
```
</details>

#### Get the top 5 seasons from Florida Teams (Florida Marlins, Tampa Bay Rays, and Miami Marlins) in attendance. How many have occured since 2000?

<details>
    <summary>Click here for solution</summary>
    
Florida baseball teams are not known for their attendance for a variety of reasons. Both MLB franchises play in domed fields, but usually do not draw large crowds. 
    
```{sql, connection=lahman}
SELECT franchID, franchName FROM teamsfranchises
    WHERE franchName=='Tampa Bay Rays'
    OR franchName=='Florida Marlins';
```

```{sql, connection=lahman}
SELECT teamID, yearID, attendance FROM teams
  WHERE franchID=='TBD' OR franchID=='FLA'
  ORDER BY attendance DESC LIMIT 10;
```
</details>

#### What pitcher has thrown the most Shutouts (SHO) in the AL since 2010? What about the NL? Please get their first and last names respectively.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=lahman}
SELECT playerID,teamID, yearID, SHO FROM pitching
  WHERE yearID>2010 AND lgID=='NL'
  ORDER BY SHO DESC LIMIT 10;
```

```{sql, connection=lahman}
SELECT playerID,teamID, yearID, SHO FROM pitching
  WHERE yearID>2010 AND lgID=='AL'
  ORDER BY SHO DESC LIMIT 10;    
```

```{sql, connection=lahman}
SELECT nameFirst, nameLast FROM people
  WHERE playerID=='leecl02' OR playerID=='hernafe02';
```
</details>

The following examples use the `chinook.db` sqlite database.

```{r}
dbListTables(chinook)
```

#### How do I select all of the rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM employees;
```
</details>

#### How do I select the first 5 rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM employees LIMIT 5;
```
</details>

#### How do I select specific rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT LastName, FirstName FROM employees;
```
    
You can switch the order in which the columns are displayed as well:
```{sql, connection=chinook}
SELECT FirstName, LastName FROM employees;
```
</details>

#### How do I select only unique values from a column?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT DISTINCT Title FROM employees;
```
</details>

#### How can I filter that match a certain criteria?

<details>
    <summary>Click here for solution</summary>
    
Select only employees with a FirstName "Steve":
```{sql, connection=chinook}
SELECT * FROM employees WHERE FirstName='Steve';
```

Select only employees with FirstName "Steve" OR FirstName "Laura":
```{sql, connection=chinook}
SELECT * FROM employees WHERE FirstName='Steve' OR FirstName='Laura';
```

Select only employees with FirstName "Steve" AND LastName "Laura":
```{sql, connection=chinook}
SELECT * FROM employees WHERE FirstName='Steve' AND LastName='Laura';
```
As expected, there are no results! There is nobody with the full name "Steve Laura".
</details>


#### List the first 10 tracks from the `tracks` table.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM tracks LIMIT 10;
```
</details>

#### How many rows or records are in the table named `tracks`?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT COUNT(*) FROM tracks;
```
</details>

#### Are there any artists with the names: "Elis Regina", "Seu Jorge", or "The Beatles"?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM artists WHERE Name='Elis Regina' OR Name='Seu Jorge' OR Name='The Beatles';
```
</details>

#### What albums did the artist with `ArtistId` of 41 make?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM albums WHERE ArtistId=41;
```
</details>

#### What are the tracks of the album with `AlbumId` of 71? Order the results from most `Milliseconds` to least. 

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM tracks WHERE AlbumId=71 ORDER BY Milliseconds DESC;
```
</details>

#### What are the tracks of the album with `AlbumId` of 71? Order the results from longest to shortest and convert `Milliseconds` to seconds. Use aliasing to name the calculated field `Seconds`. {#sql-aliasing}

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT Milliseconds/1000.0 AS Seconds, * FROM tracks WHERE AlbumId=71 ORDER BY Seconds DESC;
```
</details>

#### What are the tracks that are at least 250 seconds long?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT Milliseconds/1000.0 AS Seconds, * FROM tracks WHERE Seconds >= 250;
```
</details>

#### What are the tracks that are between 250 and 300 seconds long?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT Milliseconds/1000.0 AS Seconds, * FROM tracks WHERE Seconds BETWEEN 250 AND 300 ORDER BY Seconds;
```
</details>

#### What is the `GenreId` of the genre with name `Pop`?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT GenreId FROM genres WHERE Name='Pop';
```
</details>

#### What is the average length (in seconds) of a track with genre "Pop"?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT AVG(Milliseconds/1000.0) AS avg FROM tracks WHERE genreId=9;
```
</details>

#### What is the longest Bossa Nova track (in seconds)?

<details>
    <summary>Click here for solution</summary>
    
What is the `GenreId` of Bossa Nova?
```{sql, connection=chinook}
SELECT GenreId FROM genres WHERE Name='Bossa Nova';
```
```{sql, connection=chinook}
SELECT *, MAX(Milliseconds/1000.0) AS Seconds FROM tracks WHERE genreId=11;
```
</details>

#### Get the average price per hour for Bossa Nova music (`genreId` of 11).

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT AVG(UnitPrice/Milliseconds/1000.0/3600) AS 'Price per Hour' FROM tracks WHERE genreId=11;
```
</details>

#### Get the average time (in seconds) for tracks by genre. {#sql-groupby}

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT genreId, AVG(Milliseconds/1000.0) AS 'Average seconds per track' FROM tracks GROUP BY genreId;
```

We can use an INNER JOIN to get the name of each genre as well.
```{sql, connection=chinook}
SELECT g.Name, track_time.'Average seconds per track' FROM genres AS g INNER JOIN (SELECT genreId, AVG(Milliseconds/1000.0) AS 'Average seconds per track' FROM tracks GROUP BY genreId) AS track_time ON g.GenreId=track_time.GenreId ORDER BY track_time.'Average seconds per track' DESC;
```

</details>

#### What is the average price per track for each genre?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT genreId, AVG(UnitPrice) AS 'Average seconds per track' FROM tracks GROUP BY genreId;
```
</details>

#### What is the average number of tracks per album?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT AVG(trackCount) FROM (SELECT COUNT(*) AS trackCount FROM tracks GROUP BY albumId) AS track_count;
```
</details>

#### What is the average number of tracks per album per genre?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT genreId, AVG(trackCount) FROM (SELECT genreId, COUNT(*) AS trackCount FROM tracks GROUP BY albumId) AS track_count GROUP BY genreId;
```

```{sql, connection=chinook}
SELECT Name, avg_track_count.'Average Track Count' FROM genres AS g INNER JOIN (SELECT genreId, AVG(trackCount) AS 'Average Track Count' FROM (SELECT genreId, COUNT(*) AS trackCount FROM tracks GROUP BY albumId) AS track_count GROUP BY genreId) AS avg_track_count ON g.GenreId=avg_track_count.genreId;
```
</details>



The following examples us the `lahman.db` sqlite database.

```{r}
dbListTables(lahman)
```

