# SQL {#sql}

```{r setup, include=F}
library(RSQLite)
library(DBI)

db <- dbConnect(RSQLite::SQLite(), "chinook.db")
```

### RDBMS

### SQL in R

### SQL in Python

### Examples

The following examples us the `chinook.db` sqlite database.

```{r}
dbListTables(db)
```

#### How do I select all of the rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=db}
SELECT * FROM employees;
```
</details>

#### How do I select the first 5 rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=db}
SELECT * FROM employees LIMIT 5;
```
</details>

#### How do I select specific rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=db}
SELECT LastName, FirstName FROM employees;
```
    
You can switch the order in which the columns are displayed as well:
```{sql, connection=db}
SELECT FirstName, LastName FROM employees;
```
</details>

#### How do I select only unique values from a column?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=db}
SELECT DISTINCT Title FROM employees;
```
</details>

#### How can I filter that match a certain criteria?

<details>
    <summary>Click here for solution</summary>
    
Select only employees with a FirstName "Steve":
```{sql, connection=db}
SELECT * FROM employees WHERE FirstName='Steve';
```

Select only employees with FirstName "Steve" OR FirstName "Laura":
```{sql, connection=db}
SELECT * FROM employees WHERE FirstName='Steve' OR FirstName='Laura';
```

Select only employees with FirstName "Steve" AND LastName "Laura":
```{sql, connection=db}
SELECT * FROM employees WHERE FirstName='Steve' AND LastName='Laura';
```
As expected, there are no results! There is nobody with the full name "Steve Laura".
</details>


#### List the first 10 tracks from the `tracks` table.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=db}
SELECT * FROM tracks LIMIT 10;
```
</details>

#### How many rows or records are in the table named `tracks`?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=db}
SELECT COUNT(*) FROM tracks;
```
</details>
