# SQL {#sql}

```{r setup, include=F}
library(RMariaDB)
library(RSQLite)
library(DBI)

# Establish a connection to sqlite databases
chinook <- dbConnect(RSQLite::SQLite(), "chinook.db")
lahman <- dbConnect(RSQLite::SQLite(), "lahman.db")

# Establish a connection to mysql databases
# connection <- dbConnect(RMariaDB::MariaDB(),
#                       host="scholar-db.rcac.purdue.edu",
#                       db="elections",
#                       user="elections_user",
#                       password="Dataelect!98")
```

```{r setup, eval=F}
library(RMariaDB)
library(RSQLite)
library(DBI)

# Establish a connection to sqlite databases
chinook <- dbConnect(RSQLite::SQLite(), "chinook.db")
lahman <- dbConnect(RSQLite::SQLite(), "lahman.db")

# Establish a connection to mysql databases
connection <- dbConnect(RMariaDB::MariaDB(),
                      host="your-host.com",
                      db="your-database-name",
                      user="your-username",
                      password="your-password")
```

### RDBMS {#sql-rdbms}

### SQL in R {#sql-in-r}

#### Examples {#sql-in-r-examples}

Please see [here](https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/think-summer-examples-2020.pdf) for a variety of examples demonstrating using SQL within R.

### SQL in Python {#sql-in-python}

### Examples {#sql-examples}

The following examples us the `chinook.db` sqlite database.

```{r}
dbListTables(chinook)
```

#### How do I select all of the rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM employees;
```
</details>

#### How do I select the first 5 rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM employees LIMIT 5;
```
</details>

#### How do I select specific rows of a table called employees?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT LastName, FirstName FROM employees;
```
    
You can switch the order in which the columns are displayed as well:
```{sql, connection=chinook}
SELECT FirstName, LastName FROM employees;
```
</details>

#### How do I select only unique values from a column?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT DISTINCT Title FROM employees;
```
</details>

#### How can I filter that match a certain criteria?

<details>
    <summary>Click here for solution</summary>
    
Select only employees with a FirstName "Steve":
```{sql, connection=chinook}
SELECT * FROM employees WHERE FirstName='Steve';
```

Select only employees with FirstName "Steve" OR FirstName "Laura":
```{sql, connection=chinook}
SELECT * FROM employees WHERE FirstName='Steve' OR FirstName='Laura';
```

Select only employees with FirstName "Steve" AND LastName "Laura":
```{sql, connection=chinook}
SELECT * FROM employees WHERE FirstName='Steve' AND LastName='Laura';
```
As expected, there are no results! There is nobody with the full name "Steve Laura".
</details>


#### List the first 10 tracks from the `tracks` table.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM tracks LIMIT 10;
```
</details>

#### How many rows or records are in the table named `tracks`?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT COUNT(*) FROM tracks;
```
</details>

#### Are there any artists with the names: "Elis Regina", "Seu Jorge", or "The Beatles"?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM artists WHERE Name='Elis Regina' OR Name='Seu Jorge' OR Name='The Beatles';
```
</details>

#### What albums did the artist with `ArtistId` of 41 make?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM albums WHERE ArtistId=41;
```
</details>

#### What are the tracks of the album with `AlbumId` of 71? Order the results from most `Milliseconds` to least.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT * FROM tracks WHERE AlbumId=71 ORDER BY Milliseconds DESC;
```
</details>

#### What are the tracks of the album with `AlbumId` of 71? Order the results from longest to shortest and convert `Milliseconds` to seconds. Use aliasing to name the calculated field `Seconds`.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT Milliseconds/1000.0 AS Seconds, * FROM tracks WHERE AlbumId=71 ORDER BY Seconds DESC;
```
</details>

#### What are the tracks that are at least 250 seconds long?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT Milliseconds/1000.0 AS Seconds, * FROM tracks WHERE Seconds >= 250;
```
</details>

#### What are the tracks that are between 250 and 300 seconds long?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT Milliseconds/1000.0 AS Seconds, * FROM tracks WHERE Seconds BETWEEN 250 AND 300 ORDER BY Seconds;
```
</details>

#### What is the `GenreId` of the genre with name `Pop`?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT GenreId FROM genres WHERE Name='Pop';
```
</details>

#### What is the average length (in seconds) of a track with genre "Pop"?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT AVG(Milliseconds/1000.0) AS avg FROM tracks WHERE genreId=9;
```
</details>

#### What is the longest Bossa Nova track (in seconds)?

<details>
    <summary>Click here for solution</summary>
    
What is the `GenreId` of Bossa Nova?
```{sql, connection=chinook}
SELECT GenreId FROM genres WHERE Name='Bossa Nova';
```
```{sql, connection=chinook}
SELECT *, MAX(Milliseconds/1000.0) AS Seconds FROM tracks WHERE genreId=11;
```
</details>

#### Get the average price per hour for Bossa Nova music (`genreId` of 11).

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT AVG(UnitPrice/Milliseconds/1000.0/3600) AS 'Price per Hour' FROM tracks WHERE genreId=11;
```
</details>

#### Get the average time (in seconds) for tracks by genre.

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT genreId, AVG(Milliseconds/1000.0) AS 'Average seconds per track' FROM tracks GROUP BY genreId;
```

We can use an INNER JOIN to get the name of each genre as well.
```{sql, connection=chinook}
SELECT g.Name, track_time.'Average seconds per track' FROM genres AS g INNER JOIN (SELECT genreId, AVG(Milliseconds/1000.0) AS 'Average seconds per track' FROM tracks GROUP BY genreId) AS track_time ON g.GenreId=track_time.GenreId ORDER BY track_time.'Average seconds per track' DESC;
```

</details>

#### What is the average price per track for each genre?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT genreId, AVG(UnitPrice) AS 'Average seconds per track' FROM tracks GROUP BY genreId;
```
</details>

#### What is the average number of tracks per album?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT AVG(trackCount) FROM (SELECT COUNT(*) AS trackCount FROM tracks GROUP BY albumId) AS track_count;
```
</details>

#### What is the average number of tracks per album per genre?

<details>
    <summary>Click here for solution</summary>
```{sql, connection=chinook}
SELECT genreId, AVG(trackCount) FROM (SELECT genreId, COUNT(*) AS trackCount FROM tracks GROUP BY albumId) AS track_count GROUP BY genreId;
```

```{sql, connection=chinook}
SELECT Name, avg_track_count.'Average Track Count' FROM genres AS g INNER JOIN (SELECT genreId, AVG(trackCount) AS 'Average Track Count' FROM (SELECT genreId, COUNT(*) AS trackCount FROM tracks GROUP BY albumId) AS track_count GROUP BY genreId) AS avg_track_count ON g.GenreId=avg_track_count.genreId;
```
</details>



The following examples us the `lahman.db` sqlite database.

```{r}
dbListTables(lahman)
```

