# R {#r}

## Getting started {#getting-started-with-r}

## Variables {#r-variables}

### NA {#r-na}

NA stands for not available and, in general, represents a missing value or a lack of data.

#### How do I tell if a value is NA? {r-isna}

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Test if value is NA.
value <- NA
is.na(value)

# Does is.nan return TRUE for NA?
is.nan(value)
```
</details>

### NaN {#r-nan}

NaN stands for not a number and, in general, is used for arithmetic purposes, for example, the result of 0/0.

#### How do I tell if a value is NaN?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Test if a value is NaN.
value <- NaN
is.nan(value)

value <- 0/0
is.nan(value)

# Does is.na return TRUE for NaN?
is.na(value)
```
</details>

### NULL {#r-null}

NULL represents the null object, and is often returned when we have undefined values. 

#### How do I tell if a value is NULL?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Test if a value is NaN.
value <- NULL
is.null(value)

class(value)

# Does is.na return TRUE for NULL?
is.na(value)
```

### Dates {#r-dates}

`Date` is a class which allows you to perform special operations like subtraction, where the number of days between dates are returned. Or addition, where you can add 30 to a `Date` and a `Date` is returned where the value is 30 days in the future.

You will usually need to specify the `format` argument based on the format of your date strings. For example, if you had a string `07/05/1990`, the format would be: `%m/%d/%Y`. If your string was `31-12-90`, the format would be `%d-%m-%y`. Replace `%d`, `%m`, `%Y`, and `%y` according to your date strings. A full list of formats can be found [here](https://www.stat.berkeley.edu/~s133/dates.html).

#### How do I convert a string "07/05/1990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_string <- "07/05/1990"
my_date <- as.Date(my_string, format="%m/%d/%Y")
my_date
```
</details>

#### How do I convert a string "31-12-1990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_string <- "31-12-1990"
my_date <- as.Date(my_string, format="%d-%m-%Y")
my_date
```
</details>

#### How do I convert a string "12-31-1990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_string <- "12-31-1990"
my_date <- as.Date(my_string, format="%m-%d-%Y")
my_date
```
</details>

#### How do I convert a string "31121990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_string <- "31121990"
my_date <- as.Date(my_string, format="%d%m%Y")
my_date
```
</details>

### Factors {#r-factors}

A `factor` is R's way of representing a categorical variable. Each `factor` is essentially a numeric value with an associated name. They are a useful when a vector has only a few different values it could be, like "Male" and "Female" or "A", "B", or "C".

#### How do I test whether or not a value is a factor?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
test_factor <- factor("Male")
is.factor(test_factor)

test_factor_vec <- factor(c("Male", "Female", "Female"))
is.factor(test_factor_vec)
```
</details>

#### How do I convert a vector of strings to a vector of factors?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- c("Male", "Female", "Female")
vec <- factor(c("Male", "Female", "Female"))
```
</details>

#### How do I get the unique values a factor could hold, also known as _levels_?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- factor(c("Male", "Female", "Female"))
levels(vec)
```
</details>

#### How can I rename the levels of a vector of factors?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- factor(c("Male", "Female", "Female"))
levels(vec)
levels(vec) <- c("F", "M")
vec

# Be careful! Order matters, this is wrong:
vec <- factor(c("Male", "Female", "Female"))
levels(vec)
levels(vec) <- c("M", "F")
vec
```
</details>

## Logical operators {#r-logical-operators}

Logical operators are symbols that can be used within R to compare values or vectors of values. 

Operator | Description
---------|------------
`<`      | less than
`<=`     | less than or equal to
`>`      | greater than
`>=`     | greater than or equal to
`==`     | equal to
`!=`     | not equal to
`!x`     | negation, not x
`x|y`    | x OR y
`x&y`    | x AND y

### Examples

#### What are the values in a vector, `vec` that are greater than 5?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
vec > 5
```
</details>

#### What are the values in a vector, `vec` that are greater than or equal to 5?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
vec >= 5
```
</details>

#### What are the values in a vector, `vec` that are less than 5?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
vec < 5
```
</details>

#### What are the values in a vector, `vec` that are less than or equal to 5?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
vec <= 5
```
</details>

#### What are the values in a vector that are greater than 7 OR less than or equal to 2?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
vec > 7 | vec <=2
```
</details>

#### What are the values in a vector that are greater than 3 AND less than 6?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
vec > 3 & vec < 6
```
</details>

#### How do I get the values in `list1` that are in `list2`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
list1 <- c("this", "is", "a", "test")
list2 <- c("this", "a", "exam")
list1[list1 %in% list2]
```
</details>

#### How do I get the values in `list1` that are not in `list2`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
list1 <- c("this", "is", "a", "test")
list2 <- c("this", "a", "exam")
list1[!(list1 %in% list2)]
```
</details>

#### How can I get the number of values in a vector that are greater than 5?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- 1:10
sum(vec>5)

# Note, you do not need to do:
length(vec[vec>5])

# because TRUE==1 and FALSE==0 in R
TRUE==1
FALSE==0
```
</details>

### Resources 

**[Operators Summary](https://www.statmethods.net/management/operators.html)**

A quick list of the various operators with a few simple examples.

## Lists & Vectors {#r-lists-and-vectors}

A vector contains values that are all the same type. The following are some examples of vectors:

```{r}
# A logical vector
lvec <- c(F, T, TRUE, FALSE)
class(lvec)

# A numeric vector
nvec <- c(1,2,3,4)
class(nvec)

# A character vector
cvec <- c("this", "is", "a", "test")
class(cvec)
```
As soon as you try to mix and match types, elements are coerced to the simplest type required to represent all the data.

The order of representation is:

logical, numeric, character, list

For example:

```{r}
class(c(F, 1, 2))
class(c(F, 1, 2, "ok"))
class(c(F, 1, 2, "ok", list(1, 2, "ok")))
```

Lists are vectors that can contain any class of data. For example:

```{r}
list(TRUE, 1, 2, "OK", c(1,2,3))
```

With lists, there are 3 ways you can index.

```{r}
my_list <- list(TRUE, 1, 2, "OK", c(1,2,3), list("OK", 1,2, F))

# The first way is with single square brackets [].
# This will always return a list, even if the content 
# only has 1 component.
class(my_list[1:2])
class(my_list[3])

# The second way is with double brackets [[]].
# This will return the content itself. If the
# content is something other than a list it will
# return the value itself.
class(my_list[[1]])
class(my_list[[3]])

# Of course, if the value is a list itself, it will
# remain a list.
class(my_list[[6]])

# The third way is using $ to extract a single, named variable. 
# We need to add names first! $ is like the double bracket,
# in that it will return the simplest form.
my_list <- list(first=TRUE, second=1, third=2, fourth="OK", embedded_vector=c(1,2,3), embedded_list=list("OK", 1,2, F))
my_list$first
my_list$embedded_list
```

#### How do get the type of a vector?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_vector <- c(0, 1, 2)
typeof(my_vector)
```
</details>

#### How do I convert a character vector to a numeric?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_character_vector <- c('1','2','3','4')
as.numeric(my_character_vector)
```
</details>

#### How do I convert a numeric vector to a character?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_numeric_vector <- c(1,2,3,4)
as.character(my_numeric_vector)
```
</details>

### Indexing {#r-indexing}

Indexing enables us to access a subset of the elements in vectors and lists. There are three types of indexing: positional/numeric, logical, and reference/named.

You can create a named vector and a named list easily:

```{r}
my_vec <- 1:5
names(my_vec) <- c('a','b','c','d','e')

my_list <- list(1,2,3,4,5)
names(my_list) <- c('a','b','c','d','e')

my_list2 <- list('a' = 1, 'b' = 2, 'c' = 3,'d' = 4, 'e' = 5)
```


```{r}
# Numeric (positional) indexing:
my_vec[1:2]
my_vec[c(1,3)]

my_list[1:2]
my_list[c(1,3)]
```

```{r}
# Logical indexing:
my_vec[c(T, F, T, F, F)]

my_list[c(T, F, T, F, F)]
```

```{r}
# Named (reference) indexing:
# if there are named values:
my_vec[c("a", "c")]

my_list[c("a", "c")]
```

#### Examples

##### How can I get the first 2 values of a vector named `my_vec`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_vec <- c(1, 13, 2, 9)
names(my_vec) <- c('cat', 'dog','snake', 'otter')
my_vec[1:2]
```
</details>


##### How can I get the values that are greater than 2?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_vec[my_vec>2]
```
</details>

##### How can I get the values greater than 5 and smaller than 10?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_vec[my_vec > 5 & my_vec < 10]
```
</details>


##### How can I get the values greater than 10 or smaller than 3?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_vec[my_vec > 10 | my_vec < 3]
```
</details>

##### How can I get the values for "otter" and "dog"?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_vec[c('otter','dog')]
```
</details>

### Recycling {#r-recycling}

Often operations in `R` on two or more vectors require them to be the same length. When `R` encounters vectors with different lengths, it automatically repeats (recycles) the shorter vector until the length of the vectors is the same.

### Examples

#### Given two numeric vectors with different lengths, add them element-wise.

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1,2,3)
y <- c(0,1)
x+y
```
</details>

## Basic R functions {#r-basic-functions}

### `all` {#r-all}

`all` returns a logical value (`TRUE` or `FALSE`) if all values in a vector are `TRUE`.

#### Examples

##### Are all values in `x` positive?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1, 2, 3, 4, 8, -1, 7, 3, 4, -2, 1, 3)
all(x>0) # FALSE
```
</details>

### `any` {#r-any}

`all` returns a logical value (`TRUE` or `FALSE`) if any values in a vector are `TRUE`.

#### Examples

##### Are any values in `x` positive?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1, 2, 3, 4, 8, -1, 7, 3, 4, -2, 1, 3)
any(x>0) # TRUE
```
</details>

### `all.equal` {#r-all.equal}

`all.equal` compares two objects and tests if they are "nearly equal" (up to some provided tolerance).

#### Examples

##### Is $\pi$ equal to 3.14?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
all.equal(pi, 3.14) # FALSE
```
</details>

##### Is $\pi$ equal to 3.14 if our tolerance is 2 decimal cases?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
all.equal(pi, 3.14, tol=0.01) # TRUE
```
</details>

##### Are the vectors `x` and `y` equal?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- 1:5
y <- c('1', '2', '3', '4', '5')
all.equal(x, y) # difference in type (numeric vs. character)
all.equal(x, as.numeric(y)) # TRUE
```
</details>

### `dim` {#r-dim}

`dim` returns the dimensions of a matrix or data.frame. The first value is the rows, the second is columns.

#### Examples

##### How many dimensions does the data.frame `dat` have?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
dat <- data.frame("col1"=c(1,2,3), "col2"=c("a", "b", "c"))
dim(dat) # 3 rows and 2 columns
```
</details>

### `length` {#r-length}

`length` allows you to get or set the length of an object in R (for which a method has been defined).

#### How do I get how many values are in a vector?
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Create a vector of length 5
my_vector <- c(1,2,3,4,5)

# Calculate the length of my_vector
length(my_vector)
```
</details>

### `rep` {#r-rep}

`rep` is short for replicate. `rep` accepts some object, `x`, and up to three additional arguments: `times`, `length.out`, and `each`. `times` is the number of non-negative _times_ to repeat the whole object `x`. `length.out` specifies the end length you want the result to be. `rep` will repeat the values in `x` as many times as it takes to reach the provided `length.out`. `each` repeats each element in `x` the number of times specified by `each`.

#### Examples

##### How do I repeat values in a vector 3 times?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- c(1,2,3)
rep(vec, 3)

# or

rep(vec, times=3)
```
</details>

##### How do I repeat the values in a vector enough times to be the same length as another vector?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- c(1,2,3)
other_vec <- c(1,2,2,2,2,2,2,8)
rep(vec, length.out=length(other_vec))

# Note that if the end goal is to do something 
# like add the two vectors, this can be done
# using recycling.
rep(vec, length.out=length(other_vec)) + other_vec

vec + other_vec
```
</details>

##### How can I repeat each value inside a vector a certain amount of times?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- c(1,2,3)
rep(vec, each=3)
```
</details>

##### How can I repeat the values in one vector based on the values in another vector?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- c(1,2,3)
rep_by <- c(3,2,1)
rep(vec, times=rep_by)
```
</details>

### `rbind` and `cbind` {#r-bind}

### `which`, `which.max`, `which.min` {#r-which}

`which` enables you to find the position of the elements that are `TRUE` in a logical vector.

`which.max` and `which.min` finds the location of the maximum and minimum, respectively, of a numeric (or logical) vector. 

#### Examples

##### Given a numeric vector, return the index of the maximum value.
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1,-10, 2,4,-3,9,2,-2,4,8)
which.max(x)

# which.max is just shorthand for:
which(x==max(x))
```
</details>

##### Given a vector, return the index of the positive values.
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1,-10, 2,4,-3,9,2,-2,4,8)
which(x>0)
```
</details>

##### Given a matrix, return the indexes (row and column) of the positive values.
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- matrix(c(1,-10, 2,4,-3,9,2,-2,4,8), ncol=2)
which(x>0, arr.ind = TRUE)
```
</details>

### `grep`, `grepl`, etc. {#r-grep}

`grep` allows you to use regular expressions to search for a pattern in a string or character vector, and returns the index where there is a match.

`grepl` performs the same operation but rather than returning indices, returns a vector of logical `TRUE` or `FALSE` values.

#### Examples

##### Given a character vector, return the index of any words ending in "s".
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
grep("*.s$", c("waffle", "waffles", "pancake", "pancakes"))
```
</details>

##### Given a character vector, return a vector of the same length where each element is `TRUE` if there was a match for any word ending in "s", and `FALSE otherwise.
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
grepl("*.s$", c("waffle", "waffles", "pancake", "pancakes"))
```
</details>

### `mean` {#r-mean}

`mean` is a function that calculates the average of a vector of values.

#### How do I get the average of a vector of values?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
mean(c(1,2,3,4))
```
</details>

#### How do I get the average of a vector of values when some of the values are: `NA`, `NaN`?

<details>
    <summary>Click here for solution</summary>
    
Many R functions have the `na.rm` argument available. This argument is "a logical value indicating whether NA values should be stripped before the computation proceeds." 

```{r, eval=T}
mean(c(1,2,3,NaN), na.rm=T)

mean(c(1,2,3,NA), na.rm=T)

mean(c(1,2,NA,NaN,4), na.rm=T)
```
</details>

### `var` {#r-var}

`var` is a function that calculate the variance of a vector of values.

#### How do I get the variance of a vector of values?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
var(c(1,2,3,4))
```
</details>

#### How do I get the variance of a vector of values when some of the values are: `NA`, `NaN`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
var(c(1,2,3,NaN), na.rm=T)

var(c(1,2,3,NA), na.rm=T)

var(c(1,2,NA,NaN,4), na.rm=T)
```
</details>

#### How do I get the standard deviation of a vector of values?

<details>
    <summary>Click here for solution</summary>
    
The standard deviation is equal to the square root of the variance.

```{r, eval=T}
sqrt(var(c(1,2,3,NaN), na.rm=T))

sqrt(var(c(1,2,3,NA), na.rm=T))

sqrt(var(c(1,2,NA,NaN,4), na.rm=T))
```
</details>

### `unique` {#r-unique}

`unique` "returns a vector, data frame, or array like x but with duplicate elements/rows removed.

#### Given a vector of values, how do I return a vector of values with all duplicates removed?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vec <- c(1, 2, 3, 3, 3, 4, 5, 5, 6)
unique(vec)
```
</details>

### `order` and `sort` {#r-order}

`sort` allows you to arrange (or partially arrange) a vector into ascending or descending order.

`order` returns the position of each element of a vector in ascending (or descending order).

#### Examples

##### Given a vector, arrange it in a ascending order.

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1,3,2,10,4)
sort(x)
```
</details>

##### Given a vector, arrange it in a descending order.

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
x <- c(1,3,2,10,4)
sort(x, decreasing = TRUE)
```
</details>

##### Given a character vector, arrange it in ascending order.

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
sort(c("waffle", "pancake", "eggs", "bacon"))
```
</details>

##### Given a matrix, arrange it in ascending order using the first column.

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_mat <- matrix(c(1,5,0, 2, 10, 1, 2, 8, 9, 1,0,2), ncol=3)
my_mat[order(my_mat[,1]),]
```
</details>

### `paste` and `paste0` {#r-paste}

`paste` is a useful function to "concatenate vectors after converting to character."

`paste0` is a shorthand function where the `sep` argument is "".

#### How do I concatenate two vectors, element-wise, with a comma in between values from each vector?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
vector1 <- c("one", "three", "five")
vector2 <- c("two", "four", "six")
paste(vector1, vector2, sep=",")
```
</details>

#### How do I paste together two strings?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
paste0("abra", "kadabra")
```
</details>

#### How do I paste together three strings?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
paste0("abra", "kadabra", "alakazam")
```
</details>

### `str` {#r-str}

`str` stands for _structure_. `str` gives you a glimpse at the variable of interest. 

```{r, include=F}
df <- read.csv("test.csv")
```

#### How do I get the number of columns or features in a data.frame?

<details>
    <summary>Click here for solution</summary>
    
As you can see, there are 9 rows or obs. (short for observations), and 29 variables (which can be referred to as columns or features).
```{r, eval=F}
str(df)
```
</details>

### `names`

`names` is a function that returns the names of a an object. This includes the typical data structures: vectors, lists, and data.frames. By default, `names` will return the column names of a data.frame, not the row names.

```{r, include=F}
df <- data.frame(cat_1=c(1,2,3), cat_2=c(9,8,7), ok=c(T, T, F), other=c("first", "second", "third"))
```

#### Examples

##### How do I get the column names of a data.frame?

<details>
    <summary>Click here for solution</summary>
```{r}
# Get the column names of a data.frame
names(df)
```
</details>

##### How do I get the names of a list?

<details>
    <summary>Click here for solution</summary>
```{r}
# Get the names of a list
names(list(col1=c(1,2,3), col2=c(987)))
```
</details>

##### How do I get the names of a vector?

<details>
    <summary>Click here for solution</summary>
```{r}
# Get the names of a vector
names(c(val1=1, val2=2, val3=3))
```
</details>

##### How do I change the column names of a data.frame?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
names(df) <- c("col1", "col2", "col3", "col4")
df
```
</details>


### `colnames` & `rownames`

`colnames` is the same as `names` but specifies the column names. `rownames` is the same as `names` but specifies the row names.

### `table` & `prop.table`

`table` is a function used to build a [contingency table](https://en.wikipedia.org/wiki/Contingency_table) of counts of various factors.

`prop.table` is a function that accepts the output of `table` and rather than returning counts, returns conditional proportions.

#### Examples

```{r, include=F}
grades <- read.csv("grades.csv")
grades$sex <- factor(rep(c("M", "F", "M", "M", "F"), 2))
```

##### How do I get a count of the number of students in each year in our `grades` data.frame?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
table(grades$year)
```
</details>

##### How do I get the precentages of students in each year in our `grades` data.frame?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
prop.table(table(grades$year))
```
</details>

##### How do I get a count of the number of students in each year by sex in our `grades` data.frame?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
table(grades$year, grades$sex)
```
</details>

##### How do I get the precentages of students in each year by sex in our `grades` data.frame?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
prop.table(table(grades$year, grades$sex))
```
</details>

### `cut`

`cut` breaks a vector `x` into factors specified by the argument `breaks`. `cut` is particularly useful to break Date data into categories like "Q1", "Q2", or 1998, 1999, 2000, etc.

You can find more useful information by running `?cut.POSIXt`.

#### Examples

```{r, include=F}
times <- seq(as.POSIXct("2020-06-01 06:00"), by = "1 month", length.out = 24)
# times <- aggregate(1:24, list(day = cut(times, "days")), mean)
df <- data.frame(times=times)
df$value <- sample(1:100, size=length(df$times))
```

##### How can I create a new column in a data.frame `df` that is a factor based on the year?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df$year <- cut(df$times, breaks="year")
str(df)
```
</details>

##### How can I create a new column in a data.frame `df` that is a factor based on the quarter?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df$quarter <- cut(df$times, breaks="quarter")
str(df)
```
</details>

##### How can I create a new column in a data.frame `df` that is a factor based on every 2 weeks?

```{r, include=F}
times <- seq(as.POSIXct("2020-06-01 06:00"), by = "1 day", length.out = 45)
# times <- aggregate(1:24, list(day = cut(times, "days")), mean)
df <- data.frame(times=times)
df$value <- sample(1:100, size=length(df$times))
```

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df$biweekly <- cut(df$times, breaks="2 weeks")
```
</details>

### `subset`

`subset` is a function that helps you take subsets of data. By default, subset removes NA rows, so use with care. `subset` does not perform any operation that can't be accomplished by indexing, but can sometimes be easier to read.

Where we would normally write something like:

```{r}
grades[grades$year=="junior" | grades$sex=="M",]$grade
```

We can instead do:

```{r}
subset(grades, year=="junior" | sex=="M", select=grade)
```

But be careful, if we replace a grade with an NA, it will be removed by subset:

```{r}
grades$sex[8] <- NA
subset(grades, year=="junior" | sex=="M", select=grade)
```

Whereas indexing will not unless you specify to:

```{r}
grades[grades$year=="junior" | grades$sex=="M",]$grade
```
### `merge`

`merge` is a function that can be used to combine data.frames by row names, or more commonly, by column names. `merge` can replicate the join operations in SQL. The documentation is quite clear, and a useful resource: `?merge`.

```{r, include=F}
books <- data.frame(
  id = 1:10,
  title = c("Harry Potter and the Sorcerer's Stone", "Harry Potter and the Chamber of Secrets", "Harry Potter and the Prisoner of Azkaban", "Harry Potter and the Goblet of Fire", "Harry Potter and the Order of the Phoenix", "Harry Potter and the Half Blood Prince", "Harry Potter and the Deathly Hallows", "The Way of Kings", "The Book Thief", "The Eye of the World"),
  author_id = rep(c(1, 2, 3, 4), c(7, 1, 1, 1)),
  rating = c(4.47, 4.43, 4.57, 4.56, 4.5, 4.57, 4.62, 4.64, 4.37, 4.18)
)

authors <- data.frame(
  id = 1:10,
  name = c("J.K. Rowling", "Brandon Sanderson", "Markus Zusak", "Robert Jordan", "Agatha Christie", "Alex Kava", "Nassim Nicholas Taleb", "Neil Gaiman", "Stieg Larsson", "Antoine de Saint-Exupéry"),
  avg_rating = c(4.46, 4.39, 4.34, 4.18, 4.0, 4.02, 3.99, 4.13, 4.16, 4.3)
)
```

#### Examples

Consider the data.frame's `books` and `authors`:

```{r}
books
```

```{r}
authors
```

##### How do I merge the author information from `authors` based on `author_id` in `books` and `id` in `authors`, keeping only information from `authors` and `books` where there is a match?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# In SQL this is referred to as an INNER JOIN.
merge(books, authors, by.x="author_id", by.y="id", all=F)
```
</details>

##### How do I merge the author information from `authors` based on `author_id` in `books` and `id` in `authors`, keeping all information from `authors` regardless of whether or not there is match?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
merge(books, authors, by.x="author_id", by.y="id", all.y=T)

# or

merge(authors, books, by.x="id", by.y="author_id", all.x=T)
```
</details>

## Data.frames {#r-data-frames}

Data.frames are one of the primary data structure used very frequently when working in R. Data.frames are tables of same-sized, named columns, where each column has a single type.

You can create a data.frame easily:

```{r}
df <- data.frame(cat_1=c(1,2,3), cat_2=c(9,8,7), ok=c(T, T, F), other=c("first", "second", "third"))
head(df)
```

Regular indexing rules apply as well. This is how you index rows. Pay close attention to the trailing comma:

```{r}
# Numeric indexing on rows:
df[1:2,]
df[c(1,3),]
```

```{r}
# Logical indexing on rows:
df[c(T,F,T),]
```

```{r}
# Named indexing on rows only works 
# if there are named rows:
row.names(df) <- c("row1", "row2", "row3")
df[c("row1", "row3"),]
```

By default, if you don't include the comma in the square brackets, you are indexing the column:

```{r}
df[c("cat_1", "ok")]
```

To index columns, place expressions after the first comma:

```{r}
# Numeric indexing on columns:
df[, 1]
df[, c(1,3)]
```

```{r}
# Logical indexing on columns:
df[, c(T, F, F, F)]
```

```{r}
# Named indexing on columns.
# This is the more typical method of 
# column indexing:
df$cat_1
```

```{r}
# Another way to do named indexing on columns:
df[,c("cat_1", "ok")]
```

Of course, you can index on columns and rows:

```{r}
# Numeric indexing on columns and rows:
df[1:2, 1]
df[1:2, c(1,3)]
```

```{r}
# Logical indexing on columns and rows:
df[c(T,F,T), c(T, F, F, F)]
```

```{r}
# Named indexing on columns and rows.
# This is the more typical method of 
# column indexing:
df$cat_1[c(T,F,T)]
```

```{r}
# Another way to do named indexing on columns and rows:
row.names(df) <- c("row1", "row2", "row3")
df[c("row1", "row3"),c("cat_1", "ok")]
```

#### Examples

```{r, include=F}
df <- data.frame(cat_1=c(1,2,3), cat_2=c(9,8,7), ok=c(T, T, F), other=c("first", "second", "third"))
```

##### How can I get the first 2 rows of a data.frame named `df`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df <- data.frame(cat_1=c(1,2,3), cat_2=c(9,8,7), ok=c(T, T, F), other=c("first", "second", "third"))
df[1:2,]
```
</details>

##### How can I get the first 2 columns of a data.frame named `df`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df[,1:2]
```
</details>

##### How can I get the rows where values in the column named `cat_1` are greater than 2?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df[df$cat_1 > 2,]
```
```{r, eval=T}
df[df[, c("cat_1")] > 2,]
```

</details>

##### How can I get the rows where values in the column named `cat_1` are greater than 2 and the values in the column named `cat_2` are less than 9?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df[df$cat_1 > 2 & df$cat_2 < 9,]
```
</details>

##### How can I get the rows where values in the column named `cat_1` are greater than 2 or the values in the column named `cat_2` are less than 9?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df[df$cat_1 > 2 | df$cat_2 < 9,]
```
</details>

##### How do I sample _n_ rows randomly from a data.frame called `df`?
<details>
    <summary>Click here for solution</summary>
    
```{r, eval=F}
df[sample(nrow(df), n),]
```

Alternatively you could use the `sample_n` function from the package `dplyr`:
```{r, eval=F}
sample_n(df, n)
```
</details>

##### How can I get only columns whose names start with "cat_"?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
df <- data.frame(cat_1=c(1,2,3), cat_2=c(9,8,7), ok=c(T, T, F), other=c("first", "second", "third"))
df[, grep("^cat_", names(df))]
```
</details>

## Reading & Writing data {#r-reading-and-writing-data}

### Examples

#### How do I read a csv file called `grades.csv` into a data.frame? {r-read}

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
dat <- read.csv("./grades.csv")
head(dat)
```
</details>

#### How do I read a csv file called `grades2.csv` where instead of being comma-separated, it is semi-colon-separated, into a data.frame?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
dat <- read.csv("./grades_semi.csv", sep=";")
head(dat)
```
</details>

#### How do I prevent R from reading in strings as factors when using a function like `read.csv`?

<details>
    <summary>Click here for solution</summary>
In R 4.0+, strings are not read in as factors, so you do not need to do anything special. For R < 4.0, use `stringsAsFactors`.

```{r, eval=T}
dat <- read.csv("./grades.csv", stringsAsFactors=F)
head(dat)
```
</details>

#### How do I specify the type of 1 or more columns when reading in a csv file?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
dat <- read.csv("./grades.csv", colClasses=c("grade"="character", "year"="factor"))
str(dat)
```
</details>

#### Given a list of csv files with the same columns, how can I read them in and combine them into a single dataframe?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# We want to read in grades.csv, grades2.csv, and grades3.csv 
# into a single dataframe.

list_of_files <- c("grades.csv", "grades2.csv", "grades3.csv")

results <- data.frame()
for (file in list_of_files) {
  dat <- read.csv(file)
  results <- rbind(results, dat)
}
dim(results)
```
</details>

#### How do I create a data.frame with comma-separated data that I've copied onto my clipboard?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
# For mac
dat <- read.delim(pipe("pbpaste"),header=F,sep=",")

# For windows
dat <- read.table("clipboard",header=F,sep=",")
```
</details>

## Control flow {#r-control-flow}

### If/else statements {#r-if-else}

If, else if, and else statements are methods for controlling whether or not an operation is performed based on the result of some expression. 

#### How do I print "Success!" if my expression evaluates to `TRUE`, and "Failure!" otherwise?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Randomly assign either TRUE or FALSE to t_or_f.
t_or_f <- sample(c(TRUE,FALSE),1)

if (t_or_f == TRUE) {
  # If t_or_f is TRUE, print success
  print("Success!")
} else {
  # Otherwise, print failure
  print("Failure!")
}

# You don't need to put the full expression.
# This is the same thing because t_or_f
# is already TRUE or FALSE. 
# TRUE == TRUE evaluates to TRUE and
# FALSE == TRUE evaluates to FALSE.
if (t_or_f) {
  # If t_or_f is TRUE, print success
  print("Success!")
} else {
  # Otherwise, print failure
  print("Failure!")
}
```
</details>

#### How do I print "Success!" if my expression evaluates to `TRUE`, "Failure!" if my expression evaluates to `FALSE`, and "Huh?" otherwise?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Randomly assign either TRUE or FALSE to t_or_f.
t_or_f <- sample(c(TRUE,FALSE, "Something else"),1)

if (t_or_f == TRUE) {
  # If t_or_f is TRUE, print success
  print("Success!")
} else if (t_or_f == FALSE) {
  # If t_or_f is FALSE, print failure
  print("Failure!")
} else {
  # Otherwise print huh
  print("Huh?")
}

# In this case you need the full expression because
# "Something else" does not evaluate to TRUE or FALSE
# which will cause an error as the if and else if 
# statements expect a result of TRUE or FALSE.
if (t_or_f == TRUE) {
  # If t_or_f is TRUE, print success
  print("Success!")
} else if (t_or_f == FALSE) {
  # If t_or_f is FALSE, print failure
  print("Failure!")
} else {
  # Otherwise print huh
  print("Huh?")
}
```
</details>

### For loops {#r-for-loops}

For loops allow us to execute similar code over and over again until we've looped through all of the elements. They are useful for performing the same operation to an entire vector of input, for example. 

Using the suite of apply functions is more common in R. It is often said that the apply suite of function are much faster than for loops in R. While this used to be the case, this is no longer true. 

#### Examples

##### How do I loop through every value in a vector and print the value?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
for (i in 1:10) {
  # In the first iteration of the loop,
  # i will be 1. The next, i will be 2.
  # Etc.
  print(i)
}
```
</details>

##### How do I break out of a loop before it finishes?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
for (i in 1:10) {
  if (i==7) {
    # When i==7, we will exit the loop.
    break
  }
  print(i)
}
```
</details>

##### How do I loop through a vector of names?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
friends <- c("Phoebe", "Ross", "Rachel", "Chandler", "Joey", "Monica")
my_string <- "So no one told you life was gonna be this way, "
for (friend in friends) {
  print(paste0(my_string, friend, "!"))
}
```
</details>

##### How do I skip a loop if some expression evaluates to `TRUE`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
friends <- c("Phoebe", "Ross", "Mike", "Rachel", "Chandler", "Joey", "Monica")
my_string <- "So no one told you life was gonna be this way, "
for (friend in friends) {
  if (friend == "Mike") {
    # next, skips over the rest of the code for this loop
    # and continues to the next element
    next
  }
  print(paste0(my_string, friend, "!"))
}
```
</details>

## Apply functions {#r-apply-functions}

### Examples

```{r, include=F}
my_list <- list(
  pages = c(1,2,3,4,5),
  words = c(10,20,30,40,50),
  letters = c(100,200,300,400,500)
)
```

#### `apply` {#r-apply}

#### `lapply` {#r-lapply}

`lapply` is a function that applies a function `FUN` to each element in a vector or list, and returns a list.

##### How do I get the mean value of each vector in our list, `my_list`, in another list?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
lapply(my_list, mean)
```
</details>

#### `sapply` {#r-sapply}

`sapply` is very similar to `lapply`, however, where `lapply` always returns a list, `sapply` will simplify the output of applying the function `FUN` to each element. 

If you recall, when accessing an element in a list using single brackets `my_list[1]`, the result will always return a list. If you access an element with double brackets `my_list[[1]]`, `R` will attempt to simplify the result. This is analogous to `lapply` and `sapply`.

##### How do I get the mean value of each vector in our list, `my_list`, but rather than the result being a list, put the results in the simplest form?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
sapply(my_list, mean)
```
</details>

#### `tapply` {#r-tapply}

`tapply` is described in the documentation as a way to "apply a function to each cell of a ragged array, that is to each (non-empty) group of values given by a unique combination of the levels of certain factors." This is not a very useful description. 

An alternative way to think about `tapply`, is as a function that allows you to calculate or apply `function` to `data1` when `data1` is grouped by `data2`.

`tapply(data1, data2, function)`

A concrete example would be getting the _mean_ (`function`) _grade_ (`data1`) when _grade_ (`data1`) is grouped by _year_ (`data2`):

```{r, include=F}
grades <- read.csv("grades.csv")
```

```{r, eval=T}
grades
```

```{r}
tapply(grades$grade, grades$year, mean)
```

If your `function` (in this case _mean_), requires extra arguments, you can pass those by name to `tapply`. This is what the `...` argument in `tapply` is for. For example, if we want our _mean_ function to remove na's prior to calculating a mean we could do the following:

```{r}
tapply(grades$grade, grades$year, mean, na.rm=T)
```

## Writing functions {#r-writing-functions}

In a nutshell, a function is a set of instructions or actions packaged together in a single definition or unit. Typically, function accept 0 or more _arguments_ as input, and returns 0 or more results as output. The following is an example of a function in `R`:

```{r}
# word_count is a function that accepts a sentence as an argument,
# strips punctuation and extra space, and returns the number of 
# words in the sentence.
word_count <- function(sentence) {
  # strip punctuation and save into an auxiliary variable
  aux <- gsub('[[:punct:]]+','', sentence)
  
  # split the sentence by space and remove extra spaces
  result <- sum(unlist(strsplit(aux, " ")) != "")
  return(result)
}
test_sentence <- "this is a  sentence, with 7 words."
word_count(test_sentence)
```

The function is named `word_count`. The function has a single _parameter_ named `sentence`. The function returns a single value, `result`, which is the number of words in the provided sentence. `test_sentence` is the _argument_ to `word_count`. An _argument_ is the actual value passed to the function. We _pass_ values to functions -- this just means we use the values as _arguments_ to the function. The _parameter_, `sentence`, is the name shown in the function definition.

Functions can have helper functions. A helper function is a function defined and used within another function in order to reduce complexity or make the task at hand more clear. For example, we could have written the previous function differently:

```{r}
# word_count is a function that accepts a sentence as an argument,
# strips punctuation and extra space, and returns the number of 
# words in the sentence.
word_count <- function(sentence) {
  
  # a helper function that takes care of removing
  # punctuation and extra spaces.
  split_and_clean <- function(sentence) {
    # strip punctuation and save into an auxiliary variable
    aux <- gsub('[[:punct:]]+','', sentence)
    
    # remove extra spaces
    aux <- unlist(strsplit(aux, " "))
    
    return(aux[aux!=""])
  }
  
  # return the length of the sentence
  result <- length(split_and_clean(sentence))
  return(result)
}
test_sentence <- "this is a  sentence, with 7 words."
word_count(test_sentence)
```
Here, our helper function is named `split_and_clean`. If you try to call `split_and_clean` outside of `word_count`, you will get an error. `split_and_clean` is defined within the scope of `word_count` and is not available outside that scope. In this example, `word_count` is the _caller_, the function that _calls_ the other function, `split_and_clean`. The other function, `split_and_clean`, can be referred to as the `callee`.

In `R` functions can be passed to other functions as arguments. In general, functions that accept another function as an argument or return functions, are called higher order functions. Some examples of higher order functions in R are `sapply`, `lapply`, `tapply`, `Map`, and `Reduce`. The function passed as an argument, is often referred to as a _callback function_, as the _caller_ is expected to call back (execute) the argument at a later point in time.

### `...` {#r-ellipsis}

The ellipsis `...` in `R` can be used to pass an unknown number of arguments to a function. For example, if you look at the documentation for `sapply` (`?sapply`), you will see the following in the usage section:

```{r, eval=F}
sapply(X, FUN, ..., simplify = TRUE, USE.NAMES = TRUE)
```

In the arguments section, it says the ellipsis are "optional arguments to FUN". `sapply` uses the ellipsis as a vehicle to pass an unknown number of arguments to the callback function. In practice, this could look something like:

```{r}
dims <- function(..., sort=F) {
  args <- list(...)
  arg_names <- names(args)
  results <- lapply(args, dim)
  
  if (is.null(arg_names) | sort==FALSE) {
    # arguments not passed with a name
    return(results)
  }
  
  return(results[order(names(results))])
}

dims(grades)
dims(grades, mat)
dims(xyz=grades, abc=mat)
dims(xyz=grades, abc=mat, sort=T)
```

Here, `dims` accepts any number of data.frame-like objects, `...`, and a logical value indicating whether or not to sort the list by names. As you can see, if arguments are passed to `dims` with names, those names can be accessed within `dims` via `names(list(...))`. 

## Plotting {#r-plotting}

### `barplot` {#r-barplot}

### `boxplot` {#r-boxplot}

### `pie` {#r-pie}

### `dotchart` {#r-dotchart}

### `plot` {#r-plot}

### `lines` {#r-lines}

### `points` {#r-points}

### `abline` {#r-abline}

### `legend` {#r-legend}

### `plot_usmap` {#r-plot_usmap}

### `ggplot` {#r-ggplot}

### `ggmap` {#r-ggmap}

`ggmap` is an excellent package that provides a suite of functions that, among other things, allows you to map spatial data on top of static maps.

#### Getting started {#r-ggmap-getting-started}

To install `ggmap`, simply run `install.packages("ggmap")`. To load the library, run `library(ggmap)`. When first using this package, you may notice you need an API key to get access to certain functionality. Follow the directions [here](https://developers.google.com/maps/documentation/maps-static/get-api-key#get-key) to get an API key. It should looks somethings like: `mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg`. 

Once you've acquired the API key, you have two options:

1. Register `ggmap` with Google for the current session:

```{r, eval=F}
library(ggmap)
register_google(key="mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg")
```

2. Register `ggmap` with Google, persistently through sessions:

```{r, eval=F}
library(ggmap)
register_google(key="mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg", write=TRUE)
```

Note that if you choose option (2), your API key will be saved within your `~/.Renviron`.

#### Examples

##### How do I get a map of West Lafayette?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
map <- get_map(location="West Lafayette")
ggmap(map)
```
</details>

##### How do I zoom in and out on a map of West Lafayette?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
# zoom way out
map <- get_map(location="West Lafayette", zoom=1)
ggmap(map)

# zoom in
map <- get_map(location="West Lafayette", zoom=12)
ggmap(map)
```
</details>

##### How do I add Latitude and Longitude points to a map of Purdue University?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
points_to_add <- data.frame(latitude=c(40.433663, 40.432104, 40.428486), longitude=c(-86.916584, -86.919610, -86.920866))
map <- get_map(location="Purdue University", zoom=14)
ggmap(map) + geom_point(data = points_to_add, aes(x = longitude, y = latitude))
```
</details>

## RMarkdown {#r-rmarkdown}

To install RMarkdown simply run the following:

```r
install.packages("rmarkdown")
```

Projects in [The Data Mine](https://datamine.purdue.edu) are all written in RMarkdown. You can download the RMarkdown file by clicking on the link at the top of each project page. Each file should end in the ".Rmd" which is the file extension commonly associated with RMarkdown files.

You can find an exemplary RMarkdown file here:

https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/rmarkdown.Rmd

If you open this file in RStudio, and click on the "Knit" button in the upper left hand corner of IDE, you will get the resulting HTML file. Open this file in the web browser of your choice and compare and contrast the syntax in the `rmarkdown.Rmd` file and resulting output. Play around with the file, make modifications, and re-knit to gain a better understanding of the syntax. Note that similar input/output examples are shown in the [RMarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).

### Code chunks

Code chunks are sections within an RMarkdown file where you can write, display, and optionally evaluate code from a variety of languages:

```{r, echo=FALSE}
names(knitr::knit_engines$get())
```

The syntax is simple:

````markdown
`r ''````{language, options...}
code here...
```
````

For example:

````markdown
`r ''````{r, echo=TRUE}
my_variable <- c(1,2,3)
my_variable
```
````

Which will render like:

```{r, echo=TRUE}
my_variable <- c(1,2,3)
my_variable
```

You can find a list of chunk options [here](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf). 

#### How do I run a code chunk but not display the code above the results?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, echo=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I include a code chunk without evaluating the code itself?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, eval=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I prevent warning messages from being displayed?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, warning=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I prevent error messages from being displayed?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, error=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I run a code chunk, but not include the chunk in the final output?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, include=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I render a figure from a chunk?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r}
my_variable <- c(1,2,3)
plot(my_variable)
```
````
</details>

#### How do I create a set of slides using RMarkdown?

<details>
    <summary>Click here for solution</summary>
Please see the example Rmarkdown file [here](https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/example_slides.Rmd). 

You can change the slide format by changing the yaml header to any of: ioslides_presentation, slidy_presentation, or beamer_presentation.

By default all first and second level headers (\# and \##, respectively) will create a new slide. To manually create a new slide, you can use \*\*\*.
</details>

### Resources {#r-markdown-resources}

**[RMarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)**

An excellent quick reference for RMarkdown syntax.

**[RMarkdown Reference](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)**

A thorough reference manual showing markdown input and expected output. Gives descriptions of the various chunk options, as well as output options.

**[RStudio RMarkdown Lessons](https://rmarkdown.rstudio.com/lesson-1.html)**

A set of lessons detailing the ins and outs of RMarkdown.

**[Markdown Tutorial](https://www.markdowntutorial.com/)**

RMarkdown uses Markdown syntax for its text. This is a good, interactive tutorial to learn the basics of Markdown. This tutorial is available in multiple languages.

**[RMarkdown Gallery](https://rmarkdown.rstudio.com/gallery.html)**

This gallery highlights a variety of reproducible and interactive RMarkdown documents. An excellent resource to see the power of RMarkdown.

**[RMarkdown Chapter](https://r4ds.had.co.nz/r-markdown.html)**

This is a chapter from Hadley Wickham's excellent [_R for Data Science_](https://r4ds.had.co.nz/) book that details important parts of RMarkdown.

**[RMarkdown in RStudio](https://rmarkdown.rstudio.com/articles_intro.html)**

This is a nice article that introduces RMarkdown, and guides the user through creating their own interactive document using RMarkdown in RStudio.

**[Reproducible Research](https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html#reproducible_research)**

This is another good resource that introduces RMarkdown. Plenty of helpful pictures and screenshots.

## Tidyverse {#r-tidyverse}

### piping {#r-piping}

### `glimpse` {#r-glimpse}

### `filter` {#r-filter}

### `arrange` {#r-arrange}

### `mutate` {#r-mutate}

### `group_by` {#r-group-by}

### `str_extract` and `str_extract_all` {#r-str-extract}

`str_extract` and `str_extract_all` are useful functions from the `stringr` package. You can install the package by running:

```{r, eval=F}
install.packages("stringr")
```

`str_extract` extracts the text which matches the provided regular expression or pattern. Note that this differs from `grep` in a major way. `grep` simply returns the index in which a pattern match was found. `str_extract` returns the actual matching text. Note that `grep` typically returns the entire line where a match was found. `str_extract` returns only the part of the line or text that matches the pattern. 

For example:

```{r, eval=T}
text <- c("cat", "mat", "spat", "spatula", "gnat")

# All 5 "lines" of text were a match.
grep(".*at", text)
```

```{r, eval=T}
text <- c("cat", "mat", "spat", "spatula", "gnat")
stringr::str_extract(text, ".*at") 
```
As you can see, although all 5 words match our pattern and would be returned by `grep`, `str_extract` only returns the actual text that matches the pattern. In this case "spatula" is _not_ a "full" match -- the pattern ".\*at" only captures the "spat" part of "spatula". In order to capture the rest of the word you would need to add something like ".\*" to the end of the pattern:

```{r, eval=T}
text <- c("cat", "mat", "spat", "spatula", "gnat")
stringr::str_extract(text, ".*at.*") 
```
One final note is that you must double-escape certain characters in patterns because R treats backslashes as escape values for character constants ([stackoverflow](https://stackoverflow.com/questions/27721008/how-do-i-deal-with-special-characters-like-in-my-regex)). For example, to write `\(` we must first escape the `\`, so we write `\\(`. This is true for many character which would normally only be preceded by a single `\`.

#### Examples

##### How can I extract the text between parenthesis in a vector of texts?
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
text <- c("this is easy for (you)", "there (are) challenging ones", "text is (really awesome) (ok?)")

# Search for a literal "(", followed by any amount of any text other than more parenthesis ([^()]*), followed by a literal ")".
stringr::str_extract(text, "\\([^()]*\\)")
```

To get _all_ matches, not just the first match:
```{r, eval=T}
text <- c("this is easy for (you)", "there (are) challenging ones", "text is (really awesome) more text (ok?)")

# Search for a literal "(", followed by any amount of any text (.*), followed by a literal ")".
stringr::str_extract_all(text, "\\([^()]*\\)")
```
</details>

### `lubridate` {#r-lubridate}

`lubridate` is a fantastic package that makes the typical tasks one would perform on dates, that much easier.

#### How do I convert a string "07/05/1990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
library(lubridate)
dat <- "07/05/1990"
dat <- mdy(dat)
class(dat)
```
</details>

#### How do I convert a string "31-12-1990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_string <- "31-12-1990"
dat <- dmy(my_string)
dat
class(dat)
```
</details>

#### How do I convert a string "31121990" to a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_string <- "31121990"
my_date <- dmy(my_string)
my_date
class(my_date)
```
</details>

#### How do I extract the day, week, month, quarter, and year from a `Date`?

<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
my_date <- dmy("31121990")
day(my_date)
week(my_date)
month(my_date)
quarter(my_date)
year(my_date)
```
</details>

### Resources {#r-lubridate-resources}

**[Lubridate Cheatsheet](https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf)**

A comprehensive cheatsheet on `lubridate`. Excellent resource to immediately begin using `lubridate`.

## data.table {#r-datatable}

## SQL in R {#r-sql}

## Scraping {#r-scraping}

## `shiny` {#r-shiny}

### Rendering images {#r-shiny-render-images}


