# R {#r}

## Getting started {#getting-started-with-r}

How to install R (windows/mac/linux)
How to install RStudio
How to connect to RStudio Server on Scholar
How to get help (?, help(), get function itself)

## Variables {#r-variables}

## Logical operators {#r-logical-operators}

## Lists & Vectors {#r-lists-and-vectors}

## Basic R functions {#r-basic-functions}

### `length` {#r-length}

`length` allows you to get or set the length of an object in R (for which a method has been defined).

#### How do I get how many values are in a vector?
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
# Create a vector of length 5
my_vector <- c(1,2,3,4,5)

# Calculate the length of my_vector
length(my_vector)
```
</details>

### `grep`, `grepl`, etc. {#r-grep}

`grep` allows you to use regular expressions to search for a pattern in a string or character vector, and returns the index where there is a match.

`grepl` performs the same operation but rather than returning indices, returns a vector of logical `TRUE` or `FALSE` values.

#### Examples

##### Given a character vector, return the index of any words ending in "s".
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
grep("*.s$", c("waffle", "waffles", "pancake", "pancakes"))
```
</details>

##### Given a character vector, return a vector of the same length where each element is `TRUE` if there was a match for any word ending in "s", and `FALSE otherwise.
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
grepl("*.s$", c("waffle", "waffles", "pancake", "pancakes"))
```
</details>

### `str_extract` and `str_extract_all` {#r-str-extract}

`str_extract` and `str_extract_all` are useful functions from the `stringr` package. You can install the package by running:

```{r, eval=F}
install.packages("stringr")
```

`str_extract` extracts the text which matches the provided regular expression or pattern. Note that this differs from `grep` in a major way. `grep` simply returns the index in which a pattern match was found. `str_extract` returns the actual matching text. Note that `grep` typically returns the entire line where a match was found. `str_extract` returns only the part of the line or text that matches the pattern. 

For example:

```{r, eval=T}
text <- c("cat", "mat", "spat", "spatula", "gnat")

# All 5 "lines" of text were a match.
grep(".*at", text)
```

```{r, eval=T}
text <- c("cat", "mat", "spat", "spatula", "gnat")
stringr::str_extract(text, ".*at") 
```
As you can see, although all 5 words match our pattern and would be returned by `grep`, `str_extract` only returns the actual text that matches the pattern. In this case "spatula" is _not_ a "full" match -- the pattern ".\*at" only captures the "spat" part of "spatula". In order to capture the rest of the word you would need to add something like ".\*" to the end of the pattern:

```{r, eval=T}
text <- c("cat", "mat", "spat", "spatula", "gnat")
stringr::str_extract(text, ".*at.*") 
```
One final note is that you must double-escape certain characters in patterns because R treats backslashes as escape values for character constants ([stackoverflow](https://stackoverflow.com/questions/27721008/how-do-i-deal-with-special-characters-like-in-my-regex)). For example, to write `\(` we must first escape the `\`, so we write `\\(`. This is true for many character which would normally only be preceded by a single `\`.

#### Examples

##### How can I extract the text between parenthesis in a vector of texts?
<details>
    <summary>Click here for solution</summary>
```{r, eval=T}
text <- c("this is easy for (you)", "there (are) challenging ones", "text is (really awesome) (ok?)")

# Search for a literal "(", followed by any amount of any text other than more parenthesis ([^()]*), followed by a literal ")".
stringr::str_extract(text, "\\([^()]*\\)")
```

To get _all_ matches, not just the first match:
```{r, eval=T}
text <- c("this is easy for (you)", "there (are) challenging ones", "text is (really awesome) more text (ok?)")

# Search for a literal "(", followed by any amount of any text (.*), followed by a literal ")".
stringr::str_extract_all(text, "\\([^()]*\\)")
```
</details>

## Data.frames {#r-data-frames}

#### How do I sample _n_ rows randomly from a data.frame called `df`?
<details>
    <summary>Click here for solution</summary>
    
```{r, eval=F}
df[sample(nrow(df), n),]
```

Alternatively you could use the `sample_n` function from the package `dplyr`:
```{r, eval=F}
sample_n(df, n)
```
</details>

## Reading & Writing data {#r-reading-and-writing-data}

### Examples

#### How do I create a data.frame with comma-separated data that I've copied?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
# For mac
dat <- read.delim(pipe("pbpaste"),header=F,sep=",")

# For windows
dat <- read.table("clipboard",header=F,sep=",")
```
</details>

## Control flow {#r-control-flow}

## Apply functions {#r-apply-functions}

### `apply` {#r-apply}

### `sapply` {#r-sapply}

### `lapply` {#r-lapply}

### `tapply` {#r-tapply}

## Writing functions {#r-writing-functions}

## Plotting {#r-plotting}

### `ggplot` {#r-ggplot}

### `ggmap` {#r-ggmap}

`ggmap` is an excellent package that provides a suite of functions that, among other things, allows you to map spatial data on top of static maps.

#### Getting started {#r-ggmap-getting-started}

To install `ggmap`, simply run `install.packages("ggmap")`. To load the library, run `library(ggmap)`. When first using this package, you may notice you need an API key to get access to certain functionality. Follow the directions [here](https://developers.google.com/maps/documentation/maps-static/get-api-key#get-key) to get an API key. It should looks somethings like: `mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg`. 

Once you've acquired the API key, you have two options:

1. Register `ggmap` with Google for the current session:

```{r, eval=F}
library(ggmap)
register_google(key="mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg")
```

2. Register `ggmap` with Google, persistently through sessions:

```{r, eval=F}
library(ggmap)
register_google(key="mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg", write=TRUE)
```

Note that if you choose option (2), your API key will be saved within your `~/.Renviron`.

#### Examples

##### How do I get a map of West Lafayette?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
map <- get_map(location="West Lafayette")
ggmap(map)
```
</details>

##### How do I zoom in and out on a map of West Lafayette?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
# zoom way out
map <- get_map(location="West Lafayette", zoom=1)
ggmap(map)

# zoom in
map <- get_map(location="West Lafayette", zoom=12)
ggmap(map)
```
</details>

##### How do I add Latitude and Longitude points to a map of Purdue University?

<details>
    <summary>Click here for solution</summary>
```{r, eval=F}
points_to_add <- data.frame(latitude=c(40.433663, 40.432104, 40.428486), longitude=c(-86.916584, -86.919610, -86.920866))
map <- get_map(location="Purdue University", zoom=14)
ggmap(map) + geom_point(data = points_to_add, aes(x = longitude, y = latitude))
```
</details>

## RMarkdown {#r-rmarkdown}

To install RMarkdown simply run the following:

```r
install.packages("rmarkdown")
```

Projects in [The Data Mine](https://datamine.purdue.edu) are all written in RMarkdown. You can download the RMarkdown file by clicking on the link at the top of each project page. Each file should end in the ".Rmd" which is the file extension commonly associated with RMarkdown files.

You can find an exemplary RMarkdown file here:

https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/rmarkdown.Rmd

If you open this file in RStudio, and click on the "Knit" button in the upper left hand corner of IDE, you will get the resulting HTML file. Open this file in the web browser of your choice and compare and contrast the syntax in the `rmarkdown.Rmd` file and resulting output. Play around with the file, make modifications, and re-knit to gain a better understanding of the syntax. Note that similar input/output examples are shown in the [RMarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).

### Code chunks

Code chunks are sections within an RMarkdown file where you can write, display, and optionally evaluate code from a variety of languages:

```{r, echo=FALSE}
names(knitr::knit_engines$get())
```

The syntax is simple:

````markdown
`r ''````{language, options...}
code here...
```
````

For example:

````markdown
`r ''````{r, echo=TRUE}
my_variable <- c(1,2,3)
my_variable
```
````

Which will render like:

```{r, echo=TRUE}
my_variable <- c(1,2,3)
my_variable
```

You can find a list of chunk options [here](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf). 

#### How do I run a code chunk but not display the code above the results?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, echo=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I include a code chunk without evaluating the code itself?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, eval=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I prevent warning messages from being displayed?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, warning=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I prevent error messages from being displayed?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, error=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I run a code chunk, but not include the chunk in the final output?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r, include=FALSE}
my_variable <- c(1,2,3)
my_variable
```
````
</details>

#### How do I render a figure from a chunk?
<details>
    <summary>Click here for solution</summary>
````markdown
`r ''````{r}
my_variable <- c(1,2,3)
plot(my_variable)
```
````
</details>

#### How do I create a set of slides using RMarkdown?

<details>
    <summary>Click here for solution</summary>
Please see the example Rmarkdown file [here](https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/example_slides.Rmd). 

You can change the slide format by changing the yaml header to any of: ioslides_presentation, slidy_presentation, or beamer_presentation.

By default all first and second level headers (\# and \##, respectively) will create a new slide. To manually create a new slide, you can use \*\*\*.
</details>

### Resources {#r-markdown-resources}

**[RMarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)**

An excellent quick reference for RMarkdown syntax.

**[RMarkdown Reference](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)**

A thorough reference manual showing markdown input and expected output. Gives descriptions of the various chunk options, as well as output options.

**[RStudio RMarkdown Lessons](https://rmarkdown.rstudio.com/lesson-1.html)**

A set of lessons detailing the ins and outs of RMarkdown.

**[Markdown Tutorial](https://www.markdowntutorial.com/)**

RMarkdown uses Markdown syntax for its text. This is a good, interactive tutorial to learn the basics of Markdown. This tutorial is available in multiple languages.

**[RMarkdown Gallery](https://rmarkdown.rstudio.com/gallery.html)**

This gallery highlights a variety of reproducible and interactive RMarkdown documents. An excellent resource to see the power of RMarkdown.

**[RMarkdown Chapter](https://r4ds.had.co.nz/r-markdown.html)**

This is a chapter from Hadley Wickham's excellent [_R for Data Science_](https://r4ds.had.co.nz/) book that details important parts of RMarkdown.

**[RMarkdown in RStudio](https://rmarkdown.rstudio.com/articles_intro.html)**

This is a nice article that introduces RMarkdown, and guides the user through creating their own interactive document using RMarkdown in RStudio.

**[Reproducible Research](https://andrewbtran.github.io/NICAR/2017/reproducibility-rmarkdown/rmarkdown.html#reproducible_research)**

This is another good resource that introduces RMarkdown. Plenty of helpful pictures and screenshots.

## Tidyverse {#r-tidyverse}

## data.table {#r-datatable}

## SQL in R {#r-sql}

## Scraping {#r-scraping}

## `shiny` {#r-shiny}

### Rendering images {#r-shiny-render-images}


