# Projects {#projects}

## Templates

Our course project template can be found [here](https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/project_template.Rmd), or on Scholar: 

`/class/datamine/apps/templates/project_template.Rmd`

Students in STAT 19000, 29000, and 39000 are to use this as a template for all project submissions. The template includes a code chunk that "activates" our Python environment, and adjusts some default settings. In addition, it provides examples on how to include solutions for Python, R, Bash, and SQL. Every question should be clearly marked with a third-level header (using 3 `#`s) followed by `Question X` where `X` is the question number. Sections for question solutions should be added or removed based on the number of questions in the given project. All code chunks are to be run and solutions displayed for the compiled PDF submission.

Any format or template related questions should be asked in Piazza.

## STAT 19000

### Project 1 {#p01-190}

---

**Motivation:** In this project we are going to jump head first into The Data Mine. We will load datasets into the R environment, and introduce some core programming concepts like variables, vectors, types, etc. As we will be "living" primarily in an IDE called RStudio, we will take some time to learn how to connect to it, configure it, and run code.

**Context:** This is our first project as a part of The Data Mine. We will get situated, configure the environment we will be using throughout our time with The Data Mine, and jump straight into working with data!

**Scope:** r, rstudio, Scholar

**Learning objectives:**

```{block, type="bbox"}
- Utilize other Scholar resources: rstudio.scholar.rcac.purdue.edu, notebook.scholar.rcac.purdue.edu, desktop.scholar.rcac.purdue.edu, etc.
- Install R and setting up a working environment.
- Explain and demonstrate: positional, named, and logical indexing.
- Read and write basic (csv) data.
```

**Dataset:** `/class/datamine/data/disney/splash_mountain.csv`

##### 1. Read the webpage [here](https://www.rcac.purdue.edu/compute/scholar/). Scholar is the computing cluster you will be using throughout the semester, and your time with The Data Mine. Each _node_ is an individual machine with CPUs and memory (RAM). How many _cores_ and how much _memory_ is available, in total, for our 7 frontend nodes? How about for the sub-clusters? How much is available on your computer or laptop?

```{block, type="bbox"}
**Item(s) to submit:**

- A sentence explaining how much memory and how many cores your personal computer has.
- A sentence explaining how much memory and how many cores the 7 frontends have combined.
- A sentence explaining how much memory and how many cores the 28 sub-clusters have combined.
```

##### 2. Navigate and login to https://rstudio.scholar.rcac.purdue.edu using your Purdue Career Account credentials (without Boilerkey). This is an instance of RStudio Server running on a Scholar frontend! Frontends are labeled. So, for example, `scholar-fe01.rcac.purdue.edu` is frontend `#1`. In the lower left hand side of the RStudio screen, you should be able to see a tab labeled "Console". You can run R code by typing after the `>` and pressing enter. Check which frontend you are logged in on by running the following in the "Console" tab: `system("hostname")`. Which frontend are you in?

**Relevant topics:** running R code

```{block, type="bbox"}
**Item(s) to submit:**

- The `#` of the frontend your RStudio Server session is running on.
```

##### 3. From within RStudio, we can run every type of code that you will need to run throughout your time with The Data Mine: Python, R, Bash, SQL, etc. We've created a script for you to run to help configure settings. These settings will allow us to better serve you during the semester. Follow the directions [here](https://thedatamine.github.io/the-examples-book/python.html#python-on-scholar). Once complete, in RStudio (https://rstudio.scholar.rcac.purdue.edu), click on `Session > Restart R`. Once fully restarted, there should be a message that is printed in your "Console" tab. What does the message say?

```{block, type="bbox"}
**Item(s) to submit:**

- The sentence that is printed in the RStudio "Console".
```

##### 4. Projects in The Data Mine should all be submitted using our template found [here](https://raw.githubusercontent.com/TheDataMine/the-examples-book/master/files/project_template.Rmd) or on Scholar (`/class/datamine/apps/templates/project_template.Rmd`). At the beginning of every project, the first step should be downloading and/or copying and pasting the template into a `.Rmd` file in RStudio. Copy and paste the project template into a new RMarkdown file named `project01.Rmd`. Code chunks are parts of the RMarkdown file that contains code. You can identify what type of code a code chunk contains by looking at the _engine_ in the curly braces "{" and "}". How many of each type of code chunk are in our default template?

**Hint:** You can read about the template [here](https://thedatamine.github.io/the-examples-book/projects.html#templates). 

```{block, type="bbox"}
**Item(s) to submit:**

- A list containing the type of code chunk (r, Python, sql, etc), and how many of each code chunks our default template contains.
```

##### 5. Fill out the project template, replacing the default information with your own. If a category is not applicable to you, put N/A. This template provides examples of how to run each "type" of code we will run in this course. Look for the second R code chunk, and run it by clicking the tiny green play button in the upper right hand corner of the code chunk. What is the output?

```{block, type="bbox"}
**Item(s) to submit:**

- The output from running the R code chunk.
```

##### 6. In question (1) we answered questions about CPUs and RAM for the Scholar cluster. To do so, we needed to perform some arithmetic. Instead of using a calculator (or paper), write these calculations using R. Replace the content of the second R code chunk in our template with your calculations.

**Relevant topics:** [templates](https://thedatamine.github.io/the-examples-book/projects.html#templates)

```{block, type="bbox"}
**Item(s) to submit:**

- The R code chunk with your calculations, and output. 
```

##### 7. In (6) we got to see how you can type out arithmetic and R will calculate the result for you. One constant throughout the semester will be loading datasets into R. Load our dataset into R by running the following code:

```{r, eval=F}
dat <- read.csv("/class/datamine/data/disney/splash_mountain.csv")
```

Confirm the dataset has been read in by running the `head` function on it. `head` prints the first few rows of data:

```{r, eval=R}
head(dat)
```

`dat` is a variable which contains our data! We can name this variable anything we want, we do _not_ have to name it `dat`. Run our code to read in our dataset, this time, instead of naming our resulting dataset `dat`, name it `splash_mountain`. Place all of your code into a new R code chunk under a new level 3 header (i.e. `### Question 7`).

**Relevant topics:** [reading data in R](#r-reading-and-writing-data)

```{block, type="bbox"}
**Item(s) to submit:**

- Code used to answer this question in a code chunk in our template.
- Output of `head`.
```

##### 8. Let's pretend we are now done with our project. We've written some R code, maybe added some text explaining what we did, and we are ready to turn things in. For this course, we will turn in a variety of work, depending on the project. We will always require a PDF which contains text, code, and code output. Normally we would erase any code chunks from the template that are not used, however, for this project just keep that content. This PDF is generated by "Knitting" a PDF. In addition, if the project uses R code, you will need to copy and paste R code into an R script (file ending with `.R`). If you are submitting Python code too, you will need to copy and paste Python code into a Python script (file ending with `.py`). Let's practice. Compile your project to a PDF, create an R script and copy and paste all of your R code from your RMarkdown file (`.Rmd`), to a new file called `project01.R`. Include only the R code you wrote. Follow the directions in Brightspace to upload and submit your RMarkdown file, compiled PDF, and R script. 

**Relevant topics:** [templates](https://thedatamine.github.io/the-examples-book/projects.html#templates)

```{block, type="bbox"}
**Item(s) to submit:**

- Resulting knitted PDF.
- `project01.R` script containing all of your code from your R chunks in the `.Rmd` file.
```

---

### Project 2 {#p02-190}

---

**Motivation:** The R environment is a powerful tool to perform data analysis. R is a tool that is often compared to Python. Both have their advantages and disadvantages, and both are worth learning. In this project we will dive in head first and learn the basics while solving data-driven problems.

**Context:** Last project we set the stage for the rest of the semester. We got some familiarity with our project templates, and modified and ran some R code. In this project, we will continue to use R within RStudio to solve problems. Soon you will see how powerful R is and why it is often a more effective tool to use than spreadsheets.

**Scope:** r, vectors, indexing, recycling

**Learning objectives:**

```{block, type="bbox"}
- List the differences between lists, vectors, factors, and data.frames, and when to use each.
- Explain and demonstrate: positional, named, and logical indexing.
- Read and write basic (csv) data.
- Demonstrate the ability to use the following functions to solve data-driven problem(s): mean, var, table, cut, paste, rep, seq, sort, order, length, unique, etc.
- Explain what "recycling" is in R and predict behavior of provided statements.
```

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/disney/metadata.csv`

A public sample of the data can be found here: [/class/datamine/data/disney/metadata.csv](https://www.datadepot.rcac.purdue.edu/datamine/data/disney/metadata.csv)

##### 1. Use the `read.csv` function to load our dataset into a `data.frame` called `meta`. Note that `read.csv` _by default_ loads data into a `data.frame`. We will learn about `data.frame`s later, for now, print the first few rows of `meta`.

**Relevant topics:** reading data in r, head

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem in an R code chunk.
```

##### 2. We've provided you with R code below that will extract a column of our `data.frame` into a vector. What is the first value in the vector? What is the 50th value in the vector? That type of data is in the vector?

```{r, eval=F}
our_vec <- meta$WDWMAXTEMP
```

**Relevant topics:** indexing in r, type, [creating variables](#r-variables)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem in an R code chunk.
- The values of the first, and 50th element in the vector.
- The type of data in the vector.
```

##### 3. You can access many elements in a vector at the same time. Create three new vectors named: `first50`,  `last50`, and `mix`. `first50` should contain the first 50 values of our `our_vec` vector, and `last50` should contain the last 50 values. `mix` should contain the sum of each element of `first50` being added to each element of `last50`.

**Hint:** Make sure that `first50` and `last50` are the same length. If you get a warning message that reads "longer object length is not a multiple of shorter object length", you've done something wrong.

**Relevant topics:** indexing in r, [creating variables](#r-variables), [length](#r-length)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve this problem. 
- The `head` of each of the three vectors.
```

##### 4. In (3) we were able to rapidly add values together from two different vectors. Both vectors were the same size, hence, it was obvious which elements in each vector were added together. Create a new vector called `hot` which contains only the values which are greater than or equal to 80 (our vector contains max temperatures for days at Disney World). How many elements are in `hot`? Calculate the sum of `hot` and `first50`, do we get a warning? Read [this](https://excelkingdom.blogspot.com/2018/01/what-recycling-of-vector-elements-in-r.html) and then explain what is going on. 

**Relevant topics:** logical indexing, [length](#r-length), recycling

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve this problem. 
- 1-2 sentences explaining what is happening when we are adding two vectors of different lengths.
```

##### 5. Given what we learned in (4) how would we double every odd value in `hot`, and cut in half every even value in `hot`, in a single line of R code?

**Relevant topics:** recycling

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve this problem.
- `head` of the result.
```

##### 6. Run the code below in order to extract the `min_temp`, `max_temp`, and `mean_temp` vectors from our `data.frame`. Are the vectors all the same length? Create a new vector called `meanminmax` that contains the average of the `min_temp` and `max_temp`, element-wise. Calculate the absolute average difference between the `mean_temp` vector, and the `meanminmax` vector. What is the index of the largest difference between `mean_temp` and `meanminmax`? If you replace INDEX with the index you found, you will get information about the day: `meta[INDEX,]`. What is the largest difference between the `mean_temp` and `meanminmax` vectors? Explain whether or not that surprises you.

```{r, eval=F}
min_temp <- meta$WDWMINTEMP
mean_temp <- meta$WDWMEANTEMP
max_temp <- meta$WDWMAXTEMP
```

##### 7. The following code creates three graphics. The first two are created just using some core R functions, and the last is created using a package called `ggplot`. We will learn more about all of these things later on. For now, pick your favorite graphic, and write 1-2 sentences explaining why it is your favorite, what could be improved, and include any interesting observations (if any).

```{r, eval=F}
dat <- table(meta$SEASON)
names(dat) <- tolower(names(dat))
par(mar=c(1,8,1,1))
barplot(dat, main="Seasons", xlab="Number of Days in Each Season", las=2, horiz=TRUE, cex.names=.6)
```

```{r, eval=F}
dat <- tapply(meta$WDWMEANTEMP, meta$DAYOFYEAR, mean, na.rm=T)
seasons <- tapply(meta$SEASON, meta$DAYOFYEAR, function(x) unique(x)[1])
pal <- c("#4E79A7", "#F28E2B", "#A0CBE8",  "#FFBE7D", "#59A14F", "#8CD17D", "#B6992D", "#F1CE63", "#499894", "#86BCB6", "#E15759", "#FF9D9A", "#79706E", "#BAB0AC", "#1170aa", "#B07AA1")
colors <- factor(seasons)
levels(colors) <- pal
par(oma=c(7,0,0,0), xpd=NA)
barplot(dat, main="Average Temperature", xlab="Jan 1 (Day 0) - Dec 31 (Day 365)", ylab="Degrees in Fahrenheit", col=as.factor(colors), border = NA, space=0)
legend(-65, -50, legend=levels(factor(seasons)), lwd=5, col=pal, ncol=3, cex=0.8, box.col=NA)
```


```{r, eval=F}
library(ggplot2)
library(tidyverse)
summary_temperatures <- meta %>%
    select(MONTHOFYEAR,WDWMAXTEMP:WDWMEANTEMP) %>%
    group_by(MONTHOFYEAR) %>%
    summarise_all(mean, na.rm=T)
ggplot(summary_temperatures, aes(x=MONTHOFYEAR)) +
    geom_ribbon(aes(ymin = WDWMINTEMP, ymax = WDWMAXTEMP), fill = "#ceb888", alpha=.5) +
    geom_line(aes(y = WDWMEANTEMP), col="#5D8AA8") +
    geom_point(aes(y = WDWMEANTEMP), pch=21,fill = "#5D8AA8", size=2) +
    theme_classic() +
    labs(x = 'Month', y = 'Temperature', title = 'Average temperature range' ) +
    scale_x_continuous(breaks=1:12, labels=month.abb)
```

---

### Project 3 {#p03-190}

---

**Motivation:** `data.frame`s are the primary data structure you will work with when using R. It is important to understand how to insert, retrieve, and update data in a `data.frame`. 

**Context:** In the previous project we got our feet wet, and ran our first R code, and learned about accessing data inside vectors. In this project we will continue to reinforce what we've already learned and introduce a new, flexible data structure called `data.frame`s.

**Scope:** r, data.frames, recycling, factors

**Learning objectives:**

```{block, type="bbox"}
- Explain what "recycling" is in R and predict behavior of provided statements.
- Explain and demonstrate how R handles missing data: NA, NaN, NULL, etc.
- Demonstrate the ability to use the following functions to solve data-driven problem(s): mean, var, table, cut, paste, rep, seq, sort, order, length, unique, etc.
- Read and write basic (csv) data.
- Explain and demonstrate: positional, named, and logical indexing.
- List the differences between lists, vectors, factors, and data.frames, and when to use each.
```

**Dataset:** `/class/datamine/data/disney`

##### 1. Read the dataset `/class/datamine/data/disney/metadata.csv` into a data.frame called `meta`. Read the dataset `/class/datamine/data/disney/splash_mountain.csv` into a data.frame called `splash_mountain`. How many columns, or features are in each dataset? How many rows or observations?

**Relevant topics:** [str](#r-str)

```{block, type="bbox"}
**Item(s) to include:**

- R code used to solve the problem.
- How many columns or features in each dataset?
```

##### 2. Splash Mountain is a fan favorite ride at Disney World's Magic Kingdom theme park. `splash_mountain` contains a series of dates and datetimes. For each datetime, `splash_mountain` contains a posted minimum wait time, `SPOSTMIN`, and an actual minimum wait time, `SPOSTMIN`. What is the average minimum posted wait time for Splash Mountain? What is the standard deviation? Based on the fact that `SPOSTMIN` represents the posted minimum wait time for our ride, does our mean and standard deviation make sense? Explain.

**Hint:** If you got `NA` or `NaN` as a result, see [here](https://thedatamine.github.io/the-examples-book/r.html#r-mean).

**Relevant topics:** [mean](#r-mean), [var](#r-var), [NA](#r-na), [NaN](#r-nan)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve this problem.
- The results of running the R code. 
- 1-2 sentences explaining why or why not the results make sense.
```

##### 3. In (2) we got some peculiar values for the mean and standard deviation. If you read the "attractions" tab in the file `/class/datamine/data/disney/touringplans_data_dictionary.xlsx`, you will find that -999 is used as a value in `SPOSTMIN` and `SACTMIN` to indicate the ride as being closed. Recalculate the mean and standard deviation of `SPOSTMIN`, excluding values that are -999. Does this seem to have fixed our problem?

**Relevant topics:** [NA](#r-na), [mean](#r-mean), [var](#r-var), indexing, which

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve this problem.
- The result of running the R code.
- A statement indicating whether or not the value look reasonable now.
```

##### 4. `SPOSTMIN` and `SACTMIN` aren't the greatest feature/column names. An outsider looking at the data.frame wouldn't be able to immediately get the gist of what they represent. Change `SPOSTMIN` to `posted_min_wait_time` and `SACTMIN` to `actual_wait_time`. 

**Hint:** You can always use hard-coded integers to change names manually, however, if you use `which`, you can get the index of the column name that you would like to change. For data.frames like `meta`, this is a lot more efficient than manually counting which column is the one with a certain name.

**Relevant topics:** colnames, which

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The output from executing `names(splash_mountain)` or `colnames(splash_mountain)`.
```

##### 5. Use the `cut` function to create a new vector called `quarter` that breaks the `date` column up by quarter. Use the `labels` argument in the `factor` function to label the quarters "q1", "q2", ..., "qX" where `X` is the last quarter. Add `quarter` as a column named `quarter` in `splash_mountain`. How many

**Hint:** If you have 2 years of data, this will result in 8 quarters: "q1", ..., "q8".

**Hint:** I can generate sequential data using `seq` and `paste0`:

```{r, eval=T}
paste0("item", seq(1, 5))
```

**Relevant topics:** cut, [dates](#r-dates), [factor](#r-factors), [paste0](#r-paste), seq, [length](#r-length), [unique](#r-unique)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The `head` and `tail` of `splash_mountain`.
```

---

### Project 4 {#p04-190}

---

**Motivation:** Control flow is (roughtly) the order in which instructions are executed. We can execute certain tasks or code _if_ certain requirements are met using if/else statements. In addition, we can perform operations many times in a loop using for loops. While these are important concepts to grasp, R differs from other programming languages in that operations are usually vectorized and there is little to no need to write loops.  

**Context:** We are gaining familiarity working in RStudio and writing R code. In this project we introduce and practice using control flow in R. 

**Scope:** r, data.frames, recycling, factors, if/else, for

**Learning objectives:**

```{block, type="bbox"}
- Explain what "recycling" is in R and predict behavior of provided statements.
- Explain and demonstrate how R handles missing data: NA, NaN, NULL, etc.
- Demonstrate the ability to use the following functions to solve data-driven problem(s): mean, var, table, cut, paste, rep, seq, sort, order, length, unique, etc.
- Read and write basic (csv) data.
- Explain and demonstrate: positional, named, and logical indexing.
- List the differences between lists, vectors, factors, and data.frames, and when to use each.
- Demonstrate a working knowledge of control flow in r: if/else statements, while loops, etc.
```

**Dataset:** `/class/datamine/data/disney`

##### 1. In previous project we calculated the mean and standard deviation of the `SPOSTMIN` (posted minimum wait time). These are vectorized operations (we will learn more about this next project). Instead of using mean, use a loop to calculate the mean, just like the previous project. Do not use `sum` either.

**Hint:** Remember, if a value is NA, we don't want to include it.

**Hint:** Remember, if a value is -999, it means the ride is closed, we don't want to include it.

**Note:** This exercise should make you appreciate the variety of useful functions R has to offer!

**Relevant topics:** for loops, if/else statements, [is.na](#r-na)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The mean posted wait time.
```

##### 2. Choose one of the `.csv` files containing data for a ride. Use `read.csv` to load the file into a data.frame named `ride_name` where "ride_name" is the name of the ride you chose. Use a for loop to loop through the ride file and add a new column called `status`. `status` should contain a string whose value is either "open", or "closed". If `SPOSTMIN` or `SACTMIN` is -999, classify the row as "closed". Otherwise, classify the row as "open". After `status` is added to your data.frame, convert the column to a `factor`.

**Hint:** If you want to access two columns at once from a data.frame, you can do: `splash_mountain[i, c("SPOSTMIN", "SACTMIN")]`.

**Relevant topics:** any, for loops, if/else statements, [nrow](#r-data-frames)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The output from running `str` on `ride_name`.
```

##### 3. Typically you want to avoid using for loops (or even apply functions) when they aren't needed. Instead you can use vectorized operations and indexing. Repeat (2) without using any for loops or apply functions. Which method was faster?

**Hint:** To have multiple conditions within the `which` statement, use `|` for logical OR and `&` for logical AND.

**Hint:** You can start by assigning every value in `status` as "open", and then change the correct values to "closed".

**Relevant topics:** which

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The output from running `str` on `ride_name`.
```

##### 4. Create a pie chart for open vs. closed for `splash_mountain.csv`. First, use the `table` command to get a count of each `status`. Use the resulting table as input to the `pie` function. Make sure to give your pie chart a title that somehow indicates the ride to the audience.

**Relevant topics:** pie, table

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The resulting plot displayed as output in the RMarkdown.
```

##### 5. Loop through the vector of files we've provided below, and create a pie chart of open vs closed for each ride. Place all 6 resulting pie charts on the same image. Make sure to give each pie chart a title that somehow indicates the ride.

```{r, eval=F}
ride_names <- c("splash_mountain", "soarin", "pirates_of_caribbean", "expedition_everest", "flight_of_passage", "rock_n_rollercoaster")
ride_files <- paste0(c("/class/datamine/data/disney/"), ride_names, ".csv")
```

**Hint:** To place all of the resulting pie charts in the same image, prior to running the for loop, run `par(mfrow=c(2,3))`.

**Relevant topics:** for loop, read.csv, pie, table

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The resulting plot displayed as output in the RMarkdown.
```

---

### Project 5 {#p05-190}

---

**Motivation:** As briefly mentioned in project 4, R differs from other programming languages in that _typically_ you will want to avoid using for loops, and instead use vectorized functions and the apply suite. In this project we will demonstrate some basic vectorized operations, and how they are better to use than loops.

**Context:** While it was important to stop and learn about looping and if/else statements, in this project, we will explore the R way of doing things.

**Scope:** r, data.frames, recycling, factors, if/else, for

**Learning objectives:**

```{block, type="bbox"}
- Explain what "recycling" is in R and predict behavior of provided statements.
- Explain and demonstrate how R handles missing data: NA, NaN, NULL, etc.
- Demonstrate the ability to use the following functions to solve data-driven problem(s): mean, var, table, cut, paste, rep, seq, sort, order, length, unique, etc.
- Read and write basic (csv) data.
- Explain and demonstrate: positional, named, and logical indexing.
- List the differences between lists, vectors, factors, and data.frames, and when to use each.
- Demonstrate a working knowledge of control flow in r: if/else statements, while loops, etc.
- Demonstrate how apply functions are generally faster than using loops.
```

**Dataset:** `/class/datamine/data/fars`

To get more information on the dataset, see [here](https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/812602).

##### 1. The `fars` dataset contains a series of folders labeled by year. In each folder there is (at least) `ACCIDENT.CSV`, `PERSON.CSV`, and `VEHICLE.CSV`. If you take a peek at `ACCIDENT.CSV` you'll notice that the year isn't complete. Either add a new `YEAR` column with the _full_ year, or fix the variable some other way. Use a loop, and `rbind` to create a data.frame called `accidents`. As you are looping through each of the years (from [1975, 1981]), make sure to fix the `YEAR`.

**Relevant topics:** rbind, for loops, [read.csv](#r-str)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The result of `unique(accidents$year)`.
```

##### 2. How many accidents are there where 1+ drunk drivers were involved in an accident with a school bus?

**Hint:** Look at the variables `DRUNK_DR` and `SCH_BUS`.

**Relevant topics:** [table](#r-datatable)

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The result/answer itself.
```

##### 3. For accidents involving 1+ drunk drivers and a school bus, how many happened in each of the 7 years? Which year had the most qualifying accidents?

**Relevant topics:** [table](#r-datatable), which, indexing

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The results.
- Which year had the most qualifying accidents.
```

##### 4. Calculate the mean number of motorists involved in an accident (`PERSON`) with i drunk drivers for i in 0 through 6.

**Hint:** It is OK that there are no accidents involving just 5 drunk drivers.

**Relevant topics:** for loops, mean, indexing

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The output from running your code.
```

##### 5. Perhaps we have a theory that there are more accidents in cold weather months for Indiana and states around Indiana. Create a barplot that shows the number of accidents by `STATE` by month (`MONTH`). First, filter out all data where `STATE` is not one of: Indiana (18), Illinois (17), Ohio (39), or Michigan (26). What months have the most accidents? Are you surprised by these results? Explain why or why not? 

**Relevant topics:** %in%, barplot

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The output (plot) from running your code.
- 1-2 sentences explaining which month(s) have the most accidents and whether or not this surprises you.
```

##### 6. *(optional)* Spruce up your plot from (5). Do any of the following: 

- add vibrant (and preferably colorblind friendly) colors to your plot
- add a title
- add a legend
- add month names or abbreviations instead of numbers

**Hint:** [Here](https://www.r-graph-gallery.com/209-the-options-of-barplot.html) is a resource to get you started.

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to solve the problem.
- The output (plot) from running your code.
```

---

### Project 6 {#190-p06}

### Project 7 {#190-p07}

### Project 8 {#190-p08}

### Project 9 {#190-p09}

### Project 10 {#190-p10}

### Project 11 {#190-p11}

### Project 12 {#190-p12}

### Project 13 {#190-p13}

### Project 14 {#190-p14}

### Project 15 {#190-p15}

---

## STAT 29000

### Project 1 {#290-p01}

---

**Motivation:** In this project we will jump right into an R review. In this project we are going to break one larger data-wrangling problem into discrete parts. There is a slight emphasis on writing functions and dealing with strings. At the end of this project we will have greatly simplified a dataset, making it easy to dig into.

**Context:** We just started the semester and are digging into a large dataset, and in doing so, reviewing R concepts we've previously learned.

**Scope:** data wrangling in R, functions

**Learning objectives:**

```{block, type="bbox"}
- Comprehend what a function is, and the components of a function in R.
- Read and write basic (csv) data.
- Utilize apply functions in order to solve a data-driven problem.
```

#### Dataset: 

`/class/datamine/data/airbnb`

Often times (maybe even the majority of the time) data doesn’t come in one nice file or database. Explore the dataset in `/class/datamine/data/airbnb`.

##### 1. You may have noted that, for each country, city, and date we can find 3 files: `calendar.csv.gz`, `listings.csv.gz`, and `reviews.csv.gz` (for now, we will ignore all files in the "visualisations" folders). Let's take a look at the data in each of the three types of files. Pick a country, city and date, and read the first 50 rows of each of the 3 datasets. Provide 1-2 sentences explaining the type of information found in each, and what variable(s) could be used to join them. 

**Hint:** `read.csv` has an argument to select the number of rows we want to read.

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code used to read the first 50 rows of each dataset.
- 1-2 sentences briefly describing the information contained in each dataset.
- Name(s) of variable(s) that could be used to join them.
```

To read a compressed csv, simply use the `read.csv` function:

```{r, eval=F}
dat <- read.csv("/class/datamine/data/airbnb/brazil/rj/rio-de-janeiro/2019-06-19/data/calendar.csv.gz")
head(dat)
```

Let's work towards getting this data into an easier format to analyze. From now on, we will focus on the `listings.csv.gz` datasets.

##### 2. Write a function called `get_paths_for_country`, that, given a string with the country name, returns a vector with the full paths for all `listings.csv.gz` files, starting with `/class/datamine/data/airbnb/...`.

Some example output from `get_paths_for_country("united-states")`:

```{txt}
 [1] "/class/datamine/data/airbnb/united-states/ca/los-angeles/2019-07-08/data/listings.csv.gz"       
 [2] "/class/datamine/data/airbnb/united-states/ca/oakland/2019-07-13/data/listings.csv.gz"           
 [3] "/class/datamine/data/airbnb/united-states/ca/pacific-grove/2019-07-01/data/listings.csv.gz"     
 [4] "/class/datamine/data/airbnb/united-states/ca/san-diego/2019-07-14/data/listings.csv.gz"         
 [5] "/class/datamine/data/airbnb/united-states/ca/san-francisco/2019-07-08/data/listings.csv.gz"     
```

**Hint:** `list.files` is useful with the `recursive=T` option.

**Hint:** To exclude "visualisations" folders, try: `grep("visualisations", ..., invert=T)`.

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code for your `get_paths_for_country` function.
```

##### 3. Write a function called `get_data_for_country` that, given a string with the country name, returns a data.frames containing the all listings data for that country. Use your previously written function to help you. 

**Hint:** Use `stringAsFactors=F` in the `read.csv` function.

**Hint:** Use `do.call(rbind, <listofdataframes>)` to combine a list of dataframes into a single dataframe.

**Relevant topics:** rbind, [lapply](#r-lapply), [function](#r-basic-functions)

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code for your `get_data_for_country` function.
```

##### 4. Use your `get_data_for_country` to get the data for a country of your choice, and make sure to name the data.frame `listings`. Take a look at the following columns: `host_is_superhost`, `host_has_profile_pic`, `host_identity_verified`, and  `is_location_exact`. What is the data type for each column? Is there a more appropriate type for them? If so, which type would you recommend? 

**Hint:** Remember, there are a six types of vectors: logical, integer, double, character, complex, and raw. To see the vector data types of you can use `typeof` or `str`. The function `typeof` will return the type, while `str` will return more information. See some examples below:

```{r, eval=T}
# Using typeof
typeof(letters)
typeof(1:10)
```

```{r, eval=T}
# Using str
str(letters)
str(1:10)
```

##### 5. Write a function called `transform_column` that, given a column similar to the ones in (4) (more or less, lowercase "t"s and "f"s) transforms it to your suggested vector type in (4). Note that NA values for these columns appear as blank (`""`), and we need to be careful when transforming the data. Test your function on column `host_is_superhost`.

**Relevant topics:** toupper, as.logical

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code for your `transform_column` function.
- Type of `transform_column(listings$host_is_superhost)`.
```

##### 6. Before we can use your function, we need to determine which columns are similar to `host_is_superhost` (i.e., lowercase "t"s and "f"s) and need transformation. Create a function named `should_be_transformed` that, given a column, determines (returns `TRUE` or `FALSE`) if it contains only "t", "f", and "" values. Use your newly created function to create a vector named `columns_to_transform` which contains all columns we want to transform using the function in (5). How many columns are in this format?

**Relevant topics:** unique, %in%, any, [sapply](#r-sapply), which

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code for your `should_be_transformed` function.
- Chunk of code used to obtain `columns_to_transform`.
```

##### 7. Apply your function `transform_column` to all columns in `columns_to_transform` in your `listings` data. Make sure it worked by checking the type of columns `id` and `instant_bookable`. Note that the column `id` should have the same type as before.

**Relevant topics:** [apply](#r-apply)

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code to get your new `listings` data.
- Type of columns `id` and `instant_bookable`.
```

##### 8. Now that we have organized and cleaned our data, let's explore it! Based on your `listings` data, if you are looking at an instant bookable listing (where `instant_bookable` is `TRUE`), would you expect the location to be exact (where `is_exact_location` is `TRUE`)? Why or why not?

**Hint:** Make a frequency table, and see how many instant bookable listings have exact location.

**Relevant topics:** table

```{block, type="bbox"}
**Item(s) to submit:**

- Chunk of code to get a frequency table.
- 1-2 sentences explaining whether or not we would expect the location to be exact if we were looking at a instant bookable listing.
```

---

### Project 2 {#290-p02}

---

**Motivation:** The ability to quickly reproduce an analysis is important. It is often necessary that other individuals will need to be able to understand and reproduce an analysis. This concept is so important there are classes solely on reproducible research! In fact, there are papers that investigate and highlight the lack of reproducibility in various fields. If you are interested in reading about this topic, a good place to start is the paper titled ["Why Most Published Research Findings Are False"](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.0020124), by John Ioannidis (2005). 

**Context:** Making your work reproducible is extremely important. We will focus on the computational part of reproducibility. We will learn RMarkdown to document your analyses so others can easily understand and reproduce the computations that led to your conclusions. Pay close attention as future project templates will be RMarkdown templates.

**Scope:** Understand Markdown, RMarkdown, and how to use it to make your data analysis reproducible.

**Learning objectives:**

```{block, type="bbox"}
- Use Markdown syntax within an Rmarkdown document to achieve various text transformations.
- Use RMarkdown code chunks to display and/or run snippets of code.
```

You can find useful examples that walk you through relevant material in The Examples Book:

https://thedatamine.github.io/the-examples-book

It is highly recommended to read through, search, and explore these examples to help solve problems in this project.

**Important note:** It is highly recommended that you use https://rstudio.scholar.rcac.purdue.edu/. Simply click on the link and login using your Purdue account credentials. Use another system at your own risk. The version of RStudio on https://desktop.scholar.rcac.purdue.edu/ (which uses ThinLinc), is 99.9.9, and is known to have some strange issues when running code chunks.

Don't forget the very useful documentation shortcut `?`. To use, simply type `?` in the console, followed by the name of the function you are interested in. 

You can also look for package documentation by using `help(package=PACKAGENAME)`, so for example, to see the documentation for the package `ggplot2`, we could run:

```{r, eval=F}
help(package=ggplot2)
```

Sometimes it can be helpful to see the source code of a defined function. A [function](https://www.tutorialspoint.com/r/r_functions.htm) is any chunk of organized code that is used to perform an operation. Source code is the underlying `R` or `c` or `c++` code that is used to create the function. To see the source code of a defined function, type the function's name without the `()`. For example, if we were curious about what the function `Reduce` does, we could run:

```{r, eval=F}
Reduce
```

Occasionally this will be less useful as the resulting code will be code that calls `c` code we can't see. Other times it will allow you to understand the function better.

##### 1. Make the following text (including the asterisks) bold: `This needs to be **very** bold`. Make the following text (including the underscores) italicized: `This needs to be _very_ italicized.`

**Hint:** *Try mixing and matching ways to embolden or italicize text. Alternatively, look up "escaping characters in markdown", or see [here](https://thedatamine.github.io/the-examples-book/faqs.html#escape-characters).*

**Hint:** *Be sure to check out the [Rmarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) and our section on [Rmarkdown in the book](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown).*

**Note:** *Rmarkdown is essentially Markdown + the ability to run and display code chunks. In this question, we are actually using Markdown within Rmarkdown!*

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown), [escaping characters](https://thedatamine.github.io/the-examples-book/faqs.html#escape-characters)*

```{block, type="bbox"}
**Item(s) to submit:**

- Fill in the chunk under (1) in the `stat29000project02template.Rmd` file with 2 lines of markdown text. Note that when compiled, this text will be unmodified, regular text.
```

##### 2. Create an unordered list of your top 3 favorite academic interests (some examples could include: machine learning, operating systems, forensic accounting, etc.). Create another *ordered* list that ranks your academic interests in order of most interested to least interested.

**Hint:** *You can learn what ordered and unordered lists are [here]( https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).*

**Note:** *Similar to (1a), in this question we are dealing with Markdown. If we were to copy and paste the solution to this problem in a Markdown editor, it would be the same result as when we Knit it here.*

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- Create the lists under (2) in the `stat29000project02template.Rmd` file. Note that when compiled, this text will appear as nice, formatted lists.
```

##### 3. Browse https://www.linkedin.com/ and read some profiles. Pay special attention to accounts with an "About" section. Write your own personal "About" section using Markdown. Include the following:

- A header (your choice of size) that says "About".
- A horizontal rule directly underlining the header.
- The text of your personal "About" section that you would feel comfortable uploading to linkedin, including at least 1 link.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- Create the described profile under (3) in the `stat29000project02template.Rmd` file.
```

##### 4. Your co-worker wrote a report, and has asked you to beautify it. Knowing Rmarkdown, you agreed. Spruce up the report found under (4) in `stat29000project02template.Rmd`. At a minimum:

- Make the title pronounced.
- Make all links appear as a word or words, rather than the long-form URL.
- Organize all code into code chunks where code and output are displayed. If the output is really long, just display the code.
- Make the calls to the `library` function and the `install.packages` function be evaluated but not displayed. Make sure all warnings and errors that may eventually occur, do not appear in the final document.

Feel free to make any other changes that make the report more visually pleasing.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- Spruce up the "document" under (4) in the `stat29000project02template.Rmd` file.
```

##### 5. Create a plot using a built-in dataset like `iris`, `mtcars`, or `Titanic`, and display the plot using a code chunk. Make sure the code used to generate the plot is hidden. Include a descriptive caption for the image.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown), [plotting in r](https://thedatamine.github.io/the-examples-book/r.html#r-plotting)*

```{block, type="bbox"}
**Item(s) to submit:**

- Code chunk under (5) that creates and displays a plot using a built-in dataset like `iris`, `mtcars`, or `Titanic`.
```

##### 6. Insert the following code chunk under (5) in `stat29000project02template.Rmd`. Try knitting the document. Something should go wrong. Fix the problem and knit again. If another problem appears, fix it. What was the first problem? What was the second problem?
       
````markdown
`r ''````{r install-packages}
plot(my_variable)
```
````

**Hint:** *Take a close look at the name we give our code chunk.*

**Hint:** *Take a look at the code chunk where `my_variable` is declared.*

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- The modified version of the inserted code that fixes both problems.
- A sentence explaining what the first problem was.
- A sentence explaining what the second problem was.
```

---

### Project 3 {#290-p03}

---

**Motivation:** The ability to navigate a shell, like `bash`, and use some of its powerful tools, is very useful. The number of disciplines utilizing data in new ways is ever-growing, and as such, it is very likely that many of you will eventually encounter a scenario where knowing your way around a terminal will be useful. We want to expose you to some of the most useful `bash` tools, help you navigate a filesystem, and even run `bash` tools from within an RMarkdown file in RStudio.

**Context:** At this point in time, you will each have varying levels of familiarity with Scholar. In this project we will learn how to use the terminal to navigate a UNIX-like system, experiment with various useful commands, and learn how to execute bash commands from within RStudio in an RMarkdown file.

**Scope:** bash, RStudio

**Learning objectives:**

```{block, type="bbox"}
- Distinguish differences in /home, /scratch, and /class.
- Navigating UNIX via a terminal: ls, pwd, cd, ., .., ~, etc.
- Analyzing file in a UNIX filesystem: wc, du, cat, head, tail, etc.
- Creating and destroying files and folder in UNIX: scp, rm, touch, cp, mv, mkdir, rmdir, etc.
- Utilize other Scholar resources: rstudio.scholar.rcac.purdue.edu, notebook.scholar.rcac.purdue.edu, desktop.scholar.rcac.purdue.edu, etc.
- Use `man` to read and learn about UNIX utilities.
- Run `bash` commands from within and RMarkdown file in RStudio.
```

#### Dataset: ??

#### Public: ??

There are a variety of ways to connect to Scholar. In this class, we will _primarily_ connect to RStudio Server by opening a browser and navigating to https://rstudio.scholar.rcac.purdue.edu/, entering credentials, and using the excellent RStudio interface. 

##### 1. Navigate to https://rstudio.scholar.rcac.purdue.edu/ and login. Take some time to click around and explore this tool. We will be writing and running Python, R, SQL, and `bash` all from within this interface. Navigate to `Tools > Global Options ...`. Explore this interface and make at least 2 modifications. List what you changed.

Here are some changes Kevin likes:

- Uncheck "Restore .Rdata into workspace at startup".
- Change tab width 4.
- Check "Soft-wrap R source files".
- Check "Highlight selected line".
- Check "Strip trailing horizontal whitespace when saving".
- Uncheck "Show margin".

```{block, type="bbox"}
**Item(s) to submit:**

- List of modifications you made to your Global Options.
```

##### 2. There are four primary panes, each with various tabs. In one of the panes there will be a tab labeled "Terminal". Click on that tab. This terminal by default will run a `bash` shell right within Scholar, the same as if you connected to Scholar using ThinLinc, and opened a terminal. Very convenient! What is the default directory of your bash shell? In our list of relevant topics, we've included links to a variety of UNIX commands that may help you solve this problem. Some of the tools are super simple to use, and some are a little bit more difficult. 

**Hint:** Start by reading the section on `man`. `man` stands for manual, and you can find the "official" documentation for the command by typing `man <command_of_interest>`. For example:

```{bash, eval=F}
# read the manual for the `man` command
# use "k" or the up arrow to scroll up, "j" or the down arrow to scroll down
man man 
```

**Relevant topics:** [man](#man), [cd](#cd), [pwd](#pwd), [ls](#ls), [~](#dots), [..](#dots), [.](#dots)

```{block, type="bbox"}
**Item(s) to submit:**

- The full filepath of default directory (home directory). Ex: Kevin's is: `/home/kamstut`
- The `bash` code used to show your home directory or current working directory when the `bash` shell is first launched.
```

##### 3. Learning to navigate away from our home directory to other folders, and back again, is vital. Perform the following actions, in order:

- Write a single command to navigate to the folder containing our full datasets: `/class/datamine/data`.
- Write a command to confirm you are in the correct folder. 
- Write a command to list the files and directories within the data directory. 
- What are the names of the files? Write another command to return back to your home directory. 
- Write a command to confirm you are in the correct folder.

Note: `/` is commonly referred to as the root directory in a linux/unix filesystem. Think of it as a folder that contains _every_ other folder in the computer. `/home` is a folder within the root directory. `/home/kamstut` is the full filepath of Kevin's home directory. There is a folder `home` inside the root directory. Inside `home` is another folder named `kamstut` which is Kevin's home directory. 

**Relevant topics:** [man](#man), [cd](#cd), [pwd](#pwd), [ls](#ls), [~](#dots), [..](#dots), [.](#dots)

```{block, type="bbox"}
**Item(s) to submit:**

- Command used to navigate to the data directory.
- Command used to confirm you are in the data directory.
- Command used to list files and folders.
- List of files and folders in the data directory.
- Command used to navigate back to the home directory.
- Commnad used to confirm you are in the home directory.
```

##### 4. Let's learn about two more important concepts. `.` refers to the current working directory, or the directory displayed when you run `pwd`. Unlike `pwd` you can use this when navigating the filesystem! So, for example, if you wanted to see the contents of a file called `my_file.txt` that lives in `/home/kamstut` (so, a full path of `/home/kamstut/my_file.txt`), and you are currently in `/home/kamstut`, you could run: `cat ./my_file.txt`. `..` represents the parent folder or the folder in which your current folder is contained. So let's say I was in `/home/kamstut/projects/` and I wanted to get the contents of the file `/home/kamstut/my_file.txt`. You could do: `cat ../my_file.txt`. When you navigate a directory tree using `.`,  `..`, and `~` you create paths that are called _relative_ paths because they are _relative_ to your current directory. Alternatively, a _full_ path or (_absolute_ path) is the path starting from the root directory. So `/home/kamstut/my_file.txt` is the _absolute_ path for `my_file.txt` and `../my_file.txt` is a _relative_ path. Perform the following actions, in order:

- Write a single command to navigate to the data directory.
- Write a single command to navigate back to your home directory using a _relative_ path. Do not use `~` or plain `cd`.

**Relevant topics:** [man](#man), [cd](#cd), [pwd](#pwd), [ls](#ls), [~](#dots), [..](#dots), [.](#dots)

```{block, type="bbox"}
**Item(s) to submit:**

- Command used to navigate to the data directory.
- Command used to navigate back to your home directory that uses a _relative_ path.
```

##### 5. In Scholar, when you want to deal with _really_ large amounts of data, you want to access scratch (you can read more [here](https://www.rcac.purdue.edu/policies/scholar/)). Your scratch directory on Scholar is located here: `/scratch/scholar/$USER`. `$USER` is an environment variable containing your username. Test it out: `echo /scratch/scholar/$USER`. Perform the following actions:

- Navigate to your scratch directory. 
- Confirm you are in the correct location.
- Execute `myquota`.
- Find the location of the `myquota` bash script.
- Output the first 5 and last 5 lines of the bash script. 
- Count the number of lines in the bash script.
- How many kilobytes is the script?

**Hint:** You could use each of the commands in the relevant topics once.

**Hint:** Commands often have _options_. _Options_ are features of the program that you can trigger specifically. You can see the _options_ of a command in the `DESCRIPTION` section of the `man` pages. For example: `man wc`. You can see `-m`, `-l`, and `-w` are all options for `wc`. To test this out:

```{bash, eval=F}
# using the default wc command. "/class/datamine/data/flights/1987.csv" is the first "argument" given to the command.
wc /class/datamine/data/flights/1987.csv
# to count the lines, use the -l option
wc -l /class/datamine/data/flights/1987.csv
# to count the words, use the -w option
wc -w /class/datamine/data/flights/1987.csv
# you can combine options as well
wc -w -l /class/datamine/data/flights/1987.csv
# some people like to use a single tack `-`
wc -wl /class/datamine/data/flights/1987.csv
# order doesn't matter
wc -lw /class/datamine/data/flights/1987.csv
```

**Hint:** The `-h` option for the `du` command is useful.

**Relevant topics:** [cd](#cd), [pwd](#pwd), [type](#type), [head](#head), [tail](#tail), [wc](#wc), [du](#du)

```{block, type="bbox"}
**Item(s) to submit:**

- Command used to navigate to your scratch directory.
- Command used to confirm your location.
- Output of `myquota`.
- Command used to find the location of the `myquota` script.
- Absolute path of the `myquota` script.
- Command used to output the first 5 lines of the `myquota` script.
- Command used to output the last 5 lines of the `myquota` script.
- Command used to find the number of lines in the `myquota` script.
- Number of lines in the script.
- Command used to find out how many kilobytes the script is.
- Number of kilobytes that the script takes up.
```

##### 6. Perform the following operations:

- Navigate to your scratch directory.
- Copy and paste the file: `/class/datamine/data/flights/1987.csv` to your current directory (scratch).
- Create a new directory called `my_test_dir` in your scratch folder.
- Move the file you copied to your scratch directory, into your new folder.
- Use `touch` to create an empty file named `im_empty.txt` in your scratch folder.
- Remove the directory `my_test_dir` _and_ the contents of the directory.
- Remove the `im_empty.txt` file.

**Hint:** `rmdir` may not be able to do what you think, instead, check out the options for `rm` using `man rm`.

**Relevant topics:** [cd](#cd), [cp](#cp), [mv](#mv), [mkdir](#mkdir), [touch](#touch), [rmdir](#rmdir), [rm](#rm)

```{block, type="bbox"}
**Item(s) to submit:**

- Command used to navigate to your scratch directory.
- Command used to copy the file, `/class/datamine/data/flights/1987.csv` to your current directory (scratch).
- Command used to create a new directory called `my_test_dir` in your scratch folder.
- Command used to move the file you copied earlier `1987.csv` into your new `my_test_dir` folder.
- Command used to create an empty file named `im_empty.txt` in your scratch folder.
- Command used to remove the directory _and_ the contents of the directory `my_test_dir`.
- Command used to remove the `im_empty.txt` file.
```

---

### Project 4 {#290-p04}

---

**Motivation:** The need to search files and datasets based on the text held within is common during various parts of the data wrangling process. `grep` is an extremely powerful UNIX tool that allows you to do so using regular expressions. Regular expressions are a structured method for searching for specified patterns. Regular expressions can be very complicated, [even professionals can make critical mistakes](https://blog.cloudflare.com/details-of-the-cloudflare-outage-on-july-2-2019/). With that being said, learning some of the basics is an incredible tool that will come in handy regardless of the language you are working in.

**Context:** We've just begun to learn the basics of navigating a file system in UNIX using various terminal commands. Now we will go into more depth with one of the most useful command line tools, `grep`, and experiment with regular expressions using `grep`, R, and later on, Python.

**Scope:** grep, regular expression basics, utilizing regular expression tools in R and Python

**Learning objectives:**

```{block, type="bbox"}
- Use `grep` to search for patterns within a dataset.
- Use `cut` to section off and slice up data from the command line.
- Use `wc` to count the number of lines of input.
```


You can find useful examples that walk you through relevant material in The Examples Book:

https://thedatamine.github.io/the-examples-book

It is highly recommended to read through, search, and explore these examples to help solve problems in this project.

**Important note:** It is highly recommended that you use https://rstudio.scholar.rcac.purdue.edu/. Simply click on the link and login using your Purdue account credentials. Use another system at your own risk. The version of RStudio on https://desktop.scholar.rcac.purdue.edu/ (which uses ThinLinc), is 99.9.9, and is known to have some strange issues when running code chunks.

Don't forget the very useful documentation shortcut `?`. To use, simply type `?` in the console, followed by the name of the function you are interested in. 

You can also look for package documentation by using `help(package=PACKAGENAME)`, so for example, to see the documentation for the package `ggplot2`, we could run:

```{r, eval=F}
help(package=ggplot2)
```

Sometimes it can be helpful to see the source code of a defined function. A [function](https://www.tutorialspoint.com/r/r_functions.htm) is any chunk of organized code that is used to perform an operation. Source code is the underlying `R` or `c` or `c++` code that is used to create the function. To see the source code of a defined function, type the function's name without the `()`. For example, if we were curious about what the function `Reduce` does, we could run:

```{r, eval=F}
Reduce
```

Occasionally this will be less useful as the resulting code will be code that calls `c` code we can't see. Other times it will allow you to understand the function better.

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/movies_and_tv/the_office_dialogue.csv`

A public sample of the data can be found here: [the_office_dialogue.csv](https://www.datadepot.rcac.purdue.edu/datamine/movies-and-tv/the_office_dialogue.csv)

Answers to questions should all be answered using the full dataset located on Scholar. You may use the public samples of data to experiment with your solutions prior to running them using the full dataset.

`grep` stands for (g)lobally search for a (r)egular (e)xpression and (p)rint matching lines. As such, to best demonstrate `grep`, we will be using it with textual data. You can read about and see examples of `grep` [here](https://thedatamine.github.io/the-examples-book/unix.html#grep).

##### 1. Login to Scholar and use `grep` to find the dataset we will use this project. The dataset we will use is the only dataset to have the text "bears. beets. battlestar galactica.". What is the name of the dataset and where is it located?

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The `grep` command used to find the dataset.
- The name and location in Scholar of the dataset.
- Use `grep` and `grepl` within R to solve a data-driven problem.
```

##### 2. `grep` prints the line that the text you are searching for appears in. In project 3 we learned a UNIX command to quickly print the first _n_ lines from a file. Use this command to get the headers for the dataset. As you can see, each line in the tv show is a row in the dataset. You can count to see which column the various bits of data live in.

Write a line of UNIX commands that searches for "bears. beets. battlestar galactica." and, rather than printing the entire line, prints only the character who speaks the line, as well as the line itself.

**Hint:** *The result if you were to search for "bears. beets. battlestar galactica." should be:*

```{txt}
"Jim","Fact. Bears eat beets. Bears. Beets. Battlestar Galactica."
```

**Hint:** *One method to solve this problem would be to [pipe](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection) the output from `grep` to [`cut`](https://thedatamine.github.io/the-examples-book/unix.html#cut).*

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to perform the operation.
```

##### 3. This particular dataset happens to be very small. You could imagine a scenario where the file is many gigabytes and not easy to load completely into R or Python. We are interested in learning what makes Jim and Pam tick as a couple. Use a line of UNIX commands to create a new dataset called `jim_and_pam.csv`. Include only lines that are spoken by either Jim or Pam, or reference Jim or Pam in any way. Include only the following columns: `episode_name`, `character`, `text`, `text_w_direction`, and `air_date`. How many rows of data are in the new file? How many megabytes is the new file (to the nearest 1/10th of a megabyte)?

**Hint:** *[Redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection).*

**Hint:** *It is OK if you get an erroneous line where the word "jim" or "pam" appears as a part of another word.*

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [grep](https://thedatamine.github.io/the-examples-book/unix.html#grep), [ls](https://thedatamine.github.io/the-examples-book/unix.html#ls), [wc](https://thedatamine.github.io/the-examples-book/unix.html#wc), [redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to create the new file. 
- The number of rows of data in the new file, and the accompanying UNIX command used to find this out.
- The number of megabytes (to the nearest 1/10th of a megabyte) that the new file has, and the accompanying UNIX command used to find this out.
```

##### 4. Find all lines where either Jim/Pam/Michael/Dwight's name is followed by an exclamation mark. Use only 1 "!" within your regular expression. How many lines are there?

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The UNIX command(s) used to solve this problem.
- The number of lines where either Jim/Pam/Michael/Dwight's name is followed by an exclamation mark.
```

##### 5. Find all lines that contain the text "that's what" followed by any amount of any text and then "said". How many lines are there?

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The UNIX command used to solve this problem.
- The number of lines that contain the text "that's what" followed by any amount of text and then "said".
```

Regular expressions are really a useful semi language-agnostic tool. What this means is regardless of the programming language your are using, there will be some package that allows you to use regular expressions. In fact, we can use them in both R and Python! This can be particularly useful when dealing with strings. Load up the dataset you discovered in (1) using `read.csv`. Name the resulting data.frame `dat`.

##### 6. The `text_w_direction` column in `dat` contains the characters' lines with inserted direction that helps characters know what to do as they are reciting the lines. Direction is shown between square brackets "[" "]". Create a new column called `has_direction` that is set to `TRUE` if the `text_w_direction` column has direction, and `FALSE` otherwise. Use regular expressions and the `grepl` function in R to accomplish this.

**Hint:** *Make sure all opening brackets "[" have a corresponding closing bracket "]".*

**Hint:** *Think of the pattern as any line that has a [, followed by any amount of any text, followed by a ], followed by any amount of any text.*

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep), [grepl](#r-grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
```

##### 7. Modify your regular expression in (7) to find lines with 2 or more sets of direction.

For example, the following line has 2 directions: `dat$text_w_direction[2789]`. How many lines have more than 2 directions? How many have more than 5?

```{txt}
This is a line with [emphasize this] only 1 direction!
This is a line with [emphasize this] 2 sets of direction, do you see the difference [shrug].
```

In (6), your solution may have found a match in both lines. In this question we want it to find only lines with 2+ directions, so the first line would not be a match.

**Relevant topics:** *[length](#r-length), [grep](#r-grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
- How many lines have > 2 directions?
- How many lines have > 5 directions?
```

##### 8. Use the `str_extract_all` function from the `stringr` package to extract the direction(s) as well as the text between direction(s) from each line. Put the strings in a new column called `direction`.

```{txt}
This is a line with [emphasize this] only 1 direction!
This is a line with [emphasize this] 2 sets of direction, do you see the difference [shrug].
```

In this question, your solution may have extracted:

```{txt}
[emphasize this]
[emphasize this] 2 sets of direction, do you see the difference [shrug]
```

This is ok.

**Note:** *If you capture text between two sets of direction, this is ok. For example, if we capture "[this] is a [test]" from "if we capture [this] is a [test]", this is ok.*

**Relevant topics:** *[str_extract_all](#r-str-extract)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
```

---

### Project 5 {#290-p05}

---

**Motivation:** Becoming comfortable stringing together commands and getting used to navigating files in a terminal is important for every data scientist to do. By learning the basics of a few useful tools, you will have the ability to quickly understand and manipulate files in a way which is just not possible using tools like Microsoft Office, Google Sheets, etc.

**Context:** We've been using UNIX tools in a terminal to solve a variety of problems. In this project we will continue to solve problems by combining a variety of tools using a form of redirection called piping.

**Scope:** grep, regular expression basics, UNIX utilities, redirection, piping

**Learning objectives:**

```{block, type="bbox"}
- Use `cut` to section off and slice up data from the command line.
- Use piping to string UNIX commands together.
- Use `sort` and it's options to sort data in different ways.
- Use `head` to isolate _n_ lines of output.
- Use `wc` to summarize the number of lines in a file or in output.
- Use `uniq` to filter out non-unique lines.
- Use `grep` to search files effectively.
```


You can find useful examples that walk you through relevant material in The Examples Book:

https://thedatamine.github.io/the-examples-book

It is highly recommended to read through, search, and explore these examples to help solve problems in this project.

**Important note:** It is highly recommended that you use https://rstudio.scholar.rcac.purdue.edu/. Simply click on the link and login using your Purdue account credentials. Use another system at your own risk. The version of RStudio on https://desktop.scholar.rcac.purdue.edu/ (which uses ThinLinc), is 99.9.9, and is known to have some strange issues when running code chunks.

Don't forget the very useful documentation shortcut `?`. To use, simply type `?` in the console, followed by the name of the function you are interested in. 

You can also look for package documentation by using `help(package=PACKAGENAME)`, so for example, to see the documentation for the package `ggplot2`, we could run:

```{r, eval=F}
help(package=ggplot2)
```

Sometimes it can be helpful to see the source code of a defined function. A [function](https://www.tutorialspoint.com/r/r_functions.htm) is any chunk of organized code that is used to perform an operation. Source code is the underlying `R` or `c` or `c++` code that is used to create the function. To see the source code of a defined function, type the function's name without the `()`. For example, if we were curious about what the function `Reduce` does, we could run:

```{r, eval=F}
Reduce
```

Occasionally this will be less useful as the resulting code will be code that calls `c` code we can't see. Other times it will allow you to understand the function better.

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/amazon/amazon_fine_food_reviews.csv`

A public sample of the data can be found here: [amazon_fine_food_reviews.csv](https://www.datadepot.rcac.purdue.edu/datamine/amazon/amazon_fine_food_reviews.csv)

Answers to questions should all be answered using the full dataset located on Scholar. You may use the public samples of data to experiment with your solutions prior to running them using the full dataset.

#### Questions

##### 1. What is the `Id` of the most helpful review if we consider the review with highest `HelpfulnessNumerator` to be an indicator of helpfulness (higher is more helpful)?

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [head](https://thedatamine.github.io/the-examples-book/unix.html#head), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- Line of UNIX commands used to solve the problem.
- The `Id` of the most helpful review.
```

##### 2. What proportion of all `Summary`s are unique? Use two lines of UNIX commands to find the answer.

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [wc](https://thedatamine.github.io/the-examples-book/unix.html#wc), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- Two lines of UNIX commands used to solve the problem.
- The ratio of unique `Summary`'s.
```

##### 3. Use a simple UNIX command to create a frequency table of `Score`.

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
- The frequency table.
```

##### 4. Who is the user with the highest number of reviews? There are two columns you could use to answer this question, but which column do you think would be most appropriate and why?

**Hint:** *You may need to pipe the output to `sort` multiple times.*

**Hint:** *To create the frequency table, read through the `man` pages for `uniq`. Man pages are the "manual" pages for UNIX commands. You can read through the man pages for uniq by running the following:*

```{bash, eval=F}
man uniq
```

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [head](https://thedatamine.github.io/the-examples-book/unix.html#head), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection), [man](https://thedatamine.github.io/the-examples-book/unix.html#man)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
- The frequency table.
```

##### 5. Anecdotally, there seems to be a tendency to leave reviews when we feel strongly (either positive or negative) about a product. For the user with the highest number of reviews, would you say that they follow this pattern of extremes? Let's consider 5 star reviews to be strongly positive and 1 star reviews to be strongly negative. Let's consider anything in between neither strongly positive nor negative.

**Hint:** *You may find the solution to problem (3) useful.*

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [grep](https://thedatamine.github.io/the-examples-book/unix.html#grep), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
```

##### 6. We want to compare the most helpful review with a `Score` of 5 with the most helpful review with a `Score` of 1. Use UNIX commands to calculate these values. Write down the `ProductId` of both reviews. In the case of a tie, write down all `ProductId`'s to get full credit. In this case we are considering the most helpful review to be the review with the highest `HelpfulnessNumerator`.

**Hint:** *You can use multiple lines to solve this problem.*

**Relevant topics:** *[sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [head](https://thedatamine.github.io/the-examples-book/unix.html#head), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The lines of UNIX commands used to solve the problem.
- `ProductId`'s of both requested reviews.
```

##### 7. Using the `ProductId`'s from the previous question, create a new dataset called `reviews.csv` which contains the `ProductId`'s and `Score` of all reviews with the corresponding `ProductId`'s. 

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#head), [redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
```

##### 8. Use R to load up `reviews.csv` into a new data.frame called `dat`. Create a histogram for each products' `Score`. Compare the most helpful review `Score` with the `Score`'s given in the histogram. Based on this comparison, decide (anecdotally) whether you think people found the review helpful because the product is overrated, underrated, or correctly reviewed by the masses.

**Relevant topics:** *[read.csv](file:///Users/kamstut/Dropbox/work/datamine/the-examples-book/docs/r.html#r-reading-and-writing-data), [hist](file:///Users/kamstut/Dropbox/work/datamine/the-examples-book/docs/r.html#r-plotting)*

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to create the histograms.
- 3 histograms, 1 for each `ProductId`.
- 1-2 sentences explaining whether or not you think people found the review helpful because the produce is overrated, underrated, or correctly reviewed, and why.
```

---

### Project 6 {#290-p06}

---

**Motivation:** A bash script is a powerful tool to perform repeated tasks. RCAC uses bash scripts to automate a variety of tasks. In fact, we use bash scripts on Scholar to do things like link Python kernels to your account, fix potential isues with Firefox, etc. `awk` is a programming language designed for text processing. The combination of these tools can be really powerful and useful for a variety of quick tasks.

**Context:** This is the first part in a series of projects that are designed to exercise skills around UNIX utilities, with a focus on writing bash scripts and `awk`. You will get the opportunity to manipulate data without leaving the terminal. At first it may seem overwhelming, however, with just a little practice you will be able to accomplish data wrangling tasks really efficiently. 

**Scope:** awk, UNIX utilities, bash scripts

**Learning objectives:**

```{block, type="bbox"}
- Use `awk` to process and manipulate textual data.
- Use piping and redirection within the terminal to pass around data between utilities.
```

#### Dataset: 

The following questions will use the dataset found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/YYYY.csv) or in Scholar:

`/class/datamine/data/flights/subset/YYYY.csv` 

An example from 1987 data can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/1987.csv) or in Scholar:

`/class/datamine/data/flights/subset/1987.csv`

#### Questions 

##### 1. In previous projects we learned how to get a single column of data from a csv file. Write 1 line of UNIX commands to print the 17th column, the `Origin`, from `1987.csv`. Write another line, this time using `awk` to do the same thing. Which one do you prefer, and why?

**Relevant topics:** [cut](#cut), [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- One line of UNIX commands to solve the problem *without* using `awk`.
- One line of UNIX commands to solve the problem using `awk`.
- 1-2 sentences describing which method you prefer and why.
```

##### 2. Write a bash script that accepts a year (1987, 1988, etc.) and a column *n* and returns the *nth* column of the associated year of data.

**Relevant topics:** [awk](#awk), [bash scripts](#writing-scripts)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
```

##### 3. How many flights came into Indianapolis (IND) in 2008? First solve this problem without using `awk`, then solve this problem using *only* `awk`.

**Relevant topics:** [cut](#cut), [grep](#grep), [wc](#wc), [awk](#awk), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:** 

- One line of UNIX commands to solve the problem *without* using `awk`.
- One line of  UNIX commands to solve the problem using `awk`. 
- The number of flights that came into Indianapolis (IND) in 2008.
```

##### 4. Do you expect the number of unique origins and destinations to be the same? Find out using any command line tool you'd like. Are they indeed the same? How many unique values do we have per category (`Origin`, `Dest`)?

**Relevant topics:** [cut](#cut), [sort](#sort), [uniq](#uniq), [wc](#wc), [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- 1-2 sentences explaining whether or not you expect the number of unique origins and destinations to be the same.
- The UNIX command(s) used to figure out if the number of unique origins and destinations are the same. 
- The number of unique values per category (`Origin`, `Dest`).
```

##### 5. In (4) we found that there are not the same number of unique `Origin`'s as `Dest`'s. Find the IATA airport code for all `Origin`'s that dont appear in a `Dest` and all `Dest`'s that don't appear in an `Origin`.

**Hint:** https://www.tutorialspoint.com/unix_commands/comm.html

**Relevant topics:** [comm](#unix), [cut](#cut), [sort](#sort), [uniq](#uniq), [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The line(s) of UNIX command(s) used to answer the question.
- The list of `Origin`s that don't appear in `Dest`.
- The list of `Dest`s that don't appear in `Origin`.
```


##### 6. What was the average number of flights in 2008 per unique `Origin` with the `Dest` of "IND"? How does "PHX" (as a unique `Origin`) compare to the average?

**Hint:** You manually do the average calculation by dividing the result from (3) by the number of unique `Origin`'s that have a `Dest` of "IND".

**Relevant topics:** [awk](#awk), [sort](#sort), [grep](#grep), [wc](#wc)

```{block, type="bbox"}
**Item(s) to submit:**

- The average number of flights in 2008 per unique `Origin` with the `Dest` of "IND".
- 1-2 sentences explaining how "PHX" compares (as a unique `Origin`) to the average?
```

##### 7. Write a bash script that takes a year and IATA airport code and returns the year, and the total number of flights to and from the given airport. Example rows may look like:

```{txt, eval=F}
1987, 12345
1988, 44
```

Run the script with inputs: `1991` and `ORD`. Include the output in your submission.

**Relevant topics:** [bash scripts](#writing-scripts), [cut](#cut), [piping](#piping-and-redirection), [grep](#grep), [wc](#wc)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
- The output of the script given `1991` and `ORD` as inputs.
```

---

### Project 7 {#290-p07}

---

**Motivation:** A bash script is a powerful tool to perform repeated tasks. RCAC uses bash scripts to automate a variety of tasks. In fact, we use bash scripts on Scholar to do things like link Python kernels to your account, fix potential isues with Firefox, etc. `awk` is a programming language designed for text processing. The combination of these tools can be really powerful and useful for a variety of quick tasks.

**Context:** This is the first part in a series of projects that are designed to exercise skills around UNIX utilities, with a focus on writing bash scripts and `awk`. You will get the opportunity to manipulate data without leaving the terminal. At first it may seem overwhelming, however, with just a little practice you will be able to accomplish data wrangling tasks really efficiently. 

**Scope:** awk, UNIX utilities, bash scripts

**Learning objectives:**

```{block, type="bbox"}
- Use `awk` to process and manipulate textual data.
- Use piping and redirection within the terminal to pass around data between utilities.
```

#### Dataset: 

The following questions will use the dataset found in Scholar:
`/class/datamine/data/flights/subset/YYYY.csv` 

An example of the data for the year 1987 can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/1987.csv).

Sometimes if you are about to dig into a dataset, it is good to quickly do some sanity checks early on to make sure the data is what you expect it to be. 

##### 1. Write a line of code that prints a list of the unique values in the `DayOfWeek` column. Write a line of code that prints a list of the unique values in the `DayOfMonth` column. Write a line of code that prints a list of the unique values in the `Month` column. Use the `1987.csv` dataset. Are the results what you expected?

**Relevant topics:** [cut](#cut), [sort](#sort)

```{block, type="bbox"}
**Item(s) to submit:**

- 3 lines of code used to get a list of unique values for the chosen columns.
- 1-2 sentences explaining whether or not the results are what you expected.
```

##### 2. Our files should have 29 columns. Write a line of code that prints any lines in a file that do *not* have 29 columns. Test it on `1987.csv`, were there any rows without 29 columns?

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of code used to solve the problem.
- 1-2 sentences explaining whether or not there were any rows without 29 columns.
```

##### 3. Write a bash script that, given a "begin" year and "end" year, cycles through the associated files and prints any lines that do *not* have 29 columns.

**Relevant topics:** [awk](#awk), [bash scripts](#writing-scripts)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
- The results of running your bash scripts from year 1987 to 2008.
```

##### 4. `awk` is a really good tool to quickly get some data and manipulate it a little bit. For example, let's see the number of kilometers and miles traveled in 1990. To convert from miles to kilometers, simply multiply by 1.609344. 

**Example output:**

```{txt, eval=F}
Miles: 12345
Kilometers: 19867.35168
```

**Relevant topics:** [awk](#awk), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem. 
- The results of running the code.
```

##### 5. Use `awk` to calculate the number of `DepDelay` minutes by `DayOfWeek`. Use `2007.csv`.

**Example output:**

```{txt, eval=F}
DayOfWeek:  0
1:  1234567
2:  1234567
3:  1234567
4:  1234567
5:  1234567
6:  1234567
7:  1234567
```

**Note:** 1 is Monday.

**Relevant topics:** [awk](#awk), [sort](#sort), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
```

##### 6. It wouldn't be fair to compare the total `DepDelay` minutes by `DayOfWeek` as the number of flights may vary. One way to take this into account is to instead calculate an average. Modify (5) to calculate the average number of `DepDelay` minutes by the number of flights per `DayOfWeek`. Use `2007.csv`.

**Example output:**

```{txt, eval=F}
DayOfWeek:  0
1:  1.234567
2:  1.234567
3:  1.234567
4:  1.234567
5:  1.234567
6:  1.234567
7:  1.234567
```

**Relevant topics:** [awk](#awk), [sort](#sort), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
```

##### 7. As a quick follow-up, _slightly_ modify (6) to perform the same calculation for `ArrDelay`. Do the `ArrDelay`s and `DepDelay`s appear to have the highest delays on the same day? Use `2007.csv`.

**Example output:**

```{txt, eval=F}
DayOfWeek:  0
1:  1.234567
2:  1.234567
3:  1.234567
4:  1.234567
5:  1.234567
6:  1.234567
7:  1.234567
```

**Relevant topics:** [awk](#awk), [sort](#sort), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
- 1-2 sentences explaining whether or not the `ArrDelay`s and `DepDelay`s appear to have the highest delays on the same day.
```

---

### Project 8 {#290-p08}

---

**Motivation:** A bash script is a powerful tool to perform repeated tasks. RCAC uses bash scripts to automate a variety of tasks. In fact, we use bash scripts on Scholar to do things like link Python kernels to your account, fix potential isues with Firefox, etc. `awk` is a programming language designed for text processing. The combination of these tools can be really powerful and useful for a variety of quick tasks.

**Context:** This is the first part in a series of projects that are designed to exercise skills around UNIX utilities, with a focus on writing bash scripts and `awk`. You will get the opportunity to manipulate data without leaving the terminal. At first it may seem overwhelming, however, with just a little practice you will be able to accomplish data wrangling tasks really efficiently. 

**Scope:** awk, UNIX utilities, bash scripts

**Learning objectives:**

```{block, type="bbox"}
- Use `awk` to process and manipulate textual data.
- Use piping and redirection within the terminal to pass around data between utilities.
```

#### Dataset: 

The following questions will use the dataset found in Scholar:
`/class/datamine/data/flights/subset/YYYY.csv` 

An example of the data for the year 1987 can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/1987.csv).


Let's say we have a theory that there are more flights on the weekend days (Friday, Saturday, Sunday) than the rest of the days, on average. We can use awk to quickly check it out and see if maybe this looks like something that is true!

##### 1. Write a line of `awk` code that, prints the number of flights on the weekend days, followed by the number of flights on the weekdays for the flights during 2008.

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of `awk` code that solves the problem.
- The result: the number of flights on the weekend days, followed by the number of flights on the weekdays for the flights during 2008.
```

##### 2. Note that in (1), we are comparing 3 days to 4! Write a line of `awk` code that, prints the average number of flights on a weekend day, followed by the average number of flights on the weekdays. Continue to use data for 2008.

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of `awk` code that solves the problem.
- The result: the average number of flights on the weekend days, followed by the average number of flights on the weekdays for the flights during 2008.
```

We want to look to see if there may be some truth to the whole "snow bird" concept where people will travel to warmer states like Florida and Arizona during the Winter. Let's use the tools we've learned to explore this a little bit. 

##### 3. Take a look at `airports.csv`. In particular run the following:

```{bash, eval=F}
head airports.csv
```

Notice how all of the non-numeric text is surrounded by quotes. The surrounding quotes would need to be escaped for any comparison within `awk`. This is messy and we would prefer to create a new file called `new_airports.csv` without any quotes. Write a line of code to do this. 

**Hint:** You could use `gsub` within `awk` to replace '"' with ''.

**Hint:** If you leave out the column number argument to `gsub` it will apply the substitution to every field in every column.

**Relevant topics:** [awk](#awk), [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of `awk` code used to create the new dataset.
```

##### 4. Write a line of commands that create a new dataset called `az_fl_airports.txt` that contains a list of airport codes for all airports from both Arizona (AZ) and Florida (FL). Use the file we created in (3),`new_airports.csv`.

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands to create an array called `airports`.
```

##### 5. Wow! In (4) we discovered a lot of airports! How many airports are there? Did you expect this? Use a line of bash code to answer this question.

**Relevant topics:** [echo](#echo), [wc](#wc), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of UNIX commands used to solve the problem.
- The number of airports.
- 1-2 sentences explaining whether you expected this result and why or why not.
```

##### 6. Create a new dataset that contains all of the data for flights into or out of Florida and Arizona using 2008.csv, use the newly created dataset, `az_fl_airports.txt` in (4) to do so.

**Hint:** https://unix.stackexchange.com/questions/293684/basic-grep-awk-help-extracting-all-lines-containing-a-list-of-terms-from-one-f

**Relevant topics:** [grep](#grep)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of UNIX commands used to solve the problem.
```

##### 7. Now that you have code to complete (6), write a bash script that accepts the start year, end year, and filename containing airport codes (`az_fl_airports.txt`), and outputs the data for flights into or out of any of the airports listed in the provided filename containing airport codes using _all_ of the years of data in the provided range. Run the bash script to create a new file called `az_fl_flights.csv`.

**Relevant topics:** [bash scripts](#writing-scripts), [grep](#grep), for loop, [redirection](#redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
- The line of UNIX code you used to execute the script and create the new dataset.
```

---

### Project 9 {#290-p09}

### Project 10 {#290-p10}

### Project 11 {#290-p11}

### Project 12 {#290-p12}

--- 

**Motivation:** Databases are comprised of many tables. It is imperative that we learn how to combine data from multiple tables using queries. To do so we perform joins! In this project we will explore learn about and practice using joins on a database containing bike trip information from the Bay Area Bike Share. 

**Context:** We've introduced a variety of SQL commands that let you filter and extract information from a database in an systematic way. In this project we will introduce joins, a powerful method to combine data from different tables.

**Scope:** SQL, sqlite, joins

**Learning objectives:**

```{block, type="bbox"}
- Briefly explain the differences between left and inner join and demonstrate the ability to use the join statements to solve a data-driven problem.
- Perform grouping and aggregate data using group by and the following functions: count, max, sum, avg, like, having. 
- Showcase the ability to filter, alias, and write subqueries.
```

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/bay_area_bike_share/bay_area_bike_share.db`

A public sample of the data can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/data/bay_area_bike_share/bay_area_bike_share.db).

#### Questions 

##### 1. There are a variety of ways to join data using SQL. With that being said, if you are able to understand and use a LEFT JOIN and INNER JOIN, you can perform *all* of the other types of joins (RIGHT JOIN, FULL OUTER JOIN). Given the following two tables, use [RMarkdown](#r-rmarkdown) to display the result of performing the following query as a table:

```{sql, eval=F}
SELECT * FROM users AS u INNER JOIN dorms AS d ON u.dorm=d.id;
```

**users:**

id | first_name | last_name | dorm
---|------------|-----------|-----
1  | Alice      | Smith     | 1
2  | Bob        | Johnson   | 2
3  | Susan      | Marques   | 3
4  | Amare      | Keita     | 3
5  | Kristen    | Lakehold  | 4

**dorms:**

id | name | capacity | address
---|------|----------|--------
1  | Windsor Halls | NULL | Windsor Halls, West Lafayette, IN, 47906
2  | Cary Quadrangle | 1200 | 1016 W Stadium Ave, West Lafayette, IN 47906
3  | Hillenbrand Hall | NULL | 1301 3rd Street, West Lafayette, IN 47906

**Relevant topics:** [sql](#sql), [inner join](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- [RMarkdown](#r-rmarkdown) table displaying the result of performing the following query as a table.
```

##### 2. Using the same two tables from (1), use [RMarkdown](#r-rmarkdown) to display the result of performing the following query as a table. Explain the difference between an INNER JOIN and LEFT JOIN.

```{sql, eval=F}
SELECT * FROM users AS u LEFT JOIN dorms AS d ON u.dorm=d.id;
```

**Relevant topics:** [sql](#sql), [left join](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- [RMarkdown](#r-rmarkdown) table displaying the result of performing the following query as a table.
- 1-2 sentences explaining (in your own words) what the difference between and INNER and LEFT JOIN is.
```

##### 3. Aliases can be created for tables, fields, and even results of aggregate functions (like MIN, MAX, COUNT, AVG, etc.). In addition, you can combine fields using the `sqlite` concatenate operator `||` (see [here](https://www.sqlitetutorial.net/sqlite-string-functions/sqlite-concat/)). Write a query that returns the first 5 records of information from the `station` table formatted in the following way:

`(id) name @ (lat, long)`

For example:

`(84) Ryland Park @ (37.342725,-121.895617)`

**Relevant topics:** [aliasing](#sql-examples), concat

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve this problem. 
- The first 5 records of information from the `station` table.
```

##### 4. There is a variety of interesting weather information in the `weather` table. Write a query that finds the average `mean_temperature_f` by `zip_code`. Which is on average the warmest `zip_code`?

Use aliases to format the result in the following way:

```{txt}
Zip Code|Avg Temperature
94041|61.3808219178082
```

**Relevant topics:** [aliasing](#sql-examples), [group by](#sql-examples), avg

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve this problem.
- The results of the query copy and pasted.
```

##### 5. From (4) we can see that there are only 5 `zip_code`s with weather information. How many unique `zip_code`s do we have in the `trip` table? Write a query that finds the number of unique `zip_code`s in the `trip` table. Write another query that lists the `zip_code` and count of the number of times the `zip_code` appears. If we had originally assumed that the `zip_code` was related to the location of the trip itself, we were wrong. Can you think of a likely explanation?

Relevant topics: [group by](#sql-examples), [count](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL queries used to solve this problem.
- 1-2 sentences explainging what a possible explanation for the `zip_code`s could be.
```

##### 6. In (4) we wrote a query that finds the average `mean_temperature_f` by `zip_code`. What if we want to tack on to our results information from each row in the `station` table based on the `zip_code`s? To do, use an INNER JOIN. INNER JOIN combines tables based on specified fields, and returns only rows where there is a match in both the "left" and "right" tables.

**Hint:** Use the query from (4) as a sub query within your solution.

**Relevant topics:** [inner join](#sql-examples), subqueries, [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve this problem.
```

##### 7. In (5) we eluded that the `zip_code`s in the `trip` table aren't very consistent. Users can enter a zip code when using the app. This means that `zip_code` can be from anywhere in the world! With that being said, if the `zip_code` is one of the 5 `zip_code`s for which we have weather data (from question 4), we can add that weather information to matching rows of the `trip` table. In (6) we used an INNER JOIN to append some weather information to each row in the `station` table. For this question, write a query that performs an INNER JOIN and appends weather data from the `weather` table to the trip data from the `trip` table. Limit your solution to 5 lines. 

**Hint:** You will want to wrap your dates and datetimes in [sqlite's `date` function](https://www.sqlitetutorial.net/sqlite-date-functions/sqlite-date-function/) prior to comparison.

**Important note:** Notice that the weather data has about 1 row of weather information for each date and each zip code. This means you may have to join your data based on multiple constraints instead of just 1 like in (6).

**Relevant topics:** [inner join](#sql-examples), [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve this problem.
- First 5 lines of output.
```

--- 

### Project 13 {#290-p13}

### Project 14 {#290-p14}

### Project 15 {#290-p15}

---

**Motivation:** We've done a lot of work with SQL this semester. Let's review concepts in this project and mix and match R, Python, and SQL to solve data-driven problems.

**Context:** In this project, we will reinforce topics you've already learned, with a focus on SQL.

**Scope:** SQL, sqlite, R, Python

**Learning objectives:**

```{block, type="bbox"}
- Write and run SQL queries in `sqlite` on real-world data.
- Use SQL from within R and Python.
```

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/movies_and_tv/imdb.db`

A public sample of the data can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/data/movies-and-tv/imdb.db).

In this project we want to offer the flexibility of using your choice of R and/or Python. To keep things as consistent as possible, please use Rmarkdown on https://rstudio.scholar.rcac.purdue.edu/. See [here](#python-on-scholar) to learn how to run Python in this environment.

#### Questions 

##### 1. What is the first year where our database has > 1000 titles? Use the `premiered` column in the `titles` table as our year. What year has the most titles?

**Relevant topics:** [count](#sql-examples), [group by](#sql-examples), [order by](#sql-examples), [desc](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- 1 or more SQL queries used to answer the questions.
- What year is the first year to have > 1000 titles? 
- What year has the most titles?
```

##### 2. How many, and what are the unique `type`s from the `titles` table? From the year from (1) with the most `titles`, how many titles of each `type` are there?

```{block, type="bbox"}
**Item(s) to submit:**

- 1 or more SQL queries used to answer the questions.
- How many and what are the unique `types` from the `titles` table?
- A list of `type` and and count for the year (`premiered`) 2017.
```

F.R.I.E.N.D.S is a popular tv show. They have an interesting naming convention for the names of their episodes. They all begin with the text "The One ...". There are 6 primary characters in the show: Chandler, Joey, Monica, Phoebe, Rachel, and Ross. Let's use SQL and R to take a look at how many times each characters' names appear in the title of the episodes.

##### 3. Write a query that gets the `episode_title_id`, `primary_title`, `rating`, and `votes`, of all of the episodes of Friends (`title_id` is tt0108778). 

**Hint:** You can slightly modify the solution to question (7) in project 13.

**Relevant topics:** [inner join](#sql-examples), subqueries, [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to answer the question.
- First 5 results of the query.
```

The next couple of questions should be complete in the same language. You can use either R or Python, but you must use the same for both questions.

##### 4. Now that you have a working query, connect to the database and run the query to get the data into an R or `pandas` data frame. In previous projects, we learned how to used regular expressions to search for text. For each character, how many episodes `primary_title`s contained their name? 

**Relevant topics:** [SQL in R](#sql-in-r), [SQL in Python](#sql-in-python), [grep](#r-grep)

```{block, type="bbox"}
**Item(s) to submit:**

- R or Python code in a code chunk that was used to find the solution.
- The solution pasted below the code chunk.
```

##### 5. Create a graphic showing our results in (2) using your favorite package. Make sure the plot has a good title, x-label, y-label, and try to incorporate some of the following colors: #273c8b, #bd253a, #016f7c, #f56934, #016c5a, #9055b1, #eaab37.

**Relevant topics:** [plotting](#r-plotting)

```{block, type="bbox"}
**Item(s) to submit:**

- The R or Python code used to generate the graphic.
- The graphic in a png or jpg/jpeg format.
```

##### 6. Use any combination of SQL, R, and Python you'd like in order to find which of the following 3 genres has the highest average rating for movies (see `type` column from `titles` table): Romance, Comedy, Animation. In the `titles` table, you can find the genres in the `genres` column. There may be some overlap (i.e. a movie may have more than one genre), this is ok. 

To query rows which have the genre Action as one of its genres:

```{sql, eval=F}
SELECT * FROM titles WHERE genres LIKE '%action%';
```

**Relevant topics:** [like](#sql-examples), [inner join](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- Any code you used to solve the problem in a code chunk.
- The average rating of each of the genres listed for movies.
```

---

## STAT 39000

### Project 1 {#390-p01}

### Project 2 {#390-p02}

---

**Motivation:** The ability to quickly reproduce an analysis is important. It is often necessary that other individuals will need to be able to understand and reproduce an analysis. This concept is so important there are classes solely on reproducible research! In fact, there are papers that investigate and highlight the lack of reproducibility in various fields. If you are interested in reading about this topic, a good place to start is the paper titled ["Why Most Published Research Findings Are False"](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.0020124), by John Ioannidis (2005). 

**Context:** Making your work reproducible is extremely important. We will focus on the computational part of reproducibility. We will learn RMarkdown to document your analyses so others can easily understand and reproduce the computations that led to your conclusions. Pay close attention as future project templates will be RMarkdown templates.

**Scope:** Understand Markdown, RMarkdown, and how to use it to make your data analysis reproducible.

**Learning objectives:**

```{block, type="bbox"}
- Use Markdown syntax within an Rmarkdown document to achieve various text transformations.
- Use RMarkdown code chunks to display and/or run snippets of code.
```


You can find useful examples that walk you through relevant material in The Examples Book:

https://thedatamine.github.io/the-examples-book

It is highly recommended to read through, search, and explore these examples to help solve problems in this project.

**Important note:** It is highly recommended that you use https://rstudio.scholar.rcac.purdue.edu/. Simply click on the link and login using your Purdue account credentials. Use another system at your own risk. The version of RStudio on https://desktop.scholar.rcac.purdue.edu/ (which uses ThinLinc), is 99.9.9, and is known to have some strange issues when running code chunks.

Don't forget the very useful documentation shortcut `?`. To use, simply type `?` in the console, followed by the name of the function you are interested in. 

You can also look for package documentation by using `help(package=PACKAGENAME)`, so for example, to see the documentation for the package `ggplot2`, we could run:

```{r, eval=F}
help(package=ggplot2)
```

Sometimes it can be helpful to see the source code of a defined function. A [function](https://www.tutorialspoint.com/r/r_functions.htm) is any chunk of organized code that is used to perform an operation. Source code is the underlying `R` or `c` or `c++` code that is used to create the function. To see the source code of a defined function, type the function's name without the `()`. For example, if we were curious about what the function `Reduce` does, we could run:

```{r, eval=F}
Reduce
```

Occasionally this will be less useful as the resulting code will be code that calls `c` code we can't see. Other times it will allow you to understand the function better.

##### 1. Make the following text (including the asterisks) bold: `This needs to be **very** bold`. Make the following text (including the underscores) italicized: `This needs to be _very_ italicized.`

**Hint:** *Try mixing and matching ways to embolden or italicize text. Alternatively, look up "escaping characters in markdown", or see [here](https://thedatamine.github.io/the-examples-book/faqs.html#escape-characters).*

**Hint:** *Be sure to check out the [Rmarkdown Cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) and our section on [Rmarkdown in the book](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown).*

**Note:** *Rmarkdown is essentially Markdown + the ability to run and display code chunks. In this question, we are actually using Markdown within Rmarkdown!*

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown), [escaping characters](https://thedatamine.github.io/the-examples-book/faqs.html#escape-characters)*

```{block, type="bbox"}
**Item(s) to submit:**

- Fill in the chunk under (1) in the `stat29000project02template.Rmd` file with 2 lines of markdown text. Note that when compiled, this text will be unmodified, regular text.
```

##### 2. Create an unordered list of your top 3 favorite academic interests (some examples could include: machine learning, operating systems, forensic accounting, etc.). Create another *ordered* list that ranks your academic interests in order of most interested to least interested.

**Hint:** *You can learn what ordered and unordered lists are [here]( https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).*

**Note:** *Similar to (1a), in this question we are dealing with Markdown. If we were to copy and paste the solution to this problem in a Markdown editor, it would be the same result as when we Knit it here.*

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- Create the lists under (2) in the `stat29000project02template.Rmd` file. Note that when compiled, this text will appear as nice, formatted lists.
```

##### 3. Browse https://www.linkedin.com/ and read some profiles. Pay special attention to accounts with an "About" section. Write your own personal "About" section using Markdown. Include the following:

- A header (your choice of size) that says "About".
- A horizontal rule directly underlining the header.
- The text of your personal "About" section that you would feel comfortable uploading to linkedin, including at least 1 link.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- Create the described profile under (3) in the `stat29000project02template.Rmd` file.
```

##### 4. LaTeX is a powerful editing tool where you can create beautifully formatted equations and formulas. Replicate the equation found [here](https://wikimedia.org/api/rest_v1/media/math/render/svg/87c061fe1c7430a5201eef3fa50f9d00eac78810) as closely as possible.

**Hint:** *Lookup "latex mid" and "latex frac".*

```{block, type="bbox"}
**Item(s) to submit:**

- Replicate the equation using LaTeX under (1d) in the `stat39000project02template.Rmd` file.
```

##### 5. Your co-worker wrote a report, and has asked you to beautify it. Knowing Rmarkdown, you agreed. Spruce up the report found under (4) in `stat29000project02template.Rmd`. At a minimum:

- Make the title pronounced.
- Make all links appear as a word or words, rather than the long-form URL.
- Organize all code into code chunks where code and output are displayed. If the output is really long, just display the code.
- Make the calls to the `library` function and the `install.packages` function be evaluated but not displayed. Make sure all warnings and errors that may eventually occur, do not appear in the final document.

Feel free to make any other changes that make the report more visually pleasing.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- Spruce up the "document" under (4) in the `stat29000project02template.Rmd` file.
```

#### 6. Create a plot using a built-in dataset like `iris`, `mtcars`, or `Titanic`, and display the plot using a code chunk. Make sure the code used to generate the plot is hidden. Include a descriptive caption for the image.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown), [plotting in r](https://thedatamine.github.io/the-examples-book/r.html#r-plotting)*

```{block, type="bbox"}
**Item(s) to submit:**

- Code chunk under (5) that creates and displays a plot using a built-in dataset like `iris`, `mtcars`, or `Titanic`.
```

##### 7. Insert the following code chunk under (5) in `stat29000project02template.Rmd`. Try knitting the document. Something should go wrong. Fix the problem and knit again. If another problem appears, fix it. What was the first problem? What was the second problem?
       
````markdown
`r ''````{r install-packages}
plot(my_variable)
```
````

**Hint:** *Take a close look at the name we give our code chunk.*

**Hint:** *Take a look at the code chunk where `my_variable` is declared.*

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- The modified version of the inserted code that fixes both problems.
- A sentence explaining what the first problem was.
- A sentence explaining what the second problem was.
```

##### 8. RMarkdown is also an excellent tool to create a slide deck. Use the information [here](https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf) or [here](https://thedatamine.github.io/the-examples-book/r.html#how-do-i-create-a-set-of-slides-using-rmarkdown) to convert your solutions into a slide deck rather than the regular PDF. You may use `slidy`, `ioslides` or `beamer`. Make any needed modifications to make the solutions knit into a well-organized slide deck. Modify (2) so the bullets are incrementally presented as the slides progresses.

**Relevant topics:** *[rmarkdown](https://thedatamine.github.io/the-examples-book/r.html#r-rmarkdown)*

```{block, type="bbox"}
**Item(s) to submit:**

- The modified version of the inserted code that fixes both problems.
- A sentence explaining what the first problem was.
- A sentence explaining what the second problem was.
```

---

### Project 3 {#390-p03}

### Project 4 {#390-p04}

---

**Motivation:** The need to search files and datasets based on the text held within is common during various parts of the data wrangling process. `grep` is an extremely powerful UNIX tool that allows you to do so using regular expressions. Regular expressions are a structured method for searching for specified patterns. Regular expressions can be very complicated, [even professionals can make critical mistakes](https://blog.cloudflare.com/details-of-the-cloudflare-outage-on-july-2-2019/). With that being said, learning some of the basics is an incredible tool that will come in handy regardless of the language you are working in.

**Context:** We've just begun to learn the basics of navigating a file system in UNIX using various terminal commands. Now we will go into more depth with one of the most useful command line tools, `grep`, and experiment with regular expressions using `grep`, R, and later on, Python.

**Scope:** grep, regular expression basics, utilizing regular expression tools in R and Python

**Learning objectives:**

```{block, type="bbox"}
- Use `grep` to search for patterns within a dataset.
- Use `cut` to section off and slice up data from the command line.
- Use `wc` to count the number of lines of input.
```


You can find useful examples that walk you through relevant material in The Examples Book:

https://thedatamine.github.io/the-examples-book

It is highly recommended to read through, search, and explore these examples to help solve problems in this project.

**Important note:** It is highly recommended that you use https://rstudio.scholar.rcac.purdue.edu/. Simply click on the link and login using your Purdue account credentials. Use another system at your own risk. The version of RStudio on https://desktop.scholar.rcac.purdue.edu/ (which uses ThinLinc), is 99.9.9, and is known to have some strange issues when running code chunks.

Don't forget the very useful documentation shortcut `?`. To use, simply type `?` in the console, followed by the name of the function you are interested in. 

You can also look for package documentation by using `help(package=PACKAGENAME)`, so for example, to see the documentation for the package `ggplot2`, we could run:

```{r, eval=F}
help(package=ggplot2)
```

Sometimes it can be helpful to see the source code of a defined function. A [function](https://www.tutorialspoint.com/r/r_functions.htm) is any chunk of organized code that is used to perform an operation. Source code is the underlying `R` or `c` or `c++` code that is used to create the function. To see the source code of a defined function, type the function's name without the `()`. For example, if we were curious about what the function `Reduce` does, we could run:

```{r, eval=F}
Reduce
```

Occasionally this will be less useful as the resulting code will be code that calls `c` code we can't see. Other times it will allow you to understand the function better.

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/movies_and_tv/the_office_dialogue.csv`

A public sample of the data can be found here: [the_office_dialogue.csv](https://www.datadepot.rcac.purdue.edu/datamine/movies-and-tv/the_office_dialogue.csv)

Answers to questions should all be answered using the full dataset located on Scholar. You may use the public samples of data to experiment with your solutions prior to running them using the full dataset.

`grep` stands for (g)lobally search for a (r)egular (e)xpression and (p)rint matching lines. As such, to best demonstrate `grep`, we will be using it with textual data. You can read about and see examples of `grep` [here](https://thedatamine.github.io/the-examples-book/unix.html#grep).

##### 1. Login to Scholar and use `grep` to find the dataset we will use this project. The dataset we will use is the only dataset to have the text "bears. beets. battlestar galactica.". What is the name of the dataset and where is it located?

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The `grep` command used to find the dataset.
- The name and location in Scholar of the dataset.
- Use `grep` and `grepl` within R to solve a data-driven problem.
```

##### 2. `grep` prints the line that the text you are searching for appears in. In project 3 we learned a UNIX command to quickly print the first _n_ lines from a file. Use this command to get the headers for the dataset. As you can see, each line in the tv show is a row in the dataset. You can count to see which column the various bits of data live in.

Write a line of UNIX commands that searches for "bears. beets. battlestar galactica." and, rather than printing the entire line, prints only the character who speaks the line, as well as the line itself.

**Hint:** *The result if you were to search for "bears. beets. battlestar galactica." should be:*

```{txt}
"Jim","Fact. Bears eat beets. Bears. Beets. Battlestar Galactica."
```

**Hint:** *One method to solve this problem would be to [pipe](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection) the output from `grep` to [`cut`](https://thedatamine.github.io/the-examples-book/unix.html#cut).*

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to perform the operation.
```

##### 3. This particular dataset happens to be very small. You could imagine a scenario where the file is many gigabytes and not easy to load completely into R or Python. We are interested in learning what makes Jim and Pam tick as a couple. Use a line of UNIX commands to create a new dataset called `jim_and_pam.csv`. Include only lines that are spoken by either Jim or Pam, or reference Jim or Pam in any way. Include only the following columns: `episode_name`, `character`, `text`, `text_w_direction`, and `air_date`. How many rows of data are in the new file? How many megabytes is the new file (to the nearest 1/10th of a megabyte)?

**Hint:** *[Redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection).*

**Hint:** *It is OK if you get an erroneous line where the word "jim" or "pam" appears as a part of another word.*

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [grep](https://thedatamine.github.io/the-examples-book/unix.html#grep), [ls](https://thedatamine.github.io/the-examples-book/unix.html#ls), [wc](https://thedatamine.github.io/the-examples-book/unix.html#wc), [redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to create the new file. 
- The number of rows of data in the new file, and the accompanying UNIX command used to find this out.
- The number of megabytes (to the nearest 1/10th of a megabyte) that the new file has, and the accompanying UNIX command used to find this out.
```

##### 4. Find all lines where either Jim/Pam/Michael/Dwight's name is followed by an exclamation mark. Use only 1 "!" within your regular expression. How many lines are there?

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The UNIX command(s) used to solve this problem.
- The number of lines where either Jim/Pam/Michael/Dwight's name is followed by an exclamation mark.
```

##### 5. Find all lines that contain the text "that's what" followed by any amount of any text and then "said". How many lines are there?

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The UNIX command used to solve this problem.
- The number of lines that contain the text "that's what" followed by any amount of text and then "said".
```

##### 6. Find all of the lines where Pam is called "Beesley" instead of "Pam" or "Pam Beesley".

**Hint:** *A negative lookbehind would be one way to solve this.*

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The UNIX command used to solve this problem.
```

Regular expressions are really a useful semi language-agnostic tool. What this means is regardless of the programming language your are using, there will be some package that allows you to use regular expressions. In fact, we can use them in both R and Python! This can be particularly useful when dealing with strings. Load up the dataset you discovered in (1) using `read.csv`. Name the resulting data.frame `dat`.

##### 7. The `text_w_direction` column in `dat` contains the characters' lines with inserted direction that helps characters know what to do as they are reciting the lines. Direction is shown between square brackets "[" "]". Create a new column called `has_direction` that is set to `TRUE` if the `text_w_direction` column has direction, and `FALSE` otherwise. Use regular expressions and the `grepl` function in R to accomplish this.

**Hint:** *Make sure all opening brackets "[" have a corresponding closing bracket "]".*

**Hint:** *Think of the pattern as any line that has a [, followed by any amount of any text, followed by a ], followed by any amount of any text.*

**Relevant topics:** *[grep](https://thedatamine.github.io/the-examples-book/unix.html#grep), [grepl](#r-grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
```

##### 8. Modify your regular expression in (7) to find lines with 2 or more sets of direction.

For example, the following line has 2 directions: `dat$text_w_direction[2789]`. How many lines have more than 2 directions? How many have more than 5?

```{txt}
This is a line with [emphasize this] only 1 direction!
This is a line with [emphasize this] 2 sets of direction, do you see the difference [shrug].
```

In (7), your solution may have found a match in both lines. In this question we want it to find only lines with 2+ directions, so the first line would not be a match.

**Relevant topics:** *[length](#r-length), [grep](#r-grep)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
- How many lines have > 2 directions?
- How many lines have > 5 directions?
```

##### 9. Use the `str_extract_all` function from the `stringr` package to extract the direction(s) as well as the text between direction(s) from each line. Put the strings in a new column called `direction`.

```{txt}
This is a line with [emphasize this] only 1 direction!
This is a line with [emphasize this] 2 sets of direction, do you see the difference [shrug].
```

In this question, your solution may have extracted:

```{txt}
[emphasize this]
[emphasize this] 2 sets of direction, do you see the difference [shrug]
```

This is ok.

**Note:** *If you capture text between two sets of direction, this is ok. For example, if we capture "[this] is a [test]" from "if we capture [this] is a [test]", this is ok.*

**Relevant topics:** *[str_extract_all](#r-str-extract)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
```

##### 10. Repeat (9) but this time make sure you only capture the brackets and text within the brackets. Save the results in a new column called `direction_correct`. You can test to see if it is working by running the following code:

```{r, eval=F}
dat$direction_correct[747]
```

```{txt}
This is a line with [emphasize this] only 1 direction!
This is a line with [emphasize this] 2 sets of direction, do you see the difference [shrug].
```

In (7), your solution may have extracted:

```{txt}
[emphasize this]
[emphasize this] 2 sets of direction, do you see the difference [shrug]
```

This is ok for (7). In this question, however, we want to fix this to only extract:

```{txt}
[emphasize this]
[emphasize this] [shrug]
```

**Hint:** *This regular expression will be hard to read.*

**Hint:** *The pattern we want is: literal opening bracket, followed by 0+ of any character other than the literal [ or literal ], followed by a literal closing bracket.*

**Relevant topics:** *[str_extract_all](#r-str-extract)*

```{block, type="bbox"}
**Item(s) to submit:**

- The R code used to solve this problem.
```

---

### Project 5 {#390-p05}

---

**Motivation:** Becoming comfortable stringing together commands and getting used to navigating files in a terminal is important for every data scientist to do. By learning the basics of a few useful tools, you will have the ability to quickly understand and manipulate files in a way which is just not possible using tools like Microsoft Office, Google Sheets, etc.

**Context:** We've been using UNIX tools in a terminal to solve a variety of problems. In this project we will continue to solve problems by combining a variety of tools using a form of redirection called piping.

**Scope:** grep, regular expression basics, UNIX utilities, redirection, piping

**Learning objectives:**

```{block, type="bbox"}
- Use `cut` to section off and slice up data from the command line.
- Use piping to string UNIX commands together.
- Use `sort` and it's options to sort data in different ways.
- Use `head` to isolate _n_ lines of output.
- Use `wc` to summarize the number of lines in a file or in output.
- Use `uniq` to filter out non-unique lines.
- Use `grep` to search files effectively.
```

You can find useful examples that walk you through relevant material in The Examples Book:

https://thedatamine.github.io/the-examples-book

It is highly recommended to read through, search, and explore these examples to help solve problems in this project.

**Important note:** It is highly recommended that you use https://rstudio.scholar.rcac.purdue.edu/. Simply click on the link and login using your Purdue account credentials. Use another system at your own risk. The version of RStudio on https://desktop.scholar.rcac.purdue.edu/ (which uses ThinLinc), is 99.9.9, and is known to have some strange issues when running code chunks.

Don't forget the very useful documentation shortcut `?`. To use, simply type `?` in the console, followed by the name of the function you are interested in. 

You can also look for package documentation by using `help(package=PACKAGENAME)`, so for example, to see the documentation for the package `ggplot2`, we could run:

```{r, eval=F}
help(package=ggplot2)
```

Sometimes it can be helpful to see the source code of a defined function. A [function](https://www.tutorialspoint.com/r/r_functions.htm) is any chunk of organized code that is used to perform an operation. Source code is the underlying `R` or `c` or `c++` code that is used to create the function. To see the source code of a defined function, type the function's name without the `()`. For example, if we were curious about what the function `Reduce` does, we could run:

```{r, eval=F}
Reduce
```

Occasionally this will be less useful as the resulting code will be code that calls `c` code we can't see. Other times it will allow you to understand the function better.

#### Dataset

The following questions will use the dataset found in Scholar:

`/class/datamine/data/amazon/amazon_fine_food_reviews.csv`

A public sample of the data can be found here: [amazon_fine_food_reviews.csv](https://www.datadepot.rcac.purdue.edu/datamine/amazon/amazon_fine_food_reviews.csv)

Answers to questions should all be answered using the full dataset located on Scholar. You may use the public samples of data to experiment with your solutions prior to running them using the full dataset.

#### Questions

##### 1. What is the `Id` of the most helpful review if we consider the review with highest `HelpfulnessNumerator` to be an indicator of helpfulness (higher is more helpful)?

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [head](https://thedatamine.github.io/the-examples-book/unix.html#head), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- Line of UNIX commands used to solve the problem.
- The `Id` of the most helpful review.
```

##### 2. What proportion of all `Summary`s are unique? Use two lines of UNIX commands to find the answer.

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [wc](https://thedatamine.github.io/the-examples-book/unix.html#wc), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- Two lines of UNIX commands used to solve the problem.
- The ratio of unique `Summary`'s.
```

##### 3. Use a simple UNIX command to create a frequency table of `Score`.

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
- The frequency table.
```

##### 4. Who is the user with the highest number of reviews? There are two columns you could use to answer this question, but which column do you think would be most appropriate and why?

**Hint:** *You may need to pipe the output to `sort` multiple times.*

**Hint:** *To create the frequency table, read through the `man` pages for `uniq`. Man pages are the "manual" pages for UNIX commands. You can read through the man pages for uniq by running the following:*

```{bash, eval=F}
man uniq
```

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [head](https://thedatamine.github.io/the-examples-book/unix.html#head), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection), [man](https://thedatamine.github.io/the-examples-book/unix.html#man)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
- The frequency table.
```

##### 5. Anecdotally, there seems to be a tendency to leave reviews when we feel strongly (either positive or negative) about a product. For the user with the highest number of reviews, would you say that they follow this pattern of extremes? Let's consider 5 star reviews to be strongly positive and 1 star reviews to be strongly negative. Let's consider anything in between neither strongly positive nor negative.

**Hint:** *You may find the solution to problem (3) useful.*

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [uniq](https://thedatamine.github.io/the-examples-book/unix.html#uniq), [sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [grep](https://thedatamine.github.io/the-examples-book/unix.html#grep), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
```

##### 6. We want to compare the most helpful review with a `Score` of 5 with the most helpful review with a `Score` of 1. Use UNIX commands to calculate these values. Write down the `ProductId` of both reviews. In the case of a tie, write down all `ProductId`'s to get full credit. In this case we are considering the most helpful review to be the review with the highest `HelpfulnessNumerator`.

**Hint:** *You can use multiple lines to solve this problem.*

**Relevant topics:** *[sort](https://thedatamine.github.io/the-examples-book/unix.html#sort), [head](https://thedatamine.github.io/the-examples-book/unix.html#head), [piping](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The lines of UNIX commands used to solve the problem.
- `ProductId`'s of both requested reviews.
```

##### 7. Using the `ProductId`'s from the previous question, create a new dataset called `reviews.csv` which contains the `ProductId`'s and `Score` of all reviews with the corresponding `ProductId`'s.  

**Relevant topics:** *[cut](https://thedatamine.github.io/the-examples-book/unix.html#cut), [grep](https://thedatamine.github.io/the-examples-book/unix.html#head), [redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
```

##### 8. If we didn't use `cut` prior to searching for the `ProductId`'s in (7), we would get unwanted results. Modify the solution to (7) and explore. What is happening?

**Relevant topics:** *[cat](https://thedatamine.github.io/the-examples-book/unix.html#cat), [grep](https://thedatamine.github.io/the-examples-book/unix.html#head), [redirection](https://thedatamine.github.io/the-examples-book/unix.html#piping-and-redirection)*

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands used to solve the problem.
- 1-2 sentences explaining why we need to use `cut` first.
- 1-2 sentences explaining whether or not you think people found the review helpful because the produce is overrated, underrated, or correctly reviewed, and why.
```

##### 9. Use R to load up `reviews.csv` into a new data.frame called `dat`. Create a histogram for each products' `Score`. Compare the most helpful review `Score` with the `Score`'s given in the histogram. Based on this comparison, decide (anecdotally) whether you think people found the review helpful because the product is overrated, underrated, or correctly reviewed by the masses.

**Relevant topics:** *[read.csv](file:///Users/kamstut/Dropbox/work/datamine/the-examples-book/docs/r.html#r-reading-and-writing-data), [hist](file:///Users/kamstut/Dropbox/work/datamine/the-examples-book/docs/r.html#r-plotting)*

```{block, type="bbox"}
**Item(s) to submit:**

- R code used to create the histograms.
- 3 histograms, 1 for each `ProductId`.
```

---

### Project 6 {#390-p06}

---

**Motivation:** A bash script is a powerful tool to perform repeated tasks. RCAC uses bash scripts to automate a variety of tasks. In fact, we use bash scripts on Scholar to do things like link Python kernels to your account, fix potential isues with Firefox, etc. `awk` is a programming language designed for text processing. The combination of these tools can be really powerful and useful for a variety of quick tasks.

**Context:** This is the first part in a series of projects that are designed to exercise skills around UNIX utilities, with a focus on writing bash scripts and `awk`. You will get the opportunity to manipulate data without leaving the terminal. At first it may seem overwhelming, however, with just a little practice you will be able to accomplish data wrangling tasks really efficiently. 

**Scope:** awk, UNIX utilities, bash scripts

**Learning objectives:**

```{block, type="bbox"}
- Use `awk` to process and manipulate textual data.
- Use piping and redirection within the terminal to pass around data between utilities.
- Use output created from the terminal to create a plot using R.
```

#### Dataset: 

The following questions will use the dataset found in Scholar:
`/class/datamine/data/flights/subset/YYYY.csv` 

An example of the data for the year 1987 can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/1987.csv).

#### Questions

##### 1. In previous projects we learned how to get a single column of data from a csv file. Write 1 line of UNIX commands to print the 17th column, the `Origin`, from `1987.csv`. Write another line, this time using `awk` to do the same thing. Which one do you prefer, and why?

**Relevant topics:** [cut](#cut), [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- One line of UNIX commands to solve the problem _without_ using `awk`.
- One line of UNIX commands to solve the problem using `awk`.
- 1-2 sentences describing which method you prefer and why.
```

##### 2. Write a bash script that accepts a year (1987, 1988, etc.) and a column *n* and returns the *nth* column of the associated year of data.

**Relevant topics:** [awk](#awk), [bash scripts](#writing-scripts)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
```

##### 3. How many flights came into Indianapolis (IND) in 2008? First solve this problem without using `awk`, then solve this problem using *only* `awk`.

**Relevant topics:** [cut](#cut), [grep](#grep), [wc](#wc), [awk](#awk), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- One line of UNIX commands to solve the problem *without* using `awk`.
- One line of  UNIX commands to solve the problem using `awk`. 
- The number of flights that came into Indianapolis (IND) in 2008.
```

##### 4. Do you expect the number of unique origins and destinations to be the same? Find out using any command line tool you'd like. Are they indeed the same? How many unique values do we have per category (`Origin`, `Dest`)?

**Relevant topics:** [cut](#cut), [sort](#sort), [uniq](#uniq), [wc](#wc), [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- 1-2 sentences explaining whether or not you expect the number of unique origins and destinations to be the same.
- The UNIX command(s) used to figure out if the number of unique origins and destinations are the same. 
- The number of unique values per category (`Origin`, `Dest`).
```

##### 5. In (4) we found that there are not the same number of unique `Origin`'s as `Dest`'s. Find the IATA airport code for all `Origin`'s that dont appear in a `Dest` and all `Dest`'s that don't appear in an `Origin`.

**Hint:** https://www.tutorialspoint.com/unix_commands/comm.htm

**Relevant topics:** [comm](#unix), [cut](#cut), [sort](#sort), [uniq](#uniq), [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The line(s) of UNIX command(s) used to answer the question.
- The list of `Origin`s that don't appear in `Dest`.
- The list of `Dest`s that don't appear in `Origin`.
```

##### 6. What was the average number of flights in 2008 per unique `Origin` with the `Dest` of "IND"? How does "PHX" (as a unique `Origin`) compare to the average?

**Hint:** You manually do the average calculation by dividing the result from (3) by the number of unique `Origin`'s that have a `Dest` of "IND".

**Relevant topics:** [awk](#awk), [sort](#sort), [grep](#grep), [wc](#wc)

```{block, type="bbox"}
**Item(s) to submit:**

- The average number of flights in 2008 per unique `Origin` with the `Dest` of "IND".
- 1-2 sentences explaining how "PHX" compares (as a unique `Origin`) to the average?
```

##### 7. Write a bash script that takes a year and IATA airport code and returns the year, and the total number of flights to and from the given airport. Example rows may look like:

```{txt, eval=F}
1987, 12345
1988, 44
```

Run the script with inputs: `1991` and `ORD`. Include the output in your submission.

**Relevant topics:** [bash scripts](#writing-scripts), [cut](#cut), [piping](#piping-and-redirection), [grep](#grep), [wc](#wc)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
- The output of the script given `1991` and `ORD` as inputs.
```

##### 8. Pick your favorite airport and get its IATA airport code. Write a bash script that, given the first year, last year, and airport code, runs the bash script from (7) for all years in the provided range for your given airport, or loops through all of the files for the given airport, appending all of the data to a new file called `my_airport.csv`.

**Relevant topics:** [bash scripts](#writing-scripts), [cut](#cut), [grep](#grep), [wc](#wc), for loops, echo, [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
```

##### 9. In R, load `my_airport.csv` and create a line plot showing the year-by-year change. Label your x-axis "Year", your y-axis "Num Flights", and your title the name of the IATA airport code. Write 1-2 sentences with your observations.

**Relevant topics:** [read.csv](#r-reading-and-writing-data), lines

```{block, type="bbox"}
**Item(s) to submit:**

- Line chart showing year-by-year change in flights into and out of the chosen airport.
- R code used to create the chart.
- 1-2 sentences with your observations.
```

---

### Project 7 {#390-p07}

---

**Motivation:** A bash script is a powerful tool to perform repeated tasks. RCAC uses bash scripts to automate a variety of tasks. In fact, we use bash scripts on Scholar to do things like link Python kernels to your account, fix potential isues with Firefox, etc. `awk` is a programming language designed for text processing. The combination of these tools can be really powerful and useful for a variety of quick tasks.

**Context:** This is the first part in a series of projects that are designed to exercise skills around UNIX utilities, with a focus on writing bash scripts and `awk`. You will get the opportunity to manipulate data without leaving the terminal. At first it may seem overwhelming, however, with just a little practice you will be able to accomplish data wrangling tasks really efficiently. 

**Scope:** awk, UNIX utilities, bash scripts

**Learning objectives:**

```{block, type="bbox"}
- Use `awk` to process and manipulate textual data.
- Use piping and redirection within the terminal to pass around data between utilities.
```

#### Dataset: 

The following questions will use the dataset found in Scholar:

`/class/datamine/data/flights/subset/YYYY.csv` 

An example of the data for the year 1987 can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/1987.csv).

Sometimes if you are about to dig into a dataset, it is good to quickly do some sanity checks early on to make sure the data is what you expect it to be. 

#### Questions

##### 1. Write a line of code that prints a list of the unique values in the `DayOfWeek` column. Write a line of code that prints a list of the unique values in the `DayOfMonth` column. Write a line of code that prints a list of the unique values in the `Month` column. Use the `1987.csv` dataset. Are the results what you expected?

**Relevant topics:** [cut](#cut), [sort](#sort)

```{block, type="bbox"}
**Item(s) to submit:**

- 3 lines of code used to get a list of unique values for the chosen columns.
- 1-2 sentences explaining whether or not the results are what you expected.
```

##### 2. Our files should have 29 columns. Write a line of code that prints any lines in a file that do *not* have 29 columns. Test it on `1987.csv`, were there any rows without 29 columns?

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of code used to solve the problem.
- 1-2 sentences explaining whether or not there were any rows without 29 columns.
```

##### 3. Write a bash script that, given a "begin" year and "end" year, cycles through the associated files and prints any lines that do *not* have 29 columns.

**Relevant topics:** [awk](#awk), [bash scripts](#writing-scripts)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
- The results of running your bash scripts from year 1987 to 2008.
```

##### 4. `awk` is a really good tool to quickly get some data and manipulate it a little bit. For example, let's see the number of kilometers and miles traveled in 1990. To convert from miles to kilometers, simply multiply by 1.609344. 

Example output:

```txt
Miles: 12345
Kilometers: 19867.35168
```

**Relevant topics:** [awk](#awk), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem. 
- The results of running the code.
```

##### 5. Use `awk` to calculate the number of `DepDelay` minutes by `DayOfWeek`. Use `2007.csv`.

Example output:

```txt
DayOfWeek:  0
1:  1234567
2:  1234567
3:  1234567
4:  1234567
5:  1234567
6:  1234567
7:  1234567
```

Note: 1 is Monday.

**Relevant topics:** [awk](#awk), [sort](#sort), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
```

##### 6. It wouldn't be fair to compare the total `DepDelay` minutes by `DayOfWeek` as the number of flights may vary. One way to take this into account is to instead calculate an average. Modify (5) to calculate the average number of `DepDelay` minutes by the number of flights per `DayOfWeek`. Use `2007.csv`.

Example output:

```txt
DayOfWeek:  0
1:  1.234567
2:  1.234567
3:  1.234567
4:  1.234567
5:  1.234567
6:  1.234567
7:  1.234567
```

**Relevant topics:** [awk](#awk), [sort](#sort), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
```

##### 7. As a quick follow-up, _slightly_ modify (6) to perform the same calculation for `ArrDelay`. Do the `ArrDelay`s and `DepDelay`s appear to have the highest delays on the same day? Use `2007.csv`.

Example output:

```txt
DayOfWeek:  0
1:  1.234567
2:  1.234567
3:  1.234567
4:  1.234567
5:  1.234567
6:  1.234567
7:  1.234567
```

**Relevant topics:** [awk](#awk), [sort](#sort), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
- 1-2 sentences explaining whether or not the `ArrDelay`s and `DepDelay`s appear to have the highest delays on the same day.
```

##### 8. Anyone who has flown knows how frustrating it can be waiting for takeoff, or deboarding the aircraft. These roughly translate to `TaxiOut` and `TaxiIn` respectively. If you were to fly into or out of IND what is your expected total taxi time? Use `2007.csv`.

Note: Taxi times are in minutes.

**Relevant topics:** [awk](#awk), [grep](#grep)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
```

##### 9.  What are the IATA airport codes of the 5 airports with the greatest total taxi time for 2007? Show the total taxi time for each.

Example output:

```txt
DayOfWeek:  0
IND: 1234567
IND: 1234567
IND: 1234567
IND: 1234567
IND: 1234567
```

**Relevant topics:** [awk](#awk), [head](#head), [sort](#sort)

```{block, type="bbox"}
**Item(s) to submit:**

- The code used to solve the problem.
- The output from running the code.
```

---

### Project 8 {#390-p08}

--- 

**Motivation:** A bash script is a powerful tool to perform repeated tasks. RCAC uses bash scripts to automate a variety of tasks. In fact, we use bash scripts on Scholar to do things like link Python kernels to your account, fix potential isues with Firefox, etc. `awk` is a programming language designed for text processing. The combination of these tools can be really powerful and useful for a variety of quick tasks.

**Context:** This is the first part in a series of projects that are designed to exercise skills around UNIX utilities, with a focus on writing bash scripts and `awk`. You will get the opportunity to manipulate data without leaving the terminal. At first it may seem overwhelming, however, with just a little practice you will be able to accomplish data wrangling tasks really efficiently. 

**Scope:** awk, UNIX utilities, bash scripts

**Learning objectives:**

```{block, type="bbox"}
- Use `awk` to process and manipulate textual data.
- Use piping and redirection within the terminal to pass around data between utilities.
```

#### Dataset: 

The following questions will use the dataset found in Scholar:

`/class/datamine/data/flights/subset/YYYY.csv` 

An example of the data for the year 1987 can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/flights/subset/1987.csv).

Let's say we have a theory that there are more flights on the weekend days (Friday, Saturday, Sunday) than the rest of the days, on average. We can use awk to quickly check it out and see if maybe this looks like something that is true!

##### 1. Write a line of `awk` code that, prints the number of flights on the weekend days, followed by the number of flights on the weekdays for the flights during 2008.

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of `awk` code that solves the problem.
- The result: the number of flights on the weekend days, followed by the number of flights on the weekdays for the flights during 2008.
```

##### 2. Note that in (1), we are comparing 3 days to 4! Write a line of `awk` code that, prints the average number of flights on a weekend day, followed by the average number of flights on the weekdays. Continue to use data for 2008.

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of `awk` code that solves the problem.
- The result: the average number of flights on the weekend days, followed by the average number of flights on the weekdays for the flights during 2008.
```

We want to look to see if there may be some truth to the whole "snow bird" concept where people will travel to warmer states like Florida and Arizona during the Winter. Let's use the tools we've learned to explore this a little bit. 

##### 3. Take a look at `airports.csv`. In particular run the following:

```{bash, eval=F}
head airports.csv
```

Notice how all of the non-numeric text is surrounded by quotes. The surrounding quotes would need to be escaped for any comparison within `awk`. This is messy and we would prefer to create a new file called `new_airports.csv` without any quotes. Write a line of code to do this. 

**Hint:** You could use `gsub` within `awk` to replace '"' with ''.

**Hint:** If you leave out the column number argument to `gsub` it will apply the substitution to every field in every column.

**Relevant topics:** [awk](#awk), [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of `awk` code used to create the new dataset.
```

##### 4. Write a line of commands that create a new dataset called `az_fl_airports.txt` that contains a list of airport codes for all airports from both Arizona (AZ) and Florida (FL). Use the file we created in (3),`new_airports.csv`.

**Relevant topics:** [awk](#awk)

```{block, type="bbox"}
**Item(s) to submit:**

- The line of UNIX commands to create an array called `airports`.
```

##### 5. Wow! In (4) we discovered a lot of airports! How many airports are there? Did you expect this? Use a line of bash code to answer this question.

**Relevant topics:** [echo](#echo), [wc](#wc), [piping](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of UNIX commands used to solve the problem.
- The number of airports.
- 1-2 sentences explaining whether you expected this result and why or why not.
```

##### 6. Create a new dataset that contains all of the data for flights into or out of Florida and Arizona using 2008.csv, use the newly created dataset, `az_fl_airports.txt` in (4) to do so.

**Hint:** https://unix.stackexchange.com/questions/293684/basic-grep-awk-help-extracting-all-lines-containing-a-list-of-terms-from-one-f

**Relevant topics:** [grep](#grep)

```{block, type="bbox"}
**Item(s) to submit:**

- Line of UNIX commands used to solve the problem.
```

##### 7. Now that you have code to complete (6), write a bash script that accepts the start year, end year, and filename containing airport codes (`az_fl_airports.txt`), and outputs the data for flights into or out of any of the airports listed in the provided filename containing airport codes using _all_ of the years of data in the provided range. Run the bash script to create a new file called `az_fl_flights.csv`.

**Relevant topics:** [bash scripts](#writing-scripts), [grep](#grep), for loop, [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The content of your bash script (starting with "#!/bin/bash") in a code chunk.
- The line of UNIX code you used to execute the script and create the new dataset.
```

##### 8. Use the newly created `az_fl_flights.csv` dataset to calculate the total number of flights into and out of both states by month, and by year, for a total of 3 columns (year, month, flights). Export this information to a new file called `snowbirds.csv`.

**Relevant topics:** [awk](#awk), [redirection](#piping-and-redirection)

```{block, type="bbox"}
**Item(s) to submit:**

- The line of `awk` code used to create the new dataset, `snowbirds.csv`.
```

##### 9. Load up your newly created dataset and use either R or Python (or some other tool) to create a graphic that illustrates whether or not we believe the "snowbird effect" effects flights. Include a description of your graph, as well as your (anecdotal) conclusion.

```{block, type="bbox"}
**Item(s) to submit:**

- Code used to create the visualization in a code chunk.
- The generated plot as either a png or jpg/jpeg.
- 1-2 sentences describing your plot and your conclusion.
```

---

### Project 9 {#390-p09}

### Project 10 {#390-p10}

### Project 11 {#390-p11}

--- 

**Motivation:** Being able to use results of queries as tables in new queries (also known as writing sub-queries), and calculating values like MIN, MAX, and AVG in aggregate are key skills to have in order to write more complex queries. In this project we will learn about aliasing, writing sub-queries, and calculating aggregate values.

**Context:** We are in the middle of a series of projects focused on working with databases and SQL. In this project we introduce aliasing, sub-queries, and calculating aggregate values using a much larger dataset!

**Scope:** sql, sql in R

**Learning objectives:**

```{block, type="bbox"}
- Demonstrate the ability to interact with popular database management systems within R.
- Solve data-driven problems using a combination of SQL and R.
- Basic clauses: select, order by, limit, desc, asc, count, where, from, etc.
- Showcase the ability to filter, alias, and write subqueries.
- Perform grouping and aggregate data using group by and the following functions: count, max, sum, avg, like, having. Explain when to use having, and when to use where.
```

#### Dataset

`elections` database & `/class/datamine/data/election/itcontYYYY.txt` (for example, data for year 1980 would be `/class/datamine/data/electionitcont1980.txt`)

A public sample of the data can be found here:

https://www.datadepot.rcac.purdue.edu/datamine/data/election/itcontYYYY.txt (for example, data for year 1980 would be https://www.datadepot.rcac.purdue.edu/datamine/data/election/itcont1980.txt)

Up until now, you've been working with a neatly organized database containing baseball data. As fantastic as this database is, it would be trivial to load up the entire database in R or Python and do your analysis using `merge`-like functions. Now, we are going to deal with a much larger set of data.

##### 1. Approximately how large was the lahman database (use the sqlite database in Scholar: `/class/datamine/data/lahman/lahman.db`)? Use UNIX utilities you've learned about this semester to write a line of code to return the amount of data (in MB) in the elections folder `/class/datamine/data/election/`. How much data (in MB) is there? 

The data in that folder has been added to the `elections` database in the `elections` table. Write a SQL query that returns how many rows of data are in the database. How many rows of data are in the database?

**Hint:** This will take some time! Be patient.

**Relevant topics:** [sql](#sql), [sql in R](#r-sql), [awk](#awk), [ls](#ls)

```{block, type="bbox"}
**Item(s) to submit:**

- Approximate size of the lahman database in mb.
- Line of code (bash/awk) to calculate the size (in mb) of the entire elections dataset in `/class/datamine/data/election`.
- The size of the elections data in mb.
- SQL query used to find the number of rows of data in the `elections` table in the `elections` database.
- The number of rows in the `elections` table in the `elections` database.
```

##### 2. Write a SQL query using the `LIKE` command to find a unique list of `zip_code`s that start with "479". How many unique `zip_code`s are there that begin with "479"?

**Hint:** Make sure you only select `zip_code`s.

**Relevant topics:** [sql](#sql), like

```{block, type="bbox"}
**Item(s) to submit:**

- SQL queries used to answer the question.
- The first 5 results from running the query.
```

##### 3. Write a SQL query that counts the number of donations (rows) that are from Indiana. How many donations are from Indiana? Rewrite the query and create an _alias_ for our field so it doesn't read `COUNT(*)` but rather `Indiana Donations`. 

**Relevant topics:** [sql](#sql), [where](#sql-examples), [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to answer the question.
- The result of the SQL query.
```

##### 4. Rewrite the query in (3) so the result is displayed like the following:

```{txt}
+-------------+
| Donations   |
+-------------+
| IN: 1111778 |
+-------------+
```

**Hint:** Use CONCAT and aliasing to accomplish this.

**Relevant topics:** [sql](#sql), [aliasing](#sql-examples), concat

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to answer the question.
```

##### 5. In (2) we wrote a query that returns a unique list of `zip_code`s that start with "479". In (3) we wrote a query that counts the number of donations that are from Indiana. Use our query from (2) as a sub-query to find how many donations come from areas with `zip_code`s starting with "479". What percent of donations in Indiana come from said `zip_code`s?

**Relevant topics:** [sql](#sql), [aliasing](#sql-examples), subqueries

```{block, type="bbox"}
**Item(s) to submit:**

- SQL queries used to answer the question.
- The percentage of donations from Indiana from `zip_code`s starting with "479".
```

##### 6. In (3) we wrote a query that counts the number of donations that are from Indiana. When running queries like this, a natural "next question" is to ask the same question about another state. SQL gives us the ability to calculate functions in aggregate when grouping by a certain column. Write a SQL query that returns the state, number of donations from each state, the sum of the donations (`transaction_amt`). Which 5 states gave the most donations (highest count)? Order you result from most to least.

**Hint:** You may want to create an alias in order to sort.

**Relevant topics:** [sql](#sql), group by

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to answer the question. 
- Which 5 states gave the most donations?
```

##### 7. Write a query that gets the number of donations, and sum of donations, by year, for Indiana. Create one or more graphics that highlights the year-by-year changes. Write a short 1-2 sentences explaining your graphic(s).

**Relevant topics:** [sql in R](#sql-in-R), group by

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to answer the question.
- R code used to create your graphic(s).
- 1 or more graphics in png/jpeg format.
- 1-2 sentences summarizing your graphic(s).
```

---

### Project 12 {#390-p12}

### Project 13

---

**Motivation:** Databases you will work with won't necessarily come organized in the way that you like. Getting really comfortable writing longer queries where you have to perform many joins, alias fields and tables, and aggregate results, is important. In addition, gaining some familiarity with terms like _primary key_, and _foreign key_ will prove useful when you need to search for help online. In this project we will write some more complicated queries with a fun database. Proper preparation prevents poor performance, and that means practice!

**Context:** We are towards the end of a series of projects that give you an opportunity to practice using SQL. In this project, we will reinforce topics you've already learned, with a focus on subqueries and joins. 

**Scope:** SQL, sqlite

**Learning objectives:**

```{block, type="bbox"}
- Write and run SQL queries in `sqlite` on real-world data.
- Identify primary and foreign keys in a SQL table.
```

#### Dataset

`/class/datamine/data/movies_and_tv/imdb.db`

A public sample of the data can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/data/movies-and-tv/imdb.db).

#### Questions

##### 1. A primary key is a field in a table which uniquely identifies a row in the table. Primary keys _must_ be unique values, and this is enforced at the database level. A foreign key is a field whose value matches a primary key in a different table. A table can have 0-1 primary key, but it can have 0+ foreign keys. Examine the `titles` table. Do you think there are any primary keys? How about foreign keys? 

**Relevant topics:** primary key, foreign key

```{block, type="bbox"}
**Item(s) to submit:**

- List any primary or foreign keys in the `episodes` table.
```

##### 2. Examine the `episodes` table. Based on observation and the column names, do you think there are any primary keys? How about foreign keys?

**Relevant topics:** primary key, foreign key

```{block, type="bbox"}
**Item(s) to submit:**

- List any primary or foreign keys in the `episodes` table.
```

If you paste a `title_id` to the end of the following url, it will pull up the page for the title. For example, https://www.imdb.com/title/tt0413573 leads to the page for the TV series Grey's Anatomy. 

##### 3. Write a query to confirm that the `title_id` tt0413573 does indeed belong to Grey's Anatomy.

**Relevant topics:** [select](#sql-examples), where

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve the problem in a code chunk.
- Output of the query.
```

##### 4. The `episode_title_id` column in the `episodes` table references titles of individual episodes of a tv series. The `show_title_id` references the titles of the show itself. With that in mind, write a query that gets a list of all of the episodes and titles of Grey's Anatomy.

**Relevant topics:** [inner join](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve the problem in a code chunk.
```

##### 5. Like we explained in (3), you can find the `title_id` of a tv show, a tv show episodes, or a movie by browsing imdb.com and getting the `title_id` directly from the url. Browse imdb.com and find your favorite tv show. Get the `title_id` from the url and run the following query to confirm that the tv show is in our database:

```{sql, eval=F}
SELECT * FROM titles WHERE title_id='<title id here>';
```

Make sure to replace '<title id here>' with the `title_id` of your favorite show. If your show does not appear, or has only a single season, pick another show until you find one we have in our database with multiple seasons.

```{block, type="bbox"}
**Item(s) to submit:**

- The `title_id` of your favorite tv show.
- The output from running the provided (modified) query.
```

##### 6. We want to write a query that returns the title and rating of the highest rated episode of the tv show you chose in (5). In order to do so, first write a query that returns a list of `episode_title_id`s (found in the `episodes` table), with the `primary_title` (found in the `titles` table) of the episode. 

**Relevant topics:** [inner join](#sql-examples), [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve the problem in a code chunk.
- The first 5 results from your query.
```

##### 7. Write a query that adds the rating to the end of each episode. To do so, use the query you wrote in (6) as a subquery. Was this also your favorite episode?

**Relevant topics:** [inner join](#sql-examples), [aliasing](#sql-examples), subqueries, [desc](#sql-examples), [limit](#sql-examples), [order by](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to solve the problem in a code chunk.
- The `episode_title_id`, `primary_title`, and `rating` of the top rated episode from the tv series from (5).
- A statement saying whether it is also your favorite episode.
```

##### 8. Write a query that returns the `season_number` (from the `episodes` table), and average `rating` (from the `ratings` table) for each season. Write another query that only returns the season number and `rating` for the highest rated season. Consider the highest rated season the season with the highest average.

**Relevant topics:** [inner join](#sql-examples), [aliasing](#sql-examples), [group by](#sql-examples), having, avg

```{block, type="bbox"}
**Item(s) to submit:**

- The 2 SQL queries used to solve the problems in a code chunk.
```

##### 9. Write a query that returns the `primary_title`, and `rating` of the highest rated episode per season for your tv show from (5). 

**Relevant topics:** max, subqueries, [group by](#sql-examples), having, [inner join](#sql-examples), [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- The SQL query used to solve the problem.
- The output from your query.
- 1-2 sentences explaining whether or not you agree.
```

---
 
### Project 14 {#390-p14}

### Project 15 {#390-p15}

---

**Motivation:** We've done a lot of work with SQL this semester. Let's review concepts in this project and mix and match R, Python, and SQL to solve data-driven problems.

**Context:** In this project, we will reinforce topics you've already learned, with a focus on SQL.

**Scope:** SQL, sqlite, R, Python

**Learning objectives:**

```{block, type="bbox"}
- Write and run SQL queries in `sqlite` on real-world data.
- Use SQL from within R and Python.
```

#### Dataset

`/class/datamine/data/movies_and_tv/imdb.db`

A public sample of the data can be found [here](https://www.datadepot.rcac.purdue.edu/datamine/data/movies-and-tv/imdb.db).

In this project we want to offer the flexibility of using your choice of R and/or Python. To keep things as consistent as possible, please use Rmarkdown on https://rstudio.scholar.rcac.purdue.edu/. See [here](#python-on-scholar) to learn how to run Python in this environment.

F.R.I.E.N.D.S is a popular tv show. They have an interesting naming convention for the names of their episodes. They all begin with the text "The One ...". There are 6 primary characters in the show: Chandler, Joey, Monica, Phoebe, Rachel, and Ross. Let's use SQL and R to take a look at how many times each characters' names appear in the title of the episodes.

#### Questions 

##### 1. Write a query that gets the `episode_title_id`, `primary_title`, `rating`, and `votes`, of all of the episodes of Friends (`title_id` is tt0108778). 

**Hint:** You can slightly modify the solution to question (7) in project 13.

**Relevant topics:** [inner join](#sql-examples), subqueries, [aliasing](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- SQL query used to answer the question.
- First 5 results of the query.
```

The next couple of questions should be complete in the same language. You can use either R or Python, but you must use the same for both questions.

##### 2. Now that you have a working query, connect to the database and run the query to get the data into an R or `pandas` data frame. In previous projects, we learned how to used regular expressions to search for text. For each character, how many episodes `primary_title`s contained their name? 

**Relevant topics:** [SQL in R](#sql-in-r), [SQL in Python](#sql-in-python), [grep](#r-grep)

```{block, type="bbox"}
**Item(s) to submit:**

- R or Python code in a code chunk that was used to find the solution.
- The solution pasted below the code chunk.
```

##### 3. Create a graphic showing our results in (2) using your favorite package. Make sure the plot has a good title, x-label, y-label, and try to incorporate some of the following colors: #273c8b, #bd253a, #016f7c, #f56934, #016c5a, #9055b1, #eaab37.

**Relevant topics:** [plotting](#r-plotting)

```{block, type="bbox"}
**Item(s) to submit:**

- The R or Python code used to generate the graphic.
- The graphic in a png or jpg/jpeg format.
```

##### 4. Use any combination of SQL, R, and Python you'd like in order to find which of the following 3 genres has the highest average rating for movies (see `type` column from `titles` table): Romance, Comedy, Animation. In the `titles` table, you can find the genres in the `genres` column. There may be some overlap (i.e. a movie may have more than one genre), this is ok. 

To query rows which have the genre Action as one of its genres:

```{sql, eval=F}
SELECT * FROM titles WHERE genres LIKE '%action%';
```

**Relevant topics:** [like](#sql-examples), [inner join](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- Any code you used to solve the problem in a code chunk.
- The average rating of each of the genres listed for movies.
```

##### 5. Write a function called `top_episode` in R or in Python which accepts the path to the `imdb.db` database, as well as the `title_id` of a tv series (for example, "tt0108778" or "tt1266020"), and returns the `season_number`, `episode_number`, `primary_title`, and `rating` of the highest rated episode in the series. Test it out on some of your favorite series, and share the results.

**Relevant topics:** [functions](#writing-functions), [inner join](#sql-examples), [order by](#sql-examples)

```{block, type="bbox"}
**Item(s) to submit:**

- Any code you used to solve the problem in a code chunk.
- The results for at least 3 of your favorite tv series.
```
