---
title: "Key learnings from DataMine 19000"
author: "Your_Name_Here"
date: ''
output:
  beamer_presentation: default
  ioslides_presentation: default
  slidy_presentation: default
---

## Your_slide_title

Text here.

- Key learning #1
- Key learning #2
- Key learning #3
- Key learning #4

## Your_slide_title

Key learning #1: 1-2 sentences properly explaning the key learning.

## Your_slide_title

Key learning #1: 2 showcases of the key learning

```{r eval=F}
# R CODE HERE IF R IS INVOLVED
```

```{sql eval=F}
# SQL CODE HERE IF SQL IS INVOLVED
```

## Your_slide_title

Key learning #1: 1-2 sentences explaining why do you think this key learning is revelant to you.

## Your_slide_title

Key learning #2: 1-2 sentences properly explaning the key learning.

## Your_slide_title

Key learning #2: 2 showcases of the key learning

```{r eval=F}
# R CODE HERE IF R IS INVOLVED
```

```{sql eval=F}
# SQL CODE HERE IF SQL IS INVOLVED
```

## Your_slide_title

Key learning #2: 1-2 sentences explaining why do you think this key learning is revelant to you.

## Your_slide_title

Key learning #3: 1-2 sentences properly explaning the key learning.

## Your_slide_title

Key learning #3: 2 showcases of the key learning

```{r eval=F}
# R CODE HERE IF R IS INVOLVED
```

```{sql eval=F}
# SQL CODE HERE IF SQL IS INVOLVED
```

## Your_slide_title

Key learning #3: 1-2 sentences explaining why do you think this key learning is revelant to you.

## Your_slide_title

Key learning #4: 1-2 sentences properly explaning the key learning.

## Your_slide_title

Key learning #4: 2 showcases of the key learning

```{r eval=F}
# R CODE HERE IF R IS INVOLVED
```

```{sql eval=F}
# SQL CODE HERE IF SQL IS INVOLVED
```

## Your_slide_title

Key learning #4: 1-2 sentences explaining why do you think this key learning is revelant to you.

## Your_slide_title

- Recommendation/suggestion # 1
- Recommendation/suggestion # 2
- Recommendation/suggestion # 3
