---
title: "STAT 19000 - Project 8 Question 03"
author: "Your_Name_Here"
date: "3/23/2020"
output: 
    pdf_document: default
---


Read the hotel booking demand data into a dataframe.
```{r reading-the-data}
hotelDF <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv', stringsAsFactors = FALSE)
head(hotelDF)
```

There are `r ncol(hotelDF)` variables on `r nrow(hotelDF)` hotel bookings. Here we have two types of hotels: `r unique(hotelDF$hotel)`.

Let's compare the `reserved_room_type` and the `assigned_room_type` for the two hotel types. First we need to create a variable that compares `reserved_room_type` and `assigned_room_type`. Here we will just compare whether or not they got the exact room type they reserved. 

```{r}
# Getting variable that compares reserved_room_type and assigned_room_type
got_reserved_room_type <- hotelDF$reserved_room_type == hotelDF$assigned_room_type
# Double checking
str(got_reserved_room_type)
```

Now, let's see the frequency of people that got their reserved room type. We can see the frequency using either a table, or a barplot.

```{r}
# Frequency table
table(got_reserved_room_type)

# Barplot
barplot(table(got_reserved_room_type))
```

As expected, people tend to get the room types they reserve. Now it's your turn. Find an interesting factoid in this dataset and be sure to document it here, making it reproducible!
