---
title: "STAT 19000 - Project 8 Question 02"
author: "Your_Name_Here"
date: "3/23/2020"
output: 
    pdf_document: default
---

(Header 1) Worming through book data

(Header 2) Reading the data

We will begin by reading our dataset into a data frame. 

```{r getting-book-dataset}
myBookDF <- read.csv('https://raw.githubusercontent.com/zygmuntz/goodbooks-10k/master/books.csv',
                      stringsAsFactors = FALSE)
```

Our data has information on `r nrow(myBookDF)` books for `r ncol(myBookDF)` variables. In this analysis, we will focus only on the following variables

"book_id"
"goodreads_book_id"
"original_publication_year"
"original_title"
"title"
"language_code"
"average_rating"
"ratings_count"

Let's begin by creating a subset of our data containing the variables listed above
```{r subsetting-book-dataset}
cols_to_keep <- c("book_id", "goodreads_book_id", "original_publication_year", "original_title", "title", "language_code", "average_rating", "ratings_count")
subBookDF <- subset(myBookDF, select=cols_to_keep)
dim(subBookDF)
```

(Header 2) Analyzing the data

(Header 3) Hypothesis #1

The title's length has no association with the rating of a book.

(Header 4) Exploratory analysis

To evaluate our hypothesis, we will compare how many words we have in a title with the average rating, all in a scatterplot.

```{r testing-hypothesis}
count <- function(x) {length(strsplit(x, " ")[[1]])}
title_length <- sapply(subBookDF$original_title, count)
length(title_length)
plot(title_length, subBookDF$average_rating, xlab = "Title's length", ylab = "Average rating")
```

(Header 3) Hypothesis #2

There is a difference in rating between english and non-english books.

(Header 4) Exploratory analysis

In order to evaluate our second hypothesis, we will first remove the books that are missing the language information. For the remaining books, we will categorize based on the `language_code` variable, and compare their mean values and distributions using histograms.

