---
title: "STAT19000 - Project 8 - Example"
author: "Author"
date: "3/23/2020"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This is my title

## Section 00 - Quick comments on editing files using Markdown

Here are a few commands:

**Important: this sentence is bolded**. We could have obtained the same results by doing __Important: this sentence is bolded__. 

*Note that this sentence is italicized*. We could have achieved the same results by doing _Note that this sentence is italicized_.

~~Ignore this~~. 

## Section 01 - Getting the data

This is the first example of an RMarkdown document. RMarkdown files allows us to make a reproducible document for our data analysis. Let's redo some analysis we did in the project 2 examples.

We will use the dataset [wine_data](https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-28/winemag-data-130k-v2.csv). Note that we just made a link. To link webpages, we can use: [name we want](https://example.com).

The chunk below is the code to read the data.
```{r reading-data}
wine_data <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-28/winemag-data-130k-v2.csv', stringsAsFactors = F)
dim(wine_data)
```

`reading data` is the name of our chunk of code. You can have chunk of codes with no name. However, the names should be **unique**. That is, if you have more than one chunk of code with the same name, your code will produce an error. 

The code chunks have some options. Some of the most important are:

* **eval:** when set to `FALSE`, this option will prevent code from being evaluated (run).
* **include:** when set to `FALSE`, this option will run the code, but will not show the code nor it's results in the document.
* **echo:** when set to `FALSE`, this option will not show the code but will show the result.
* **message and/or warning:** when set to `FALSE`, this option will not show messages and/or warnings.
* **results:** when set to 'hide', this option will hide the output.
* **fig.show:** when set to 'hide', this option will will hide plots.
* **error:** when set to `TRUE`, this option will continue to render even if your code returns an error. This is very useful if you want to share a particular error with someone to get their help with it.
* **cache:** when set to `TRUE`, this option will save the output of the code chunk to reuse in future knits/renderings. Useful when you have some computations that take a long time. Be careful when using it.

If we wanted to hide our code but show the results we would do:
```{r reading-data-hidden, echo=FALSE}
wine_data <- read.csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-28/winemag-data-130k-v2.csv', stringsAsFactors = FALSE)
dim(wine_data)
```

You may have noticed that we used a list to show some options in code chunks. There are two types of lists we can do: *unordered* or *ordered*. For the unordered we use:

* My item 1
* My item 2
    + My item 2.a
    + My item 2.b
    
For the ordered we use:

1. My ordered item 1
2. My ordered item 2
    + My ordered item 2.a
    + My ordered item 2.b
    
## Section 02 - Exploratory data analysis

Our dataset has `r nrow(wine_data)` rows and `r ncol(wine_data)` columns. Let's look at the distribution of `points`.

```{r average-rating-distribution, echo=FALSE}
hist(wine_data$points)
```

Looking at the country distribution may be interesting.
```{r language}
table(wine_data$country)
```

We could do a similar table by hand, we could do:

US            | Spain         | Portugal      | France        | Argentina     | Other         |
------------- | ------------- | ------------- | ------------- | ------------- | ------------- |
54504         | 6645          | 5691          | 22093          | 3800         | 37238         |
